---
title: "Ex-2-1: Quarto Basics"
abstract: "Quarto"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

Quarto runs on your machine, not in your browser, so every exercise here is
something you do in RStudio.

Several of these ask you to **break something on purpose**. That is deliberate.
Almost everything that goes wrong with Quarto announces itself through an error
message, and the fastest way to learn those messages is to cause them
deliberately while you are calm, rather than meet them for the first time at
eleven o'clock the night before an assignment is due.

# Setup {#sec-setup}

1. You cloned <https://github.com/tmieno2/quarto-examples> in lecture 02-0.
   Find that folder now and confirm these exist:
   - `templates/sample_qmd.qmd`
   - `templates/custom.scss`
   - `templates/nebraska-n.jpg`
2. Make a new folder `quarto-practice`, and in RStudio do
   `File -> New Project -> Existing Directory` on it.
3. Inside it, make a `data` folder and put a copy of `corn_yields.csv` in it.
   That file is in the datasets repository from
   [Ex-3-1](3-1-input-output.qmd).

::: {.callout-important title="It worked if"}
`here::here()` in the console prints the path to `quarto-practice`.
:::

::: {.callout-tip collapse="true" title="If you cannot find your clone"}
Open Sublime Merge, open the repository, then `File -> Open Containing Folder`.
If you deleted it, clone it again; it takes twenty seconds.
:::

# Your first render

**Task.**

1. `File -> New File -> Quarto Document`. Save it as `first.qmd` in
   `quarto-practice`.
2. Delete everything below the YAML header.
3. Add a `##` heading, a sentence of ordinary text, and one R chunk that draws a
   plot from a built-in dataset.
4. Click **Render**.

```{r}
#| eval: false
plot(mtcars$wt, mtcars$mpg)
```

::: {.callout-important title="It worked if"}
A file called `first.html` appears **next to** `first.qmd` in the Files pane,
and your figure is inside it.
:::

::: {.callout-note collapse="true" title="What just happened"}
Render did three things, in this order:

1. started a brand new R session and ran every chunk in order, top to bottom
2. pasted each chunk's results underneath its code
3. converted the whole thing, text and code and results, into html

The `.qmd` is your source and the `.html` is the output. You edit the first and
you hand in the second. They are two separate files and they always will be.

This is also the answer to "why not just use Word?". In Word the numbers in your
document are things you typed. Here they are things your document computed, and
they cannot disagree with your data.
:::

# Make the output show what you want

The default shows your code, your results, and every message the packages
decide to emit. That is almost never what a reader wants.

**Task.** Write a chunk that loads `tidyverse` and draws a plot, then produce
each of these four outputs. Each needs a different chunk option, and you should
find them before opening the answer.

a. The figure appears; the code is hidden; no package startup noise.
b. The code appears, nicely formatted, but does not run.
c. The chunk runs, but nothing whatever appears in the html.
d. Everything from (a), applied to every chunk in the document at once, without
   editing them one by one.

::: {.callout-important title="It worked if"}
For (c) you can *prove* the chunk ran even though nothing appeared. Define an
object inside it and use that object in a later, visible chunk.
:::

::: {.callout-note collapse="true" title="Answer"}
````
a.  #| echo: false
    #| message: false
    #| warning: false

b.  #| eval: false

c.  #| include: false

d.  in the YAML header:

    execute:
      echo: false
      message: false
      warning: false
````

The distinction to hold on to:

- `eval` controls whether the code **runs**
- `echo` controls whether the code is **shown**
- `include: false` hides everything, code and output both, while still running

`include: false` is exactly what you want for the `library()` chunk at the top of
a report. It has to run, and nobody wants to read it.

For (d), note that a chunk-level option always beats the global one, so you can
set a document-wide default and then override it for the two chunks where you
actually do want to show the code.
:::

# Label your chunks

**Task.**

1. Give each chunk in `first.qmd` a label describing what it does.
2. Label the plotting chunk `fig-mpg`, give it a `fig-cap`, and refer to it from
   your text with `@fig-mpg`.
3. Now break it: copy a labelled chunk and paste it lower down without changing
   the label. Render, and read the error.
4. Break it again: rename `fig-mpg` to `fig_mpg`, with an underscore. Render and
   look carefully at the cross-reference in the output.

::: {.callout-important title="It worked if"}
Step 3 stops the render with a clear complaint. Step 4 does **not** stop the
render, and instead prints `?fig_mpg` into your text.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
```{r}
#| label: fig-mpg
#| fig-cap: "Fuel economy falls with weight."
#| echo: false
plot(mtcars$wt, mtcars$mpg)
```

Heavier cars are less efficient (@fig-mpg).
````

Three things labels buy you: the error message names the chunk that failed
rather than "chunk 7"; the cached results have somewhere to live; and figures
and tables become cross-referenceable.

Two rules:

- **Labels must be unique.** A duplicate gives
  `Duplicate chunk label 'fig-mpg'`, and the chunk it blames is the second one
  even though the real cause is usually that you copy-pasted the first.
- **Hyphens, never underscores.** `fig_mpg` is a perfectly legal label and the
  chunk runs fine, but the cross-reference silently fails and prints
  `?fig_mpg`. Nothing warns you. This is the single most common way a
  cross-reference breaks.
:::

# The render does not know what your console knows

**Task.** Break this one on purpose.

1. In the **console**, and only the console, type `secret <- 42`.
2. In `first.qmd`, add a chunk containing just `secret * 2`. Run it with the
   green arrow. It works.
3. Now click Render.

::: {.callout-important title="It worked if"}
The render fails with `object 'secret' not found`, even though the identical
line worked when you clicked the green arrow thirty seconds earlier.
:::

::: {.callout-note collapse="true" title="Why, and what to do about it"}
Rendering starts a **completely new R session**. Nothing in your Environment
pane is available to it, not your data, not your packages, not the object you
made an hour ago and forgot about.

This is a feature rather than an annoyance. It is the guarantee that your
document contains everything needed to reproduce itself, which is what makes it
work on my computer as well as yours.

The habit that follows: your qmd must create everything it uses. Load the
packages, read the data, build the objects, all inside chunks in the document.

The check, which takes ten seconds and should become automatic before you submit
anything:

```
Session -> Restart R
Run -> Run All
```

If that fails, your document depends on something invisible, and it will fail
for me too.
:::

# Diagnose three failed renders

Cause each of these, **read the message before fixing it**, and write down which
chunk Quarto blamed.

a. Misspell a function name, for instance `libary(tidyverse)`.
b. Call a function from a package you have not installed, for instance
   `gganimate::animate()`.
c. Ask for a column that does not exist, for instance `corn$Yeild`.

Then: make (a) render anyway, printing the error into the document instead of
stopping.

::: {.callout-note collapse="true" title="What each looks like, and the fix"}
**a.** `could not find function "libary"`. Quarto names the chunk and the line.
Fix the spelling. Remember that this exact message also appears when a package
is not loaded, so "could not find function" always means one of three things:
typo, not loaded, or not installed.

**b.** `there is no package called 'gganimate'`. Install it. Note this is a
*different* message from (a), and the difference is informative: it tells you
the package is missing rather than merely unloaded.

**c.** `object 'Yeild' not found`, wrapped in dplyr's `In argument:` framing if
you were inside a verb. First check `names()` against what you typed, and watch
capitalisation: `Yield` and `yield` are different columns.

**Rendering anyway:**

```
#| error: true
```

The error is printed into the document and the render continues. These lecture
notes use it deliberately, to show you what a failure looks like. Do not leave
it switched on in an assignment: it lets broken code through silently, which is
worse than failing.
:::

::: {.callout-tip collapse="true" title="Reading a Quarto error"}
Work from the bottom up. The last few lines are usually R's actual complaint;
everything above is Quarto and knitr explaining how they got there. The line
beginning `Quitting from` names the chunk and line number, and is the fastest
thing to look for.
:::

# Caching, and the trap that comes with it

Once a chunk takes real time, you do not want to run it on every render.

**Task.**

1. Add a chunk that deliberately takes a few seconds:

```{r}
#| eval: false
slow_result <- {Sys.sleep(5); mean(mtcars$mpg)}
```

2. Render, and notice how long it takes. Render again, and notice it is no
   faster.
3. Turn on caching for that chunk. Render twice more and compare.
4. Now spring the trap. Put the data in an *earlier* chunk, have the cached
   chunk use it, then change the earlier chunk. Render.
5. Fix it two ways.

::: {.callout-important title="It worked if"}
At step 4 your document reports a number computed from data that no longer
exists anywhere in the file.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
```{r}
#| label: prep
d <- mtcars[mtcars$cyl == 4, ]     # change 4 to 6 at step 4
```

```{r}
#| label: slow-bit
#| cache: true
slow_result <- {Sys.sleep(5); mean(d$mpg)}
```
````

The trap: `cache: true` reruns a chunk when **that chunk's own code** changes.
Editing `prep` does not change `slow-bit`, so the cache is kept and your
document silently reports last week's answer. Nothing warns you, and the number
looks perfectly reasonable.

Two fixes:

```
#| dependson: prep      # rerun me when prep changes
```

or delete the `_cache` folder next to your qmd, which throws everything away and
starts clean. That folder is also the first thing to try when caching behaves
inexplicably.

The rule of thumb: cache chunks that are slow *and* self-contained. Do not
cache a chunk that reads data, because the whole point of caching is to not
notice the data changed.
:::

::: {.callout-tip collapse="true" title="freeze, for whole documents"}
On a website, `freeze: auto` in `_quarto.yml` stops Quarto re-running a document
at all unless its source has changed. That is the right tool for a site with
thirty pages where only one has been edited. `cache` is per chunk; `freeze` is
per document.
:::

# Paths, from inside a render

You met this in [Ex-3-1](3-1-input-output.qmd). It is here again because it
breaks Quarto documents specifically, and because seeing it twice is how it
sticks.

**Task.**

1. Make a subfolder `analysis` and create `analysis/report.qmd` inside it.
2. Add a chunk reading `read_csv("data/corn_yields.csv")`.
3. Render. It fails.
4. Read the error carefully. It tells you which folder R looked in.
5. Fix it so the same line works whether you render from `analysis/` or paste it
   into the console at the project root.

::: {.callout-important title="It worked if"}
`report.qmd` renders, and the exact same line also works pasted into the
console, unchanged.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
read_csv(here::here("data", "corn_yields.csv"))
```

When Quarto renders `analysis/report.qmd`, the working directory becomes
`analysis/`, so a relative path looks for `analysis/data/corn_yields.csv`.
`here()` always builds from the project root, so it does not care where the
rendering document lives.
:::

# Make an html you can actually submit

This is the one that decides whether your assignment arrives readable.

**Task.**

1. Render a document containing a figure. Notice that a folder named
   `something_files` appeared next to your html.
2. Copy **only** the `.html` file to your Desktop. Not the folder. Open it.
3. Observe what happened to your figure.
4. Add `embed-resources: true` under `html:` in the YAML, render again, and
   repeat step 2.

::: {.callout-important title="It worked if"}
The second copy shows the figure on the Desktop, with no folder next to it.
:::

::: {.callout-note collapse="true" title="Answer, and why this matters"}
```yaml
---
title: "My report"
format:
  html:
    embed-resources: true
---
```

By default Quarto writes the html plus a `_files` folder holding every figure,
and the html points at that folder. Separate the two and the figures vanish.

When you email the html on its own, or upload just that file, you have separated
them. Your document arrives with broken images and you have no way of knowing,
because on your machine the folder is still sitting right there and everything
looks perfect.

`embed-resources: true` packs the figures, the styling, and everything else into
the single html file. It gets bigger. Submit it anyway.
:::

::: {.callout-tip collapse="true" title="Check it properly"}
Zip the single html file on its own and mail it to yourself, then open it from
the download. That is the closest thing to what your grader will actually do,
and it catches problems that dragging to the Desktop does not.
:::

# Numbers in your prose that update themselves

**Task.**

1. Give `first.qmd` a proper header: title, your name, today's date, a table of
   contents, and section numbering.
2. Read `corn_yields.csv` in a hidden chunk.
3. Write a sentence in ordinary prose reporting the number of observations and
   the mean yield, using inline R rather than typing the numbers.
4. Now filter the data to irrigated observations only, re-render, and watch the
   sentence change.

::: {.callout-important title="It worked if"}
You did not retype a single number in step 4.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
---
title: "Corn yields in Nebraska"
author: "Your Name"
date: today
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
---

The dataset has `r nrow(corn)` observations, with a mean yield of
`r round(mean(corn$Yield), 1)` bushels per acre.
````

Inline code is a single backtick, the letter `r`, a space, your expression, and
a closing backtick. It runs in the same session as your chunks, so it can use
any object they created.

This one exercise is the whole argument for Quarto over Word. In Word that
sentence is a number you typed once, and the moment the data changes it quietly
becomes false. Here it cannot be wrong, because nobody typed it.

Keep the expressions short. If a sentence needs three lines of computation,
compute it in a hidden chunk and put the resulting object inline.
:::

# Put it together

**Task.** Produce a short report on the corn yield data, from scratch, that:

- has a title, your name, a date, a table of contents, and numbered sections
- reads its data with `here()`
- hides every `library()` call and all package messages
- contains at least one labelled, captioned figure, cross-referenced from the
  text
- reports at least two numbers inline rather than typed
- renders to a single self-contained html

Then run the real test: `Session -> Restart R`, then Render, then copy only the
html somewhere else and open it.

::: {.callout-important title="It worked if"}
It renders from a clean session, and the copy opens correctly somewhere else on
your disk. That is the standard every assignment in this course has to meet.
:::
