---
title: "Ex-2-2: Revealjs Slides and Websites"
abstract: "Quarto"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

All of these are done in RStudio. Work in the `quarto-practice` project from
[Ex-2-1](2-1-quarto-basics.qmd), and keep your `quarto-examples` clone open in
a Finder or Explorer window, because two exercises copy files out of it.

# Turn a report into a deck

A revealjs deck is not a different kind of document. It is the same document
with a different output format, and this exercise exists to make that concrete.

**Task.**

1. Copy `first.qmd` from Ex-3-1 to `slides.qmd`.
2. Change `format: html` to `format: revealjs`, and render.
3. Look at what happened to your headings.
4. Take your longest slide and split it in two with a second `##`.
5. Add a `#` heading somewhere, and note how it differs from `##`.

::: {.callout-important title="It worked if"}
The deck opens in a browser, the arrow keys move between slides, and pressing
`o` shows every slide at once.
:::

::: {.callout-note collapse="true" title="Answer"}
```yaml
---
title: "Corn yields in Nebraska"
format: revealjs
---
```

Every `##` became the title of a new slide, and everything beneath it became
that slide's content. A single `#` produces a section-divider slide instead.

That is the entire slide-breaking rule, and it has a useful consequence:
restructuring a deck is just moving headings around. There is no slide object to
drag, no master to fight with.
:::

::: {.callout-warning collapse="true" title="If you got a blank first slide"}
Anything between the YAML header and the first `##` becomes its own untitled
slide. A stray HTML comment, a `<style>` block, a leftover sentence. Move it
below a heading or delete it.

While you are here, press `o` and check the whole deck for blank slides. They
are easy to miss when you are paging through one at a time.
:::

# Fit two things side by side

**Task.** Build one slide with a figure on the left and the bullets explaining
it on the right, roughly two thirds / one third. Then make the bullets appear
one at a time as you press the arrow key.

::: {.callout-important title="It worked if"}
Nothing overflows the bottom of the slide, and the first arrow press reveals a
bullet rather than jumping to the next slide.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
## Yields by irrigation status

:::: {.columns}

::: {.column width="65%"}
```{r}
#| echo: false
ggplot(corn) + geom_boxplot(aes(x = factor(Irrigated), y = Yield))
```
:::

::: {.column width="35%"}
::: {.incremental}
- Irrigated fields yield more
- The gap widens in dry years
- The spread is narrower under irrigation
:::
:::

::::
````

Two things to notice. The outer `::::` has four colons because it wraps the two
three-colon column blocks; the outer fence always needs strictly more colons
than the ones inside it. Getting this wrong produces a slide where the columns
do not appear and your content stacks vertically instead.

And `.incremental` wraps the *list*, not the slide, because it is a property of
the list.
:::

# Reveal things in an order you choose

`.incremental` walks down a list in order. Fragments let you decide what appears
when, and in what sequence.

**Task.**

1. Make a slide with three separate paragraphs, revealed one at a time.
2. Make one of them appear **second** even though it is written third.
3. Make one of them fade out again after being shown.

::: {.callout-important title="It worked if"}
The middle paragraph on screen is not the middle paragraph in your source file.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
## Fragments

::: {.fragment}
This appears first.
:::

::: {.fragment .fade-out}
This appears next, then fades away.
:::

::: {.fragment fragment-index=1}
This is written last but appears second.
:::
````

`fragment-index` sets the order explicitly and overrides source order. Other
useful variants are `.fade-in`, `.fade-out`, `.highlight-red`, and
`.semi-fade-out` for dimming a point rather than removing it.

Use these sparingly. One well-chosen reveal focuses attention; six of them make
an audience wait for a machine.
:::

# Rescue a slide with too much on it

**Task.**

1. Deliberately overload a slide: paste in a long code chunk and a large table
   so the content runs off the bottom.
2. Fix it two different ways, one slide each: shrink everything on the slide,
   and let the slide scroll.
3. Say which you would use for a table of results, and which for one long code
   chunk.

::: {.callout-important title="It worked if"}
In one version everything is visible at a smaller size. In the other the content
stays full size and scrolls with the mouse.
:::

::: {.callout-note collapse="true" title="Answer"}
```markdown
## A slide with a lot on it {.smaller}

## A slide with a lot on it {.scrollable}
```

`.smaller` is right when the audience needs the whole thing at once, for example
a results table where the comparison across rows *is* the point.

`.scrollable` is right when they only need one part at a time, for example
walking through a long function line by line.

`.scrollable` is a quiet trap in a live talk: anything below the fold is
invisible to everyone who is not driving, and presenters routinely forget it is
there. If the content genuinely does not fit, the honest fix is usually two
slides.
:::

# Walk an audience through code

Showing thirty lines of code at once tells an audience nothing. Highlighting
three lines at a time tells them everything.

**Task.**

1. Put a chunk of eight to ten lines on a slide, code only, not run.
2. Make it step through: first the whole thing dimmed, then lines 1 to 2, then
   line 5, then lines 7 to 9.
3. Add line numbers.

::: {.callout-important title="It worked if"}
Pressing the arrow key moves the highlight without advancing the slide.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
```{r}
#| eval: false
#| code-line-numbers: "|1-2|5|7-9"
corn <-
  read_csv(here("data", "corn_yields.csv"))

corn_summary <-
  corn %>%
  filter(Year == 2008) %>%
  group_by(Irrigated) %>%
  summarize(mean_yield = mean(Yield)) %>%
  arrange(desc(mean_yield))
```
````

The pipes in `code-line-numbers` separate the steps. A leading `|`, as here,
means "start with nothing highlighted", which is usually what you want so the
audience sees the shape of the code before you start pointing at it.

This is the single most useful revealjs feature for teaching code, and it is
worth reaching for whenever a chunk is longer than about five lines.
:::

# Tabsets

**Task.** Put three versions of the same figure on one slide, in tabs, so you
can flip between them while talking rather than jumping between slides.

::: {.callout-important title="It worked if"}
Clicking a tab changes the figure and the slide does not advance.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
## Three ways to show the same thing

::: {.panel-tabset}

### Boxplot

```{r}
#| echo: false
ggplot(corn) + geom_boxplot(aes(x = factor(Irrigated), y = Yield))
```

### Histogram

```{r}
#| echo: false
ggplot(corn) + geom_histogram(aes(x = Yield))
```

### Density

```{r}
#| echo: false
ggplot(corn) + geom_density(aes(x = Yield, fill = factor(Irrigated)))
```

:::
````

The `###` headings inside the tabset become the tab labels. Tabsets are how
these lecture notes fit a whole section onto one slide, and they are the main
tool for keeping a deck from sprawling.

One warning: tabs are invisible in a PDF export. Everything flattens into one
long slide. If the deck has to work as a handout, use them lightly.
:::

# Make the figures the right size

Figure sizing on slides confuses everybody once, because there are two different
things you can change and they do different jobs.

**Task.**

1. Put a figure on a slide and render. It is probably too big.
2. Shrink it by changing how large the *image file* is drawn.
3. Undo that, and shrink it instead by changing how much of the *slide* it
   occupies.
4. Compare the two results, looking specifically at the axis text.

::: {.callout-important title="It worked if"}
In one version the text shrinks with the figure. In the other it stays the same
size and the figure gets smaller around it.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
```{r}
#| fig-width: 5
#| fig-height: 3
#| echo: false
ggplot(corn) + geom_point(aes(x = FIPS, y = Yield))
```

```{r}
#| out-width: "60%"
#| echo: false
ggplot(corn) + geom_point(aes(x = FIPS, y = Yield))
```
````

`fig-width` and `fig-height` are in inches, and they set the size of the canvas
R draws on. Text is drawn at a fixed point size on that canvas, so a smaller
canvas means *relatively larger* text.

`out-width` scales the finished image on the slide, shrinking everything
together, text included.

For slides you almost always want `fig-width`/`fig-height`, because tiny axis
labels are the most common defect in a student deck and scaling with
`out-width` makes them worse. Draw a smaller figure rather than shrinking a
large one.
:::

# Make it look like the course decks

**Task.**

1. Copy `templates/custom.scss` and `templates/nebraska-n.jpg` from your
   `quarto-examples` clone into `quarto-practice`.
2. Apply the stylesheet to `slides.qmd`.
3. Add the image as a logo, and a footer with your name.
4. Open `custom.scss` in RStudio, change one colour, and re-render.

::: {.callout-important title="It worked if"}
The logo appears on every slide, including ones you never touched, and your
colour change is visible.
:::

::: {.callout-note collapse="true" title="Answer"}
```yaml
---
title: "Corn yields in Nebraska"
format:
  revealjs:
    theme: [default, custom.scss]
    logo: nebraska-n.jpg
    footer: "Your Name | AECN 896-05"
---
```

`theme: [default, custom.scss]` means "start from the default theme, then apply
my changes on top of it". Writing `theme: custom.scss` on its own throws the
default away, and most things immediately stop looking like anything.

Inside the scss file, the block between `/*-- scss:defaults --*/` and
`/*-- scss:rules --*/` is where variables like `$body-bg` and `$link-color`
live. Changing a variable there is usually easier than writing a new rule.
:::

# Speaker notes and a PDF

**Task.**

1. Add speaker notes to two slides.
2. Present the deck and open the speaker view. Confirm your notes appear there
   and not on the slide.
3. Export the whole deck to PDF.

::: {.callout-important title="It worked if"}
The PDF has one page per slide and your speaker notes are not in it.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
## A slide

Some content everyone can see.

::: {.notes}
Remember to explain why the 2012 drought shows up so clearly here.
:::
````

Press `s` while presenting for the speaker view. It opens a second window with
your notes, a preview of the next slide, and a timer. Put that on your laptop
and the deck on the projector.

For the PDF: add `?print-pdf` to the end of the deck's URL in Chrome, then use
the browser's print dialog and save as PDF. Incremental content and fragments
flatten onto a single page, which is usually what you want in a handout but is
worth checking before you distribute it.
:::

# Share a deck that survives the trip

**Task.**

1. Render the deck, then move **only** `slides.html` to your Desktop and open
   it.
2. Fix the problem, re-render, and repeat.

::: {.callout-important title="It worked if"}
The deck opens on the Desktop with its theme, logo, and figures intact, and no
folder alongside it.
:::

::: {.callout-note collapse="true" title="Answer"}
```yaml
format:
  revealjs:
    embed-resources: true
```

Same lesson as the html report in Ex-3-1, and it bites harder here, because a
deck depends on the reveal.js library itself and not just on your figures.
Without it, a deck sent by email arrives as a column of unstyled text.
:::

::: {.callout-warning collapse="true" title="If your deck is blank apart from the footer"}
You are opening the file directly off your disk and the browser is refusing to
load the scripts the deck needs. This is a browser security rule, not a mistake
on your part.

Either add `embed-resources: true`, or use `quarto preview slides.qmd` while you
are working, which serves the deck properly and reloads it every time you save.
:::

# Websites: add a page and wire it up

Start a fresh project for this one.

**Task.**

1. `File -> New Project -> New Directory -> Quarto Website`. Render it and look
   at the two pages you were given.
2. Open `_quarto.yml` and find the navbar section.
3. Create `analysis.qmd` with a heading, a sentence, and a figure.
4. Add it to the navbar so it appears in the menu.
5. Render the whole site and click your new link.

::: {.callout-important title="It worked if"}
Your page is reachable by clicking, not by typing its filename into the address
bar.
:::

::: {.callout-note collapse="true" title="Answer"}
```yaml
website:
  title: "My site"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: about.qmd
        text: About
      - href: analysis.qmd
        text: Analysis
```

Creating the file is not enough. `_quarto.yml` is what decides the structure of
the site, and a `.qmd` that nothing links to is a page nobody can find.

Notice also that the rendered site goes to a separate output folder. You edit
the `.qmd` files and never the rendered output, because the next full render
overwrites all of it.
:::

::: {.callout-tip collapse="true" title="Two extras worth trying"}
Group pages under a dropdown:

```yaml
      - text: Results
        menu:
          - href: analysis.qmd
            text: Analysis
          - href: robustness.qmd
            text: Robustness
```

And add `freeze: auto` under `execute:` in `_quarto.yml`. Without it, every
render re-runs every page on the site, which becomes unbearable at about page
ten.
:::
