---
title: "Ex-3-1: Importing and Exporting Files"
abstract: "Data Wrangling"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

None of these run in your browser. Reading and writing files is the one thing a
browser cannot do, so all of this happens in RStudio on your own machine. That
is the point of the lecture: code that only works on your computer is not much
use to anybody, and you cannot learn that lesson without a real computer.

Work through these in order. Several exercises use objects built in earlier
ones, and the last two are the ones that decide whether your assignments run.

# Setup {#sec-setup}

Do this once. Everything below depends on it.

1. Clone <https://github.com/tmieno2/data-science-course-supplementary-datasets>
   if you have not already. Lecture 02-0 shows you how.
2. Make a new folder called `ch3-practice` wherever you keep coursework.
3. In RStudio: `File -> New Project -> Existing Directory`, and choose
   `ch3-practice`. **This step is what makes `here()` work later.**
4. Inside `ch3-practice`, make two subfolders, `data` and `analysis`.
5. Copy these five files from the cloned repository into `ch3-practice/data`:
   `corn_yields.csv`, `corn_yields.xls`, `corn_yields.dta`, `corn_yields.rds`,
   and `corn_yields_exp_rownames.csv`.

You should end up with this:

```
ch3-practice/
├── ch3-practice.Rproj
├── analysis/
└── data/
    ├── corn_yields.csv
    ├── corn_yields.dta
    ├── corn_yields.rds
    ├── corn_yields.xls
    └── corn_yields_exp_rownames.csv
```

The data is corn yields for Nebraska counties: one row per county per
irrigation status, with `Yield` measured in bushels per acre.

::: {.callout-important title="It worked if"}
`here::here()` typed in the console prints the path to `ch3-practice`, and
nothing longer.
:::

::: {.callout-warning collapse="true" title="If it printed something else"}
`here()` looks upward from the current folder for a `.Rproj` file. If it printed
a parent folder, you either opened a different project, or you created the
project one level up. Close RStudio, double-click `ch3-practice.Rproj`, and
check again.

If it printed your home folder, you have no project open at all.
:::

# One dataset, four formats

The same corn yield data is stored four ways. Read all four and confirm you got
the same thing.

**Task.** In a new script, read each file and report `class()` and `dim()` of
each object.

::: {.callout-important title="It worked if"}
Three of them are 161 rows by 9 columns. **One of them is not, and you should
work out why before opening the answer.**
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
library(tidyverse)
library(readxl)   # NOT loaded by library(tidyverse)
library(haven)
library(here)

corn_csv <- read_csv(here("data", "corn_yields.csv"))
corn_xls <- read_excel(here("data", "corn_yields.xls"), sheet = 1)
corn_dta <- read_dta(here("data", "corn_yields.dta"))
corn_rds <- readRDS(here("data", "corn_yields.rds"))

lapply(list(corn_csv, corn_xls, corn_dta, corn_rds), dim)
lapply(list(corn_csv, corn_xls, corn_dta, corn_rds), class)
```

All four come back as 161 by 9 tibbles, so on the face of it nothing is wrong.
The catch is not visible in this output at all: `corn_yields.xls` has **two**
sheets, and by asking for `sheet = 1` you silently took half the data. The csv
holds 2008 only; the workbook holds 2008 and 2009 on separate sheets.

This is the failure mode worth internalising. A wrong file path throws an error
and you fix it in ten seconds. Reading the wrong *sheet* throws nothing, and you
find out three weeks later when your regression uses half the sample.
Exercise 4 deals with it properly.
:::

::: {.callout-warning collapse="true" title="could not find function \"read_excel\""}
`readxl` is installed by `tidyverse` but is **not loaded** by
`library(tidyverse)`, unlike `readr`, `dplyr`, and `ggplot2`. It needs its own
`library(readxl)`. This catches everybody exactly once.
:::

# `read.csv()` renames your columns

Open `corn_yields.csv` in a text editor, or in RStudio via
`File -> Open File`, and look at the seventh column heading. It is
`Data item`, with a space in it.

**Task.**

1. Read the file with `read.csv()` and with `read_csv()`.
2. Print `names()` of each and find the difference.
3. Extract that column from each object.

::: {.callout-important title="It worked if"}
You can state what `read.csv()` did to the name without being told, and the
extraction that works on one object fails on the other.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
base_df <- read.csv(here("data", "corn_yields.csv"))
tidy_tb <- read_csv(here("data", "corn_yields.csv"))

names(base_df)[7]   # "Data.item"  <- renamed behind your back
names(tidy_tb)[7]   # "Data item"  <- left alone

base_df$Data.item          # works
tidy_tb$`Data item`        # works, note the backticks
tidy_tb$Data.item          # NULL: that column does not exist
```

`read.csv()` replaces characters it considers illegal in a variable name with
dots. That is convenient right up to the moment you are working from a codebook
that says the variable is called `Data item` and you cannot find it in your
data.

`read_csv()` leaves names exactly as they were and you refer to awkward ones
with backticks. Backticks are worth learning now, because you will need them
again the moment a dataset has a column called `2008` or `GDP (millions)`.
:::

::: {.callout-tip collapse="true" title="Cleaning names properly"}
When every name in a file is awkward, do not fix them one at a time:

```{r}
#| eval: false
corn <- janitor::clean_names(read_csv(here("data", "corn_yields.csv")))
names(corn)   # year, state, fips, county_name, ..., data_item, irrigated, yield
```

`janitor::clean_names()` converts everything to lowercase with underscores.
Many people run it on every file they read, and it saves a lot of backticks.
:::

# A column you did not ask for

`corn_yields_exp_rownames.csv` was written out carelessly. Read it and look at
what came back.

**Task.**

1. Read it with both `read.csv()` and `read_csv()`. Compare `dim()` against the
   161 by 9 you got in Exercise 1.
2. Name the extra column in each case.
3. Get rid of it.
4. Work out which line of R code created this file.

::: {.callout-important title="It worked if"}
You end up with 9 columns, and you can name the argument that would have
prevented the problem in the first place.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
r_base <- read.csv(here("data", "corn_yields_exp_rownames.csv"))
r_tidy <- read_csv(here("data", "corn_yields_exp_rownames.csv"))

dim(r_base)        # 161 10  <- one column too many
names(r_base)[1]   # "X"
names(r_tidy)[1]   # "...1"

# drop it
r_tidy <- select(r_tidy, -1)
```

The file was written with `write.csv(data, "file.csv")` and no
`row.names = FALSE`, so R wrote the row numbers out as an unnamed first column.
When it is read back, that column of 1 to 161 becomes a variable.

The fix when writing: use `readr::write_csv()`, which never does this, or pass
`row.names = FALSE` to `write.csv()`.

Two names for the same problem are worth remembering, because you will meet
both: `X` from `read.csv()`, and `...1` from `read_csv()`, which is what readr
calls any column whose header is blank.
:::

# List the sheets before you read them

**Task.**

1. Run `excel_sheets()` on `corn_yields.xls` **before** reading anything.
2. Read each sheet by name, not by number.
3. Stack them into one dataset and check the result.

::: {.callout-important title="It worked if"}
Your combined dataset has 322 rows and `unique(Year)` returns both 2008 and
2009.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
excel_sheets(here("data", "corn_yields.xls"))
#> [1] "corn_yields_08" "corn_yields_09"

corn_08 <- read_excel(here("data", "corn_yields.xls"), sheet = "corn_yields_08")
corn_09 <- read_excel(here("data", "corn_yields.xls"), sheet = "corn_yields_09")

corn_all <- bind_rows(corn_08, corn_09)

dim(corn_all)            # 322 9
unique(corn_all$Year)    # 2008 2009
```

The sheets are called `corn_yields_08` and `corn_yields_09`, which is not what
anybody would have guessed. That is precisely why `excel_sheets()` comes first.

Note the asymmetry in how the two mistakes fail. Guessing a sheet *name* gets
you `Error: Sheet '2008' not found`, which is annoying and instantly fixable.
Guessing a sheet *number* gets you no error at all and half your data missing.
Always list, then read.
:::

::: {.callout-tip collapse="true" title="Reading every sheet at once"}
```{r}
#| eval: false
path <- here("data", "corn_yields.xls")

corn_all <-
  excel_sheets(path) %>%
  lapply(function(x) read_excel(path, sheet = x)) %>%
  bind_rows()
```

This is a preview of Chapter 5. It reads however many sheets there are, so it
does not break when someone adds 2010 to the workbook.
:::

# Spreadsheets made by human beings

Real spreadsheets almost never start with a tidy header in cell A1. There is a
title, a blank row, some notes off to the side, and a few cells containing `-`
where a number should be.

**Task.** Build one yourself so you can practise the fix.

1. Open `corn_yields.xls` in Excel or Numbers.
2. Insert three rows above the header. Put a title in the first, leave the
   second blank, and write a note in the third.
3. Replace three yield values with `-` and a couple more with `n/a`.
4. Save it as `corn_messy.xlsx` in your `data` folder.
5. Read it so that the result is identical in structure to the clean version:
   right header, and those cells read as `NA` rather than as text.

::: {.callout-important title="It worked if"}
`Yield` comes back numeric, not character, and `sum(is.na(corn$Yield))` matches
the number of cells you sabotaged.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
corn_messy <-
  read_excel(
    here("data", "corn_messy.xlsx"),
    sheet = 1,
    skip  = 3,                        # the title, the blank row, the note
    na    = c("", "NA", "n/a", "-")   # everything that means "missing"
  )

class(corn_messy$Yield)          # "numeric"
sum(is.na(corn_messy$Yield))     # however many cells you sabotaged
```

The diagnostic to remember: if a column you know is numeric comes back as
`character`, it is almost always because one cell contains something that is
not a number. A dash, an `n/a`, a footnote marker, a stray space. One cell in
ten thousand is enough to turn the whole column into text, and every arithmetic
operation you then attempt will fail.

Two other arguments worth knowing:

- `range = "B4:J165"` reads one rectangle and ignores notes elsewhere on the
  sheet
- `col_names = FALSE` for a sheet with no header at all, which gives you
  `...1`, `...2`, and so on to rename yourself
:::

# Stop `read_csv()` guessing

`read_csv()` decides each column's type by looking at the first thousand rows.
It is usually right. When it is wrong, it is wrong in a way you cannot undo
afterwards.

**Task.**

1. Create this file to see the problem clearly:

```{r}
#| eval: false
writeLines(
  c("county_id,yield",
    "01001,158",
    "01003,164",
    "31019,173"),
  here("data", "ids.csv")
)
```

2. Read it with `read_csv()` and look at `county_id`.
3. Read it again so that `county_id` keeps its leading zero.
4. Now do the same thing for `FIPS` in `corn_yields.csv`, and say why a FIPS
   code should never be stored as a number.

::: {.callout-important title="It worked if"}
`county_id` reads `01001`, not `1001`.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
bad <- read_csv(here("data", "ids.csv"))
bad$county_id      # 1001 1003 31019  <- the leading zero is gone forever

good <- read_csv(
  here("data", "ids.csv"),
  col_types = cols(county_id = col_character())
)
good$county_id     # "01001" "01003" "31019"

# same idea on the real file
corn <- read_csv(
  here("data", "corn_yields.csv"),
  col_types = cols(FIPS = col_character())
)
```

A FIPS code is an **identifier**, not a quantity. The average of two FIPS codes
is meaningless, and `01001` and `1001` are not the same county. Once R has
stored it as a number the zero is not recoverable, because it was never
written down.

The general rule: anything that identifies rather than measures should be
character. County codes, zip codes, well IDs, plot numbers, phone numbers.
If you would never add two of them together, it is not a number.
:::

::: {.callout-tip collapse="true" title="Other useful col_types shortcuts"}
```{r}
#| eval: false
# the compact form: one letter per column, in order
read_csv("file.csv", col_types = "cdid")   # character, double, integer, date

# set the type of a few and let readr guess the rest
read_csv("file.csv", col_types = cols(FIPS = col_character()))

# read absolutely everything as text, then convert deliberately
read_csv("file.csv", col_types = cols(.default = col_character()))
```

The last one is a genuinely useful defensive move with an unfamiliar file.
:::

# STATA files bring their labels with them

**Task.**

1. Read `corn_yields.dta` with `read_dta()`.
2. Print `County_name`. Compare it with `County_name` from the csv.
3. Make the STATA version show county names.

::: {.callout-important title="It worked if"}
`County_name` reads `BUFFALO`, not `8`.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
corn_dta <- read_dta(here("data", "corn_yields.dta"))

head(corn_dta$County_name)
#> <labelled<double>[6]>: County_name
#> [1]  8  8 19 19 21 21

head(corn_csv$County_name)
#> [1] "BUFFALO" "BUFFALO" "CUSTER" "CUSTER" ...

# convert every labelled column in one go
corn_dta <- haven::as_factor(corn_dta)
head(corn_dta$County_name)
#> [1] BUFFALO BUFFALO CUSTER CUSTER ...
```

STATA stores text variables as integers with a lookup table attached, which is
efficient and completely invisible inside STATA. `haven` imports both parts
faithfully, so you get numbers with the labels riding along as an attribute.
That is the honest translation, but it is not what you want to group by or
plot.

`as_factor()` applied to the whole tibble resolves every labelled column at
once. If you ever find yourself grouping by county and getting integers back,
this is why.
:::

# Straight from the internet

`read_csv()` accepts a URL anywhere it accepts a file path.

**Task.**

1. Read `corn_yields.csv` directly from the course dataset repository, without
   downloading it first:

```
https://raw.githubusercontent.com/tmieno2/data-science-course-supplementary-datasets/master/corn_yields.csv
```

2. Confirm it matches what you read from disk.
3. Then give two reasons not to build an analysis this way.

::: {.callout-important title="It worked if"}
`all.equal()` on the two objects returns `TRUE`.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
url <- "https://raw.githubusercontent.com/tmieno2/data-science-course-supplementary-datasets/master/corn_yields.csv"

corn_web <- read_csv(url)
all.equal(corn_web, corn_csv)   # TRUE
```

Note it must be the **raw** URL. The ordinary github.com page address gives you
a web page containing the data, not the data, and `read_csv()` will either
error or hand you a column of HTML.

Two reasons not to rely on it:

1. It re-downloads on every run, including every time you render. That is slow,
   and it fails entirely when you are on a plane.
2. A URL is somebody else's promise. Files get moved, repositories get renamed,
   agencies redesign their websites. When that happens your analysis stops
   working and the data you used is gone.

Download once, keep the file in your project, read the local copy, and cite the
URL in your paper. Reading from a URL is for exploring, not for anything you
intend to keep.
:::

# What a csv throws away

**Task.**

1. Take the tibble from `read_csv()` and add a factor column whose levels are
   deliberately **not** alphabetical.
2. Write it out twice, once with `write_csv()` and once with `saveRDS()`.
3. Read both back and compare `class()` and `levels()` of that column.

::: {.callout-important title="It worked if"}
The two round-trips disagree, and you can say what a csv file is physically
incapable of storing.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
corn2 <-
  corn_csv %>%
  mutate(
    irrigation = factor(
      ifelse(Irrigated == 1, "irrigated", "dryland"),
      levels = c("irrigated", "dryland")   # deliberately not alphabetical
    )
  )

write_csv(corn2, here("data", "roundtrip.csv"))
saveRDS(corn2,   here("data", "roundtrip.rds"))

from_csv <- read_csv(here("data", "roundtrip.csv"))
from_rds <- readRDS(here("data", "roundtrip.rds"))

class(from_csv$irrigation)    # "character"  <- the factor is gone
class(from_rds$irrigation)    # "factor"
levels(from_rds$irrigation)   # "irrigated" "dryland"  <- order survived
```

A csv is a text file. It stores the *values* and nothing else: no classes, no
factor level order, no attributes, no dates as dates. An rds stores the R
object itself, so everything survives exactly.

Use csv when a human or another program has to open it. Use rds for anything
you are handing back to R, especially intermediate results partway through an
analysis.

This matters more than it looks. In Chapter 4, the order of a factor's levels
is the order of the bars in your figure. Round-trip through csv and your
carefully ordered categories come back alphabetical.
:::

# rds holds any object, not just data

The lecture says an rds file stores "a single R object, not necessarily a
dataset". That distinction is worth proving to yourself.

**Task.**

1. Fit a regression of `Yield` on `Irrigated`.
2. Save the fitted model, not its output, to an rds file.
3. Restart R (`Session -> Restart R`), so your Environment is empty.
4. Read the model back and call `summary()` and `predict()` on it.
5. Then save a list containing the model, the data, and today's date, all in one
   file, and read that back too.

::: {.callout-important title="It worked if"}
After restarting R, you get a full regression summary without re-reading the
data or re-fitting anything.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
model <- lm(Yield ~ Irrigated, data = corn_csv)
saveRDS(model, here("data", "yield_model.rds"))

# --- Session -> Restart R ---

model <- readRDS(here("data", "yield_model.rds"))
summary(model)
predict(model, newdata = data.frame(Irrigated = c(0, 1)))

# a list of several things in one file
results <- list(
  model = model,
  data  = corn_csv,
  run_on = Sys.Date()
)
saveRDS(results, here("data", "results.rds"))

results <- readRDS(here("data", "results.rds"))
results$model
results$run_on
```

This is what makes rds different in kind from the other three formats. csv,
xlsx, and dta all store *tables*. An rds stores whatever you give it: a model,
a list, a figure, a function, a list of lists of models.

The practical use comes in Chapter 5. When one step of your analysis takes
twenty minutes, you run it once, save the result to rds, and every later script
reads that instead of re-running it.
:::

# The exercise that decides whether your assignment runs

This one is the whole reason the lecture spends a third of its time on paths.

**Task.**

1. In the console, run `read_csv("data/corn_yields.csv")`. It works.
2. Run `setwd("~")`, then run the same line again. Read the error.
3. Restart R (`Session -> Restart R`). Create `analysis/report.qmd` containing a
   chunk with that same line, and render it. Read the error.
4. Now rewrite the line using `here()` and confirm it works from the console
   **and** from the rendered qmd, with no other changes.

::: {.callout-important title="It worked if"}
One single line of code works from both places. If you had to write the path
one way in the qmd and another way in the console, you have not finished.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
# breaks depending on where you happen to be standing
read_csv("data/corn_yields.csv")

# works from anywhere inside the project, including a qmd in a subfolder
read_csv(here::here("data", "corn_yields.csv"))
```

Step 3 is the important one, and it surprises nearly everybody. When you render
a qmd, R sets the working directory to **the folder the qmd is in**, not the
folder the project is in. So `analysis/report.qmd` looks for
`analysis/data/corn_yields.csv`, which does not exist.

Your code worked interactively and failed on render. That is the single most
common way an assignment arrives broken, and it is why "but it works on my
machine" is not a defence: it did not even work on your machine, it worked in
your console.

`here()` builds every path from the project root, so it does not care which
folder the document asking the question lives in.
:::

::: {.callout-warning collapse="true" title="Why not just setwd() at the top of the qmd?"}
Two reasons. `setwd()` inside a chunk does not persist the way you expect
during a render, and knitr resets the directory between chunks. More
importantly it hard-codes a path that exists on exactly one computer in the
world, so the first line fails when anyone else opens your file.

An R Project plus `here()` solves both, and requires you to think about it
exactly once.
:::

# Read the error, do not guess

No code to write for this one. For each message below, say what it means and
what the *first* thing you would check is. All four are real messages, copied
verbatim from R.

**1.**
```
Error : 'corn_yields.csv' does not exist in current working directory
('/Users/taro/ch3-practice/analysis').
```

**2.**
```
Error in read_excel("corn_yields.xls") :
  could not find function "read_excel"
```

**3.**
```
Error: '\U' used without hex digits in character string (<input>:1:14)
```

**4.**
```
Error in readRDS("corn_yields.csv") : unknown input format
```

::: {.callout-note collapse="true" title="Answers"}
**1.** The file is not where R looked. Note that R tells you exactly where it
looked, in the brackets, and it is the `analysis` folder rather than the project
root. So this is a rendered qmd, or somebody ran `setwd()`. First check:
`getwd()`, then rewrite the path with `here()`.

**2.** The function does not exist *as far as R currently knows*. That means one
of exactly three things: the package is not loaded, the package is not
installed, or you misspelled the name. Here it is the readxl trap, installed by
tidyverse but not loaded by it. First check: is there a `library(readxl)` above
this line? Note that the message never suggests `library()`, so you have to
know to ask.

**3.** A Windows path written with single backslashes. R reads `\U` as the start
of an escape sequence, not as a folder separator. First check: the path string.
Either double the backslashes, `C:\\Users\\...`, or use forward slashes,
`C:/Users/...`, or stop writing the top of the path at all and use `here()`.
The character position at the end, `1:14`, points at where R gave up.

**4.** The right function pointed at the wrong format. `readRDS()` opened the
file, found text where it expected a compressed R object, and gave up. This is
the lecture's central rule failing in practice: the function must match the
format, and no function reads everything. First check: the file extension
against the function name. Here it is `.csv`, so it wants `read_csv()`.

Worth noticing how much less helpful this message is than the first three. It
does not name the file, does not say what format it expected, and does not
suggest anything. Some errors tell you what to do and some only tell you that
something is wrong; the second kind is why the extension-to-function mapping is
worth memorising rather than guessing at.
:::
