---
title: "Ex-5-1: Functions, Loops, and Parallelization"
abstract: "Function and Loop"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

The deck already gives you browser exercises on writing functions and on
`lapply()`. This handout covers the half that a browser cannot do: reading a
folder full of files, timing one approach against another, and running things on
more than one core. All of it happens in RStudio.

Timing is the point of several of these exercises, and timing in a browser is
meaningless. Parallel processing in a browser is not merely slow, it is
unavailable.

# Setup {#sec-setup}

1. Find your `quarto-examples` clone from lecture 02-0. It contains
   `data/data-for-loop-demo/`, which holds 60 files: 30 corn experiments and 30
   soybean experiments.
2. Make a project folder `ch5-practice`, open it as an RStudio Project, and copy
   that whole `data-for-loop-demo` folder into it.
3. Install what you need:

```{r}
#| eval: false
install.packages(c("future.apply", "microbenchmark", "tictoc"))
```

::: {.callout-important title="It worked if"}
`length(list.files(here::here("data-for-loop-demo")))` returns 60.
:::

# Look before you loop

Never write a loop over files you have not opened one of.

**Task.**

1. List the corn files only, with their full paths.
2. Read the first one and report its dimensions and column names.
3. Read the first **soybean** file and compare its column names with the corn
   one. Something is wrong. Say what.

::: {.callout-important title="It worked if"}
You have 30 corn paths, and you can state the problem with the soybean files
before you write any loop at all.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
library(tidyverse)
library(here)

corn_files <-
  list.files(
    here("data-for-loop-demo"),
    pattern    = "corn_experiment",
    full.names = TRUE
  )

length(corn_files)   # 30

first <- readRDS(corn_files[1])
dim(first)           # 1000 4
names(first)         # "N_rate" "v" "corn_yield" "field_id"

soy_first <- readRDS(here("data-for-loop-demo", "soy_experiment_1.rds"))
names(soy_first)     # "N_rate" "v" "corn_yield" "field_id"   <- corn_yield!
```

The soybean files have a column called `corn_yield`. The column holds soybean
yields; only the name is wrong.

This is worth pausing on, because it is not a trick. Datasets you are handed
routinely have names inherited from whatever script produced them, and nothing
in R will ever warn you. If you bound the corn and soybean files together
trusting the names, you would have 60,000 rows of "corn yield", 30,000 of which
are soybeans, and every number after that point would be wrong while looking
entirely plausible.

`full.names = TRUE` matters too. Without it you get bare filenames, which are
not paths, and every `readRDS()` fails.
:::

# Read 30 files without writing 30 lines

**Task.**

1. Use `lapply()` to read all 30 corn files into a list.
2. Combine them into one dataset.
3. Check the row count, then check how many distinct fields you have.
4. That last number is wrong. Work out why, and fix it so each row knows which
   file it came from.

::: {.callout-important title="It worked if"}
Your combined dataset has 30,000 rows and 30 distinct field identifiers.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
# the naive version
corn_all <-
  lapply(corn_files, readRDS) %>%
  bind_rows()

nrow(corn_all)                       # 30000, correct
n_distinct(corn_all$field_id)        # 1  <- wrong
```

Every file contains `field_id = 1`. The identifier is not unique across files,
so once you bind them you can no longer tell the 30 experiments apart. Thirty
thousand rows that all claim to be field 1.

The fix is to attach the identifier as you read, because that is the only moment
you know which file you are holding:

```{r}
#| eval: false
read_one <- function(path) {
  readRDS(path) %>%
    mutate(
      source_file = basename(path),
      field       = parse_number(basename(path))
    )
}

corn_all <-
  lapply(corn_files, read_one) %>%
  bind_rows()

n_distinct(corn_all$field)   # 30
```

The general principle: whatever distinguishes one iteration from another has to
be recorded *inside* the iteration. Once the loop is over, that information is
gone. This is the single most common bug in file-reading loops, and it is silent.
:::

::: {.callout-tip collapse="true" title="The tidyverse shortcut"}
```{r}
#| eval: false
corn_all <-
  corn_files %>%
  set_names(basename(.)) %>%
  purrr::map(readRDS) %>%
  bind_rows(.id = "source_file")
```

`bind_rows(.id =)` takes the list's names and writes them into a column, which
handles the identifier problem for you. Worth knowing, but write the explicit
version at least once first, because the explicit version is what tells you what
`.id` is actually doing.
:::

# One function, then loop it

**Task.** Write a function that takes a file path and returns a **one-row**
summary of that experiment: the field number, the number of observations, the
mean yield, and the yield at the highest nitrogen rate applied.

Then use it to build a 30-row summary table across all corn fields, and find
the three fields with the highest mean yield.

::: {.callout-important title="It worked if"}
The result has exactly 30 rows and one row per file.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
summarize_one <- function(path) {
  d <- readRDS(path)

  tibble(
    field      = parse_number(basename(path)),
    n_obs      = nrow(d),
    mean_yield = mean(d$corn_yield),
    yield_at_max_n = d$corn_yield[which.max(d$N_rate)]
  )
}

corn_summary <-
  lapply(corn_files, summarize_one) %>%
  bind_rows()

corn_summary %>% arrange(desc(mean_yield)) %>% head(3)
```

Two habits worth copying. The function takes **one** path and returns **one**
row, so you can test it on a single file before trusting it on thirty. And it
returns a tibble rather than a vector, so `bind_rows()` produces a proper table
with named columns.

Test it on one file first, every time:

```{r}
#| eval: false
summarize_one(corn_files[1])
```

If that is wrong, the loop is wrong thirty times over, and it is far harder to
see the problem in thirty wrong answers than in one.
:::

# Loop over two things at once

**Task.** Build a summary across **both** crops: 60 rows, one per file, with a
`crop` column distinguishing them. Do it without copying your loop twice.

While you are at it, rename that misleading `corn_yield` column to `yield`.

::: {.callout-important title="It worked if"}
60 rows, `count(crop)` gives 30 and 30, and no column is called `corn_yield`.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
# every combination of crop and field number
grid <- expand.grid(
  crop  = c("corn", "soy"),
  field = 1:30,
  stringsAsFactors = FALSE
)

summarize_case <- function(i) {
  crop  <- grid$crop[i]
  field <- grid$field[i]

  path <- here("data-for-loop-demo", paste0(crop, "_experiment_", field, ".rds"))

  readRDS(path) %>%
    rename(yield = corn_yield) %>%     # the name is wrong in the soy files
    summarize(
      crop       = crop,
      field      = field,
      mean_yield = mean(yield),
      .groups    = "drop"
    )
}

all_summary <-
  lapply(seq_len(nrow(grid)), summarize_case) %>%
  bind_rows()

count(all_summary, crop)   # corn 30, soy 30
```

The pattern to take away: when you need to loop over combinations, build a
**data frame of the combinations first**, then loop over its row numbers. Your
function takes one integer, looks up what that row means, and does the work.

This scales to three or four varying inputs without nesting loops, and it makes
the parallel version in exercise 7 a one-word change. Nested loops do neither.
:::

# Do you actually need the loop?

**Task.**

1. Compute profit at 1,000 nitrogen rates using a `for` loop that grows a
   vector with `c()`.
2. Do the same with `lapply()`.
3. Do the same with a single vectorized expression.
4. Time all three on 1,000 values, then on 100,000.

Yield is `120 + 25 * log(N)`. Profit is `4 * yield - 0.5 * N`.

::: {.callout-important title="It worked if"}
All three give identical answers, and the gap between them widens dramatically
when you go from 1,000 to 100,000.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
library(microbenchmark)

profit_of <- function(N) 4 * (120 + 25 * log(N)) - 0.5 * N

N_seq <- seq(1, 300, length = 1000)

#--- 1. grow a vector: the slow way ---#
loop_grow <- function(N_seq) {
  out <- c()
  for (i in seq_along(N_seq)) out <- c(out, profit_of(N_seq[i]))
  out
}

#--- 2. lapply ---#
via_lapply <- function(N_seq) unlist(lapply(N_seq, profit_of))

#--- 3. vectorized ---#
vectorized <- function(N_seq) profit_of(N_seq)

all.equal(loop_grow(N_seq), vectorized(N_seq))   # TRUE

microbenchmark(
  loop_grow(N_seq),
  via_lapply(N_seq),
  vectorized(N_seq),
  times = 10
)
```

The vectorized version wins by a wide margin, and the margin grows with the
input, because `profit_of()` was already vectorized. Calling it once on 1,000
values does the work in compiled code; calling it 1,000 times on one value pays
R's function-call overhead 1,000 times.

The `c()` version is the worst, and worth understanding separately. Each `c()`
allocates a **new** vector and copies everything into it, so building a vector
of length n does work proportional to n². At 1,000 it is merely bad; at 100,000
you will wait. If you must loop, pre-allocate:

```{r}
#| eval: false
out <- numeric(length(N_seq))
for (i in seq_along(N_seq)) out[i] <- profit_of(N_seq[i])
```

The rule: before parallelizing anything, check whether you should be looping at
all. Vectorizing is usually a bigger win than parallelizing, and it is free.
:::

# Parallelize it

**Task.**

1. Find out how many cores you have.
2. Set up parallel processing with `plan()`.
3. Rerun the 60-file summary from exercise 5 with `future_lapply()` instead of
   `lapply()`, and time both.
4. Then explain why the speedup is disappointing.

::: {.callout-important title="It worked if"}
The parallel version produces an identical 60-row result.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
library(future.apply)
library(tictoc)

parallel::detectCores()

plan(multisession, workers = 6)   # leave yourself a couple of cores

tic()
seq_ver <- bind_rows(lapply(seq_len(nrow(grid)), summarize_case))
toc()

tic()
par_ver <- bind_rows(future_lapply(seq_len(nrow(grid)), summarize_case))
toc()

all.equal(seq_ver, par_ver)   # TRUE
```

`lapply` becomes `future_lapply` and nothing else changes. That really is the
whole API.

Why the speedup disappoints: each of these tasks takes a few milliseconds, and
starting a worker session, shipping the data to it, and collecting the result
back costs more than that. Parallelization has a fixed overhead per task, so it
pays only when each task is slow enough to dwarf it. Sixty fast tasks is exactly
the case where it does not.

Exercise 8 gives you a workload where it does pay.
:::

::: {.callout-warning collapse="true" title="The multicore trap"}
`plan(multicore)` uses forking, which does not exist on Windows and is disabled
inside RStudio on every platform.

Ask for it where it is unavailable and R does **not** warn you and does **not**
fall back to something else. It quietly runs everything sequentially. Your code
works, your timings are unchanged, and you conclude that parallelization is
overrated.

Use `plan(multisession)`. It works everywhere.
:::

::: {.callout-tip collapse="true" title="Shut it down"}
`plan(sequential)` when you are finished. Worker sessions hold memory, and six
idle copies of a large dataset will make your machine unhappy.
:::

# Grid search, and a workload worth parallelizing

Recall the problem from the deck. Corn yield responds to nitrogen as
`120 + 25 * log(N)`, and profit is `Pc * yield - Pn * N`. You want the
profit-maximizing nitrogen rate.

**Task.**

1. Find the best rate by grid search over `N` from 1 to 300, for `Pc = 4` and
   `Pn = 0.5`.
2. Check your answer against the closed-form solution.
3. Now do it for **every** combination of 40 corn prices and 40 nitrogen prices,
   sequentially. Time it.
4. Do the same in parallel. Time it. This time it should pay.

::: {.callout-important title="It worked if"}
Step 1 gives approximately 200, which is exactly what calculus predicts, and
step 4 is meaningfully faster than step 3.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
#--- 1. one grid search ---#
best_n <- function(Pc, Pn, N_grid = seq(1, 300, length = 2000)) {
  profit <- Pc * (120 + 25 * log(N_grid)) - Pn * N_grid
  N_grid[which.max(profit)]
}

best_n(4, 0.5)   # about 200
```

**Checking it.** Differentiate profit with respect to `N`:

$$
\frac{d}{dN}\left[P_C(120 + 25\log N) - P_N N\right] = \frac{25 P_C}{N} - P_N = 0
\quad\Longrightarrow\quad N^* = \frac{25 P_C}{P_N}
$$

With `Pc = 4` and `Pn = 0.5`, that is `25 * 4 / 0.5 = 200`. Your grid search
should land within a fraction of a unit of it, and the gap tells you how coarse
your grid is.

Always check a numerical answer against something you can derive, whenever you
can derive anything. A grid search that quietly returns the edge of its own grid
looks exactly like a grid search that worked.

```{r}
#| eval: false
#--- 3 and 4. many price combinations ---#
prices <- expand.grid(
  Pc = seq(3, 7, length = 40),
  Pn = seq(0.3, 0.9, length = 40)
)

solve_row <- function(i) {
  tibble(
    Pc = prices$Pc[i],
    Pn = prices$Pn[i],
    N_star = best_n(prices$Pc[i], prices$Pn[i],
                    N_grid = seq(1, 300, length = 20000))
  )
}

tic(); seq_res <- bind_rows(lapply(seq_len(nrow(prices)), solve_row));        toc()
tic(); par_res <- bind_rows(future_lapply(seq_len(nrow(prices)), solve_row)); toc()
```

1,600 grid searches over 20,000 candidate rates each is enough work per task
that the overhead stops mattering, and the parallel version wins.

That is the general shape of when to parallelize: many tasks, each slow, none
depending on the others. If your tasks are fast, vectorize instead. If they
depend on each other, you cannot parallelize them at all.
:::

::: {.callout-tip collapse="true" title="Plot the answer"}
```{r}
#| eval: false
ggplot(par_res) +
  geom_raster(aes(x = Pc, y = Pn, fill = N_star)) +
  scale_fill_viridis_c() +
  labs(x = "Corn price ($/bu)", y = "Nitrogen price ($/lb)",
       fill = "Optimal N\n(lb/acre)")
```

The result should be a smooth surface. If yours has stripes or blocks in it, the
grid is too coarse and neighbouring price combinations are snapping to the same
grid point.
:::

# Put it together

**Task.** Write one script that, for all 60 experiment files:

- reads each file, attaching crop and field identifiers
- renames the yield column correctly
- fits `yield ~ log(N_rate)` for each field separately
- uses each field's own fitted coefficients to find its profit-maximizing
  nitrogen rate at `Pc = 4`, `Pn = 0.5`
- returns one tidy 60-row table
- runs the expensive part in parallel

Then save the result as an rds and make a figure of optimal rate by crop.

::: {.callout-important title="It worked if"}
It runs end to end from `Session -> Restart R`, and the sequential and parallel
versions give identical results.
:::

::: {.callout-note collapse="true" title="A hint on structure"}
Build the combination table first, write a function that handles **one** row of
it, test that function on row 1, and only then wrap it in `future_lapply()`.

That order matters. Debugging inside a parallel loop is genuinely unpleasant:
error messages come back from a worker session with no context, and printed
output may not come back at all. Get it right on one case sequentially, then
parallelize a function you already trust.
:::
