---
title: "Ex-6-1: Organizing a Reproducible Project"
abstract: "Code Projects"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

There is no code to run in most of this. That is not a defect of the exercises:
organizing a project is not a coding skill, and you cannot practise it in a
browser or on a toy example. What you can do is build a real structure once,
deliberately, and then use it for the rest of the course.

The test that every exercise here is working toward is a single question. **If
you sent this folder to a stranger, could they reproduce your results without
asking you anything?** Everything below is a way of making the answer yes.

# Build the structure

**Task.** Create a project called `nitrogen-study` with the folder layout the
lecture recommends.

1. `File -> New Project -> New Directory -> New Project`, named
   `nitrogen-study`.
2. Create the folders below. Make them all now, even the ones you have nothing
   to put in yet.
3. Put a `.gitkeep` file, or a short `readme.md`, in each empty folder.

```
nitrogen-study/
├── nitrogen-study.Rproj
├── Code/
│   ├── DataPrep/
│   └── Analysis/
├── Data/
│   ├── Raw/
│   └── Processed/
├── Literature/
├── Results/
└── Writing/
```

::: {.callout-important title="It worked if"}
`fs::dir_tree()` or the Files pane shows all nine folders, and
`here::here()` prints the `nitrogen-study` path.
:::

::: {.callout-note collapse="true" title="Doing it from the console"}
```{r}
#| eval: false
library(here)

dirs <- c(
  "Code/DataPrep", "Code/Analysis",
  "Data/Raw", "Data/Processed",
  "Literature", "Results", "Writing"
)

for (d in dirs) dir.create(here(d), recursive = TRUE, showWarnings = FALSE)

# stop empty folders disappearing when you share the project
for (d in dirs) file.create(here(d, ".gitkeep"))

fs::dir_tree(here())
```

The empty-folder point is not pedantry. Git does not track folders, only files,
so an empty `Data/Raw` will not survive being cloned. Someone who clones your
project then gets a script that writes to a folder that does not exist, and a
confusing error.

Why create folders you have nothing for yet? Because the alternative is deciding
where something goes at the moment you are busy with it, which is when you will
decide badly.
:::

# Sort out somebody else's mess

This is how most projects actually arrive: everything in one folder, named
however it happened.

**Task.** Create this mess in a scratch folder, then reorganize it into your
structure without looking at the answer.

```{r}
#| eval: false
mess <- here::here("mess")
dir.create(mess, showWarnings = FALSE)

files <- c(
  "data.csv", "data_v2.csv", "data_v2_FINAL.csv",
  "analysis.R", "analysis_new.R", "Untitled1.R",
  "plot.png", "figure for paper.png",
  "notes.docx", "Smith et al 2019.pdf",
  "cleaned.rds", "regression output.txt",
  "script that makes the map.R", "final_paper.qmd"
)

file.create(file.path(mess, files))
```

For each file, decide which folder it belongs in and what it should be called.
Three of them you cannot place from the name alone. Identify which, and say what
you would need to know.

::: {.callout-important title="It worked if"}
Every file has a destination and a new name, and you can state the run order of
the code files from their names alone.
:::

::: {.callout-note collapse="true" title="One reasonable answer"}
| Original | Goes to | Renamed |
|---|---|---|
| `data.csv`, `data_v2.csv`, `data_v2_FINAL.csv` | `Data/Raw/` | **cannot decide from the name** |
| `cleaned.rds` | `Data/Processed/` | `nitrogen-cleaned.rds` |
| `analysis.R`, `analysis_new.R` | `Code/Analysis/` | **cannot decide from the name** |
| `Untitled1.R` | probably nowhere | delete, or open it and find out |
| `script that makes the map.R` | `Code/Analysis/` | `02-3-make-map.R` |
| `plot.png`, `figure for paper.png` | `Results/` | `fig-yield-response.png` |
| `regression output.txt` | `Results/` | `reg-main.txt` |
| `Smith et al 2019.pdf` | `Literature/` | keep |
| `notes.docx` | `Literature/` or `Writing/` | keep |
| `final_paper.qmd` | `Writing/` | `manuscript.qmd` |

The three you cannot place are the three sets of near-duplicates:
`data` / `data_v2` / `data_v2_FINAL`, and `analysis` / `analysis_new`. Nothing in
the names tells you which is current. You have to open them, compare them, and
guess, and if the author is not available you may never know.

That is the actual cost of bad naming, and it is why `_FINAL` is a warning sign
rather than a solution. Two files whose names differ only by `_new` are a
question you have left for your future self, who will have forgotten the answer.

Note also what disappears once files are named for the order they run in.
`01-1-`, `01-2-`, `02-1-` is not decoration: it is the instruction for
reproducing the project, written where nobody can lose it.

Spaces in filenames are a separate small nuisance. They are legal, and they
require quoting in every shell command and break some tools outright. Use
hyphens.
:::

# Name files for the order they run in

**Task.** Your study needs the following steps. Give each one a filename and put
it in the right folder.

1. download county-level weather data from an API
2. download county boundaries
3. clean the nitrogen trial data
4. merge trial data with weather
5. estimate the main regression
6. run a robustness check with a different specification
7. produce the figures and tables for the paper

Then write a `readme.md` at the project root telling a stranger what to run.

::: {.callout-important title="It worked if"}
Sorting `Code/DataPrep` and `Code/Analysis` alphabetically gives you the correct
execution order, with no readme required to work it out.
:::

::: {.callout-note collapse="true" title="Answer"}
```
Code/DataPrep/
  01-1-download-weather-data.R
  01-2-download-county-boundaries.R
  01-3-clean-trial-data.R
  01-4-merge-trial-weather.R

Code/Analysis/
  02-1-main-regression.R
  02-2-robustness-alt-spec.R
  02-3-gen-figures-tables.R
```

The numbering carries two pieces of information: the `01` versus `02` says which
stage, and the `-1`, `-2` says the order within it. Alphabetical sort equals
execution order, which means the folder listing *is* the documentation.

A readme worth writing:

````markdown
# Nitrogen study

## What this is
Estimates corn yield response to nitrogen using on-farm trial data,
2020-2023.

## How to reproduce
1. Open `nitrogen-study.Rproj` in RStudio.
2. Run `Code/DataPrep/` in filename order. Requires an API key in
   `.Renviron` as `NOAA_KEY`.
3. Run `Code/Analysis/` in filename order.
4. Render `Writing/manuscript.qmd`.

Total runtime is about 20 minutes; step 01-1 is most of it.

## Data
`Data/Raw/` is read-only and is not regenerated by any script.
See `Data/Raw/metadata.md` for sources and download dates.
````

The API key line is the sort of thing that gets forgotten and makes a project
unreproducible in a way that looks like a bug. Note what is written down and
what is deliberately *not*: the key itself never goes in the repository.
:::

# Make raw data read-only, and mean it

The lecture's first rule is that you never overwrite raw data. Here you enforce
it rather than remembering it.

**Task.**

1. Put a csv in `Data/Raw/`. Any of the corn yield files will do.
2. Write `Data/Raw/metadata.md` recording where the file came from, when you
   downloaded it, and what the columns mean.
3. Make the file read-only at the operating-system level.
4. Try to overwrite it from R and observe what happens.
5. Write a `Code/DataPrep` script that reads it, changes something, and saves
   the result to `Data/Processed/`.

::: {.callout-important title="It worked if"}
Writing to the raw file **fails with an error**, and your processed file appears
in `Data/Processed/`.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
raw <- here("Data", "Raw", "corn_yields.csv")

#--- make it read-only ---#
Sys.chmod(raw, mode = "0444")

#--- now try to clobber it ---#
readr::write_csv(data.frame(x = 1), raw)
#> Error: Permission denied

#--- the right shape for a DataPrep script ---#
corn <- readr::read_csv(raw)

corn_clean <-
  corn %>%
  janitor::clean_names() %>%
  dplyr::filter(!is.na(yield))

saveRDS(corn_clean, here("Data", "Processed", "corn-clean.rds"))
```

Why bother, when you could simply be careful? Because being careful is a
promise you make to yourself at the moment you have the least attention. The
`0444` makes it impossible instead of merely inadvisable, and it costs one line.

Raw data is the only thing in the project you cannot regenerate. Every script,
every figure, every processed dataset can be rebuilt by rerunning code. If you
overwrite `Data/Raw/`, you have destroyed the irreplaceable part.

`metadata.md` is the other half. A csv with no provenance is nearly worthless
six months later, because you cannot answer the first question a referee asks:
where did this come from and when?
:::

# Read your own code as a stranger

**Task.** Here is some code that works. Reformat it so a person can read it.

```{r}
#| eval: false
d<-read.csv("Data/Raw/corn_yields.csv");d2<-d[d$Yield>100&d$Irrigated==1,]
res=lm(Yield~Irrigated+FIPS,data=d);summary(res)
Final_Data_2<-aggregate(d2$Yield,by=list(d2$County_name),FUN=mean)
```

1. Fix it by hand first: one statement per line, spaces around operators,
   consistent assignment, names that say what things are.
2. Then install `styler` and use `Addins -> Style selection` on the original.
3. Compare. Name two things you fixed that `styler` did not.

::: {.callout-important title="It worked if"}
You can name something wrong with this code that no automatic tool can fix.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
corn <- read.csv(here("Data", "Raw", "corn_yields.csv"))

irrigated_high_yield <-
  corn %>%
  dplyr::filter(Yield > 100, Irrigated == 1)

yield_model <- lm(Yield ~ Irrigated + FIPS, data = corn)
summary(yield_model)

mean_yield_by_county <-
  irrigated_high_yield %>%
  dplyr::group_by(County_name) %>%
  dplyr::summarize(mean_yield = mean(Yield))
```

`styler` fixes spacing, indentation, and line breaks. It will not fix:

- **the names.** `d`, `d2`, `res`, and `Final_Data_2` mean nothing. `styler`
  cannot know that `d2` is the irrigated high-yield subset, so it leaves the
  name alone. Naming is the part that carries the meaning, and it is entirely on
  you.
- **`=` versus `<-`.** Mixing them is legal and inconsistent. Pick `<-`.
- **the actual bug.** `lm(Yield ~ Irrigated + FIPS)` treats `FIPS` as a
  *number*, so the model claims yield rises linearly with county code. It should
  be `factor(FIPS)`. No formatter will ever tell you this, and unreadable code
  is exactly where this kind of error hides.

That last point is the argument for style. Readable code is not a courtesy, it
is how you find your own mistakes.
:::

::: {.callout-tip collapse="true" title="Setting up styler"}
```{r}
#| eval: false
install.packages("styler")
```

Then highlight code and use `Addins -> Style selection`. Addins have no keyboard
shortcut by default; assign one under
`Tools -> Modify Keyboard Shortcuts...` by searching for "Style selection".

Do not confuse this with `cmd`/`ctrl` + `shift` + `A`, which is RStudio's own
Reformat Code. That is built in, works without `styler`, and does not follow the
tidyverse style.
:::

# Stop typing the same thing

**Task.**

1. Open `Tools -> Edit Code Snippets...` and read a few of the R ones.
2. Write a snippet `hdr` that inserts a commented section divider.
3. Write a snippet `rdd` that inserts a `readRDS(here(...))` call with the
   cursor left in the right place.
4. Write one that inserts your standard qmd setup chunk.

::: {.callout-important title="It worked if"}
Typing the shortcut and pressing Tab inserts the text with your cursor where you
want it.
:::

::: {.callout-note collapse="true" title="Answer"}
Snippet definitions are indented with **tabs**, not spaces. This is the one thing
that goes wrong, and it fails silently.

````{verbatim}
snippet hdr
	# /*=================================================*/
	#' # ${1:section name}
	# /*=================================================*/

snippet rdd
	readRDS(here::here("Data", "Processed", "${1:file}.rds"))

snippet qsetup
	```{r}
	#| label: setup
	#| include: false
	library(tidyverse)
	library(here)
	```
````

`${1:name}` is a placeholder: the cursor lands there, the text is pre-selected,
and Tab moves to `${2:...}` if there is one. That is what makes a snippet worth
more than copy-paste.

These are per-user, not per-project, so they follow you to every project you
open. Ten minutes here is repaid for years.
:::

# Audit a real project

**Task.** Clone <https://github.com/tmieno2/Sample-Reproducible-Project> and
read it the way somebody reproducing it would, not the way an author would.

Answer these without running anything:

1. What is this project about, and how quickly could you tell?
2. Where is the instruction telling you what to run, and in what order?
3. Which folder holds data you could not regenerate?
4. Pick one number in the writing. Trace it back to the code that produced it
   and the data that fed it.
5. What would you need from the author that the repository does not give you?

::: {.callout-important title="It worked if"}
You completed step 4. If you cannot trace one number back to its source, either
the project is not reproducible or you have not finished reading it.
:::

::: {.callout-note collapse="true" title="What to look for"}
Step 4 is the whole exercise. Everything else is orientation.

Tracing a number backwards is the only test that actually distinguishes a
project that *looks* organized from one that is reproducible. A tidy folder tree
proves nothing; a number you can follow from the manuscript, to the script that
computed it, to the dataset it came from, proves everything.

You may also notice `renv.lock` in the repository. That records the exact version
of every package used, so a future R can reinstall the same versions. It is the
answer to the failure mode none of the folder structure addresses: your code is
fine, your data is fine, and a package changed under you.

Step 5 is worth being honest about. Almost every real project has a gap: an API
key, a licensed dataset, a manual step someone did in Excel once. The point is
not that gaps are shameful, it is that they should be **written down** rather
than discovered.
:::

# The only test that counts

**Task.** Take a project you have actually done — an assignment from this course
will do — and reorganize it into this structure. Then:

1. Rename every file so the run order is obvious.
2. Make `Data/Raw` read-only and write its `metadata.md`.
3. Write the root `readme.md`.
4. Delete everything in `Data/Processed`, `Results`, and every rendered output.
5. Restart R, and rebuild the whole project from raw data by following your own
   readme.
6. Swap projects with a classmate and reproduce each other's.

::: {.callout-important title="It worked if"}
Step 5 rebuilds everything with no manual intervention and no errors, and your
classmate gets your numbers in step 6 without messaging you once.
:::

::: {.callout-warning collapse="true" title="What usually breaks at step 5"}
In rough order of frequency:

- an absolute path (`/Users/yourname/...`) that exists on one machine
- a script that depends on an object another script left in the Environment,
  rather than reading it from `Data/Processed`
- `library()` calls for packages that are not installed on a clean machine
- a step that was done by hand once and never written down
- a file in `Results/` that no script produces, because you made it
  interactively and forgot

Every one of these is invisible until you delete the outputs and try again. That
is why step 4 is in the list, and why "it runs on my machine" is not evidence of
anything.
:::
