---
title: "Ex-7-1: Working with Dates"
abstract: "Date and String"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
webr:
  packages: ['dplyr', 'ggplot2', 'lubridate', 'gt']
  cell-options:
    editor-font-scale: 0.7
    out-width: 100%
filters:
  - webr
---

These run in your browser. Click **Run Code** in a cell to execute it, or
highlight part of it and press `cmd`/`ctrl` + Enter.

We use `pizzaplace` from the `gt` package: a year of pizza orders from 2015,
with a date, a time, a name, a size, and a price. It is a good dataset for this
lecture because its `date` column arrives as **text**, which is exactly the
problem the whole chapter exists to solve.

# Data

Run this first. Every exercise below uses `pizza`.

```{webr-r}
#| autorun: true
data(pizzaplace, package = "gt")

pizza <- pizzaplace

pizza
```

# It looks like a date, but it is not

**Task.**

1. Check the class of `pizza$date`.
2. Try to find the earliest and latest order date with `min()` and `max()`. Look
   carefully at whether the answer is trustworthy.
3. Try to work out how many days the shop was open by subtracting one from the
   other. Read the error.
4. Convert the column to a real `Date` and repeat step 3.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
class(pizza$date)          # "character"

min(pizza$date)            # "2015-01-01" -- looks right!
max(pizza$date)            # "2015-12-31"

max(pizza$date) - min(pizza$date)
#> Error in max(pizza$date) - min(pizza$date) :
#>   non-numeric argument to binary operator

pizza <- dplyr::mutate(pizza, date = lubridate::ymd(date))

class(pizza$date)                       # "Date"
max(pizza$date) - min(pizza$date)       # Time difference of 364 days
```

Step 2 is the trap worth dwelling on. `min()` and `max()` on character strings
compare them **alphabetically**, and they returned the right answer here only
because `YYYY-MM-DD` happens to sort the same way alphabetically as it does
chronologically.

That is luck, not correctness. Had the dates been written `01/01/2015`, the
alphabetical minimum would have been the first of *January of any year*, and
your answer would have been silently wrong with no error at all.

Only step 3 tells you something is amiss, and only because subtraction is
undefined for text. Never trust that a column is a date because it looks like
one; check `class()`.
:::

# Date and time together

The `time` column is separate, and also text.

**Task.** Build a single `datetime` column combining `date` and `time`, then find
the hour of day with the most orders.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
pizza <-
  pizza %>%
  dplyr::mutate(datetime = lubridate::ymd_hms(paste(date, time)))

pizza$datetime[1]      # "2015-01-01 11:38:36 UTC"

pizza %>%
  dplyr::count(hour = lubridate::hour(datetime)) %>%
  dplyr::arrange(desc(n))
```

`paste()` glues the two strings, and `ymd_hms()` parses the result. The function
name is the format: year-month-day hour-minute-second. There is a matching
function for every ordering you will meet, so a date written `12/31/2015` needs
`mdy()` and one written `31-12-2015` needs `dmy()`.

That naming scheme is the whole reason `lubridate` is worth loading. In base R
the same job needs a format string like `"%Y-%m-%d %H:%M:%S"`, which you have to
look up every time.
:::

# Pull dates apart

**Task.**

1. Add columns for the month, the day of the week, and the day of the year.
2. Find the busiest day of the week.
3. Make the weekday names readable rather than numbers.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
pizza <-
  pizza %>%
  dplyr::mutate(
    month   = lubridate::month(date, label = TRUE),
    weekday = lubridate::wday(date, label = TRUE),
    doy     = lubridate::yday(date)
  )

pizza %>%
  dplyr::count(weekday) %>%
  dplyr::arrange(desc(n))
#> Fri 8242, Sat 7493, Thu 7478, ...
```

Friday is the busiest, which is the sort of finding that would be invisible while
the column was still text.

`label = TRUE` is what turns `6` into `Fri`. Without it you get an integer, and
you then have to remember whether the week starts on Sunday or Monday. With it
you get an **ordered factor**, which is what you want for plotting, because the
bars come out Sunday through Saturday rather than alphabetically from Fri to
Wed.
:::

# Aggregate by month without losing the date

**Task.** Compute total revenue by month. Do it twice:

1. using `month()`
2. using `floor_date()`

Then plot revenue over time using each, and say which one you would use for a
figure.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
#--- 1: month() gives you a label, not a date ---#
by_month_label <-
  pizza %>%
  dplyr::group_by(month = lubridate::month(date, label = TRUE)) %>%
  dplyr::summarize(revenue = sum(price))

#--- 2: floor_date() gives you a Date ---#
by_month_date <-
  pizza %>%
  dplyr::group_by(month = lubridate::floor_date(date, "month")) %>%
  dplyr::summarize(revenue = sum(price))

by_month_date
#> 2015-01-01  69793
#> 2015-02-01  65160
#> 2015-03-01  70397 ...

ggplot(by_month_date) +
  geom_line(aes(x = month, y = revenue))
```

Use `floor_date()` for anything you intend to plot. It rounds every date down to
the first of its month while **keeping it a Date**, so `ggplot2` puts it on a
proper time axis with correctly spaced months.

`month(label = TRUE)` gives you a factor. That is fine for a table and wrong for
a time-series figure, because a factor axis is evenly spaced categories rather
than time, and it falls apart the moment your data covers more than one year:
January 2015 and January 2016 collapse into a single "Jan".

`floor_date()` also takes `"week"`, `"quarter"`, and `"year"`, which is how you
change the frequency of a figure without touching anything else.
:::

# Date arithmetic

**Task.**

1. How many days between the first and last order?
2. How many *distinct* dates appear in the data? Compare with your answer to (1)
   and explain the gap.
3. For each order, compute how many days into the year it fell.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
span <- max(pizza$date) - min(pizza$date)
span                                  # Time difference of 364 days
as.numeric(span)                      # 364

dplyr::n_distinct(pizza$date)         # 358

pizza %>%
  dplyr::mutate(
    days_in = as.numeric(date - lubridate::ymd("2015-01-01"))
  ) %>%
  dplyr::select(date, days_in)
```

The gap is the point. The year spans 365 calendar days, 364 of them between the
first and last order, but only **358** distinct dates appear. The shop was shut
on seven days, and the only way to see that is to compare the span against the
count.

This matters whenever you assume a daily series is complete. Seven missing days
will not announce themselves: a bar chart just has slightly fewer bars, and a
mean is computed over whatever rows exist.

Note that subtracting two Dates gives a `difftime`, which prints its units. Wrap
it in `as.numeric()` when you want a plain number to do arithmetic with.
:::

::: {.callout-tip collapse="true" title="Finding the missing days"}
```{r}
#| eval: false
all_days <- seq(min(pizza$date), max(pizza$date), by = "day")
setdiff(all_days, unique(pizza$date)) %>% as.Date(origin = "1970-01-01")
```

Building the complete sequence and comparing against what you have is the
general technique for finding gaps in any time series.
:::

# Adding a month is not what you think

This is the exercise most worth doing slowly.

**Task.**

1. Take the date `2015-01-31` and add one month with `+ months(1)`.
2. Look at the result.
3. Now do the same for `2015-01-30`, `2015-01-29`, and `2015-01-28`.
4. Find the operator that does what you meant.

::: {.panel-tabset}

### Work here

```{webr-r}
library(lubridate)

ymd("2015-01-31") + months(1)
```

### Answer

```{r}
#| eval: false
#| code-fold: true
ymd("2015-01-31") + months(1)      # NA
ymd("2015-01-30") + months(1)      # NA
ymd("2015-01-29") + months(1)      # NA
ymd("2015-01-28") + months(1)      # "2015-02-28"

#--- what you meant ---#
ymd("2015-01-31") %m+% months(1)   # "2015-02-28"
ymd("2015-01-31") %m-% months(1)   # "2014-12-31"
```

Adding one month to 31 January asks for 31 February, which does not exist, so
`lubridate` returns `NA`. It is being honest: there is no correct answer, and
guessing one would be worse.

`%m+%` and `%m-%` say "roll back to the last valid day of the target month"
instead, giving 28 February.

Why this deserves an exercise rather than a footnote: imagine building a monthly
panel by repeatedly adding a month to a start date. Every month whose start is
the 29th, 30th, or 31st becomes `NA`. Your panel quietly loses about a third of
its rows, and `NA` propagates silently through every mean you compute afterwards.

Use `%m+%` for anything involving months or years. Plain `+` is fine for days
and weeks, which have no such ambiguity.
:::

# Rounding and leap years

**Task.**

1. Round the timestamps to the nearest hour, and separately round them down.
   Compare.
2. Which of 2000, 2001, 2004, and 2100 are leap years? Predict before running.
3. Explain the two that surprise people.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
round_date(pizza$datetime[1], "hour")   # 11:38:36 -> 12:00:00
floor_date(pizza$datetime[1], "hour")   # 11:38:36 -> 11:00:00

leap_year(c(2000, 2001, 2004, 2100))
#> TRUE FALSE TRUE FALSE
```

2000 is a leap year and 2100 is not, which catches nearly everybody. The rule is
divisible by 4, **except** centuries, **except** centuries divisible by 400. So
2000 qualifies through the exception to the exception, and 2100 does not.

Use `floor_date()` for binning, because it is what puts every observation in the
interval it actually falls in. `round_date()` moves half your 11:38s into the
noon bucket, which is wrong if you are counting orders per hour.
:::

# Put it together

**Task.** Produce a figure showing average daily revenue by weekday, split by
pizza size, starting from the raw `pizzaplace` data.

You will need to: parse the date, aggregate to daily revenue by size, then
average those daily totals within each weekday. Think about the order of those
last two steps, because doing them the other way round answers a different
question.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
daily <-
  pizzaplace %>%
  dplyr::mutate(date = lubridate::ymd(date)) %>%
  dplyr::group_by(date, size) %>%
  dplyr::summarize(revenue = sum(price), .groups = "drop")

by_weekday <-
  daily %>%
  dplyr::group_by(weekday = lubridate::wday(date, label = TRUE), size) %>%
  dplyr::summarize(mean_revenue = mean(revenue), .groups = "drop")

ggplot(by_weekday) +
  geom_col(aes(x = weekday, y = mean_revenue, fill = size),
           position = "dodge") +
  labs(x = NULL, y = "Mean daily revenue ($)")
```

The order matters. Aggregating to daily totals **first** and then averaging gives
you the average revenue on a Friday. Averaging prices directly by weekday gives
you the average price of a pizza sold on a Friday, which is a different quantity
and a much less interesting one.

Getting this backwards is a genuine analysis error rather than a coding one, and
no error message will appear. When you group twice, say out loud what one row of
each intermediate result represents.

Note also that `weekday` comes out correctly ordered Sunday to Saturday, because
`label = TRUE` produced an ordered factor. That is the payoff from exercise 3.
:::
