---
title: "Ex-7-2: Manipulating Strings"
abstract: "Date and String"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
webr:
  packages: ['dplyr', 'ggplot2', 'stringr', 'gt']
  cell-options:
    editor-font-scale: 0.7
    out-width: 100%
filters:
  - webr
---

These run in your browser. We use `pizzaplace` again, because its text columns
have exactly the properties that make string work necessary: names written with
underscores, an order id with structure inside it, and category labels that are
almost but not quite presentable.

Every function here is from `stringr`, and they all follow the same shape:
`str_something(the_strings, the_pattern)`. The strings come first, which means
they pipe.

# Data

```{webr-r}
#| autorun: true
data(pizzaplace, package = "gt")

pizza <- pizzaplace

#--- what the text columns look like ---#
unique(pizza$name)
unique(pizza$type)
head(pizza$id)
```

# Find rows by what the text contains

**Task.**

1. How many orders were for a chicken pizza? Chicken pizzas have `ckn` somewhere
   in the `name`.
2. How many distinct chicken pizzas are there, and what are they called?
3. Now find every order that is either a veggie or a supreme `type`, using a
   single pattern rather than two comparisons.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
library(stringr)

sum(str_detect(pizza$name, "ckn"))                    # 11050
dplyr::n_distinct(pizza$name[str_detect(pizza$name, "ckn")])   # 6
unique(pizza$name[str_detect(pizza$name, "ckn")])
#> "thai_ckn" "bbq_ckn" "southw_ckn" "cali_ckn" "ckn_pesto" "ckn_alfredo"

#--- inside filter, which is how you will actually use it ---#
chicken <- dplyr::filter(pizza, str_detect(name, "ckn"))

#--- either/or in one pattern ---#
sum(str_detect(pizza$type, "veggie|supreme"))         # 23636
```

`str_detect()` returns `TRUE`/`FALSE` for each element, which is exactly what
`filter()` wants.

Two things worth noticing. `ckn` appears at the start of some names and the end
of others, and `str_detect()` does not care: it looks anywhere in the string.
That is usually what you want and occasionally not, which is what anchors are
for in exercise 6.

And `|` means "or" inside a pattern. `str_detect(type, "veggie|supreme")` is
shorter and less error-prone than `type == "veggie" | type == "supreme"`,
especially once there are five alternatives.
:::

# Make text presentable

`classic_dlx` is a database name, not something you put in a figure legend.

**Task.** Turn `name` into something readable: underscores replaced with spaces,
and each word capitalized. Then plot the ten most-ordered pizzas with the
readable names on the axis.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
pizza <-
  pizza %>%
  dplyr::mutate(
    name_nice = str_to_title(str_replace_all(name, "_", " "))
  )

unique(pizza$name_nice)[1:4]
#> "Hawaiian" "Classic Dlx" "Mexicana" "Thai Ckn"

top10 <-
  pizza %>%
  dplyr::count(name_nice, sort = TRUE) %>%
  head(10)

ggplot(top10) +
  geom_col(aes(x = n, y = reorder(name_nice, n))) +
  labs(x = "Orders", y = NULL)
```

Note `str_replace_all()` rather than `str_replace()`. The plain version replaces
only the **first** match, so `pep_msh_pep` would come back as `pep msh_pep`. One
of the pizza names has two underscores, which is the only reason you would catch
this before it reached a figure.

Do the cleaning in a `mutate()` into a new column rather than overwriting `name`.
Keeping the machine-readable version means you can still join, filter, and match
on it; the pretty version is for display only.

The `reorder(name_nice, n)` in the plot is what sorts the bars by count instead
of alphabetically.
:::

# Take strings apart

The `id` column looks like `2015-000001`: a year, then a sequence number.

**Task.**

1. Split `id` into a year column and an order-number column.
2. Confirm the year always matches the year in `date`.
3. How many distinct order ids are there, and why is that fewer than the number
   of rows?

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
pizza <-
  pizza %>%
  dplyr::mutate(
    id_year = str_split_fixed(id, "-", 2)[, 1],
    id_num  = str_split_fixed(id, "-", 2)[, 2]
  )

head(pizza$id_year)   # "2015" "2015" ...
head(pizza$id_num)    # "000001" "000002" ...

all(pizza$id_year == substr(pizza$date, 1, 4))   # TRUE

dplyr::n_distinct(pizza$id)   # 21350
nrow(pizza)                   # 49574
```

There are 21,350 orders but 49,574 rows, because **one row is one pizza, not one
order**. An order containing three pizzas appears as three rows sharing an id.

That is the single most important thing to know about this dataset, and the id
column is what tells you. If you had computed "average order value" as
`mean(price)` you would have got the average *pizza* price and been wrong by a
factor of more than two.

On the mechanics: `str_split_fixed(x, "-", 2)` returns a **matrix** with two
columns, which is why the `[, 1]` and `[, 2]` are needed. Use it rather than
`str_split()` when you know how many pieces you want; `str_split()` returns a
list, which is more awkward inside `mutate()`.
:::

::: {.callout-tip collapse="true" title="The tidier way to split a column"}
```{r}
#| eval: false
pizza %>% tidyr::separate_wider_delim(id, "-", names = c("id_year", "id_num"))
```

One call, two columns, no matrix indexing. Worth knowing once you have done it
by hand.
:::

# Rebuild an id with padding

**Task.** You are given order numbers as plain integers: `c(7, 42, 1350)`.
Rebuild ids in the original format, `2015-000007` and so on.

::: {.panel-tabset}

### Work here

```{webr-r}
nums <- c(7, 42, 1350)
```

### Answer

```{r}
#| eval: false
#| code-fold: true
nums <- c(7, 42, 1350)

str_c("2015-", str_pad(nums, width = 6, side = "left", pad = "0"))
#> "2015-000007" "2015-000042" "2015-001350"
```

`str_pad()` is what keeps ids the same length. Fixed-width ids sort correctly as
text, which unpadded ones do not: alphabetically, `"1350"` comes before `"42"`.

This is the same problem as the leading zeros in FIPS codes from
[Ex-3-1](3-1-input-output.qmd), approached from the other direction. There you
were stopping R from destroying the padding on the way in; here you are creating
it on the way out. Both exist because an identifier is text that happens to look
like a number.

`str_c()` is `paste0()` with one useful difference: it propagates `NA` instead of
turning it into the literal string `"NA"`. `paste0("x", NA)` gives `"xNA"`, which
is almost never what you want and is very hard to spot later.
:::

# Build labels for a figure

**Task.** Create a single label column combining the readable name, the size, and
the price, formatted like `Classic Dlx (M) - $16.00`. Then show the five most
expensive distinct pizza-size combinations using those labels.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
labelled <-
  pizza %>%
  dplyr::distinct(name_nice, size, price) %>%
  dplyr::mutate(
    label = str_c(name_nice, " (", size, ") - $", format(price, nsmall = 2))
  ) %>%
  dplyr::arrange(desc(price))

head(labelled$label, 5)
```

`str_c()` takes as many pieces as you like and is vectorized, so this builds
50,000 labels in one call rather than in a loop.

The `format(price, nsmall = 2)` is there because `str_c()` converts numbers using
R's default rules, which drop trailing zeros: a price of 16 becomes `"16"`, not
`"16.00"`, and your labels come out inconsistent. Anywhere you paste a number
into text for people to read, decide the number of decimal places explicitly.
:::

# Anchors and counting

**Task.**

1. Find the pizza names that **start** with `ckn`, as opposed to containing it
   anywhere.
2. Find names containing two or more underscores.
3. What are the shortest and longest names?

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
nm <- unique(pizza$name)

nm[str_detect(nm, "^ckn")]          # "ckn_pesto" "ckn_alfredo"
nm[str_detect(nm, "ckn$")]          # "thai_ckn" "bbq_ckn" "southw_ckn" "cali_ckn"

nm[str_count(nm, "_") >= 2]         # "pep_msh_pep"

range(str_length(nm))               # 7 12
nm[str_length(nm) == max(str_length(nm))]
```

`^` anchors to the start of the string and `$` to the end. Compare with exercise
1, where the unanchored `"ckn"` matched all six names: anchoring is how you turn
"contains" into "begins with".

`str_count()` counts matches rather than reporting whether there are any, which
is how you find the awkward case that breaks a naive `str_replace()`.

Those two characters are your entire introduction to regular expressions, and
between them and `|` you can express most of what you actually need day to day.
:::

# Clean a column nobody validated

**Task.** Real category columns arrive like this. Clean it so that
`dplyr::count()` gives three categories rather than nine.

```{webr-r}
messy <- c("Corn", "corn ", " CORN", "corn", "Soybean", "soybean",
           "SOY BEAN", "Wheat ", "wheat")
```

::: {.panel-tabset}

### Work here

```{webr-r}
messy <- c("Corn", "corn ", " CORN", "corn", "Soybean", "soybean",
           "SOY BEAN", "Wheat ", "wheat")
```

### Answer

```{r}
#| eval: false
#| code-fold: true
cleaned <-
  messy %>%
  str_trim() %>%                       # drop leading/trailing spaces
  str_to_lower() %>%                   # one case
  str_replace_all(" ", "")             # "soy bean" -> "soybean"

table(cleaned)
#> corn 4, soybean 3, wheat 2
```

Three operations, in this order, fix most category columns you will ever meet:
trim the whitespace, force one case, then deal with internal spacing.

Order matters a little. Trimming before lowercasing is not important, but doing
both before you compare is essential, and doing the internal-space replacement
last keeps it from interfering with the trim.

The reason this exercise exists: `" CORN"`, `"corn "`, and `"Corn"` are three
different strings, so `group_by()` gives you three groups, and your summary table
splits one crop across three rows. Nothing errors. You just get an answer that
is quietly wrong, and it is easy to miss in a table with forty rows.

Whenever you are about to group by a text column, run `unique()` on it first and
look at the result.
:::

# Put it together

**Task.** Produce a table of revenue by pizza category, where:

- the categories come from `type`, with readable labels (`Classic`, not
  `classic`)
- chicken pizzas are pulled out as their own category regardless of `type`
- the ingredient count for each pizza name is estimated as the number of
  underscore-separated pieces
- the output is sorted by revenue

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
result <-
  pizza %>%
  dplyr::mutate(
    category = dplyr::case_when(
      str_detect(name, "ckn") ~ "Chicken",
      TRUE                    ~ str_to_title(type)
    ),
    n_pieces = str_count(name, "_") + 1
  ) %>%
  dplyr::group_by(category) %>%
  dplyr::summarize(
    revenue    = sum(price),
    n_pizzas   = dplyr::n(),
    mean_pieces = mean(n_pieces)
  ) %>%
  dplyr::arrange(desc(revenue))

result
```

Two things worth noting.

`case_when()` evaluates its conditions **in order** and stops at the first match,
which is what makes "chicken first, then whatever `type` says" work. Reverse the
two lines and every chicken pizza gets its `type` label instead, because that
condition now matches first.

`str_count(name, "_") + 1` counts separators and adds one to get pieces, which is
the standard fencepost adjustment. `hawaiian` has no underscore and one piece;
`pep_msh_pep` has two and three.
:::
