---
title: "Ex-8-1: Make tables with `flextable`"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
webr:
  packages: ['dplyr', 'ggplot2', 'datasets', 'flextable']
  cell-options:
    editor-font-scale: 0.7
    out-width: 100%
filters:
  - webr
---

```{r}
#| include: false 
library(dplyr)
library(ggplot2)
library(flextable)
```

## Exercise 1: Create a Simple Flextable

**Dataset:** `mtcars`

1.  Load the `mtcars` dataset into R.
2.  Use the `flextable` package to create a table that displays the first 10 rows of the dataset for `mpg`, `cyl`, `disp`, and `hp`.
3.  Apply styles to the table
    1. make the texts in the header bold using `bold()`
    2. make the texts of the `cyl` column center-aligned using ``
    3. make the texts of the `hp` column colored red using ``

Here is the finished table you are trying to make:

```{r}
#| echo: false
(
ft <- 
  flextable(
    mtcars[1:10, ],
    col_keys = c("mpg", "cyl", "disp", "hp")
  ) %>%
  # Apply simple formatting
  theme_box() %>%
  bold(part = "header")
) 
```



::: {.panel-tabset}
### Work here
```{webr-r}
#| results: asis
data(mtcars)


```

### Answer
```{r, eval = FALSE}
#| code-fold: true
# Create a simple flextable
ft <- 
  flextable(mtcars) %>%
  # Apply simple formatting
  theme_box() %>%
  bold(part = "header")
```

:::


## Exercise 2: Conditional Formatting

**Dataset:** `iris`

1.  Load the `iris` dataset.
2.  Create a `flextable` that displays the first 10 rows of the dataset.
3.  Apply conditional formatting: Highlight cells in the `Sepal.Length` column where the values are greater than 6.0 in a different color.

Here is the finished table you are trying to make:

```{r}
#| echo: false
(
ft <- 
  flextable(iris[1:10,]) %>%
  # Apply conditional formatting to Sepal.Length column
  bg(i = ~ Sepal.Length > 6, j = "Sepal.Length", bg = "#FFDDC1")
)
```


::: {.panel-tabset}
### Work here
```{webr-r}
# Load iris dataset
data(iris)
```

### Answer
```{r, eval = FALSE}
#| code-fold: true
# Create flextable for first 10 rows
ft <- 
  flextable(iris[1:10,]) %>%
  # Apply conditional formatting to Sepal.Length column
  bg(i = ~ Sepal.Length > 6, j = "Sepal.Length", bg = "#FFDDC1")

# Display the table
ft
```

:::


## Exercise 3: Merging and Grouping Columns

**Dataset:** `airquality`

1.  Load the `airquality` dataset.
2.  Create a flextable using selected columns: `Month`, `Day`, `Ozone`, `Wind`.
3.  Group the table by the `Month` column and merge the cells of the `Month` column.
4.  Add a caption to the table.

Here is the finished table you are trying to make:

```{r}
#| echo: false
(
ft <- 
  airquality[, c("Month", "Day", "Ozone", "Wind")] %>%
  # Create flextable
  flextable() %>%
  # Merge Month column
  merge_v(j = "Month") %>%
  # Add a caption
  set_caption("Airquality Data Table with Grouped Month")
) 
```

::: {.panel-tabset}
### Work here
```{webr-r}
# Load airquality dataset
data(airquality)
```

### Answer
```{r, eval = FALSE}
#| code-fold: true
# Select columns
ft <- 
  airquality[, c("Month", "Day", "Ozone", "Wind")] %>%
  # Create flextable
  flextable() %>%
  # Merge Month column
  merge_v(j = "Month") %>%
  # Add a caption
  set_caption("Airquality Data Table with Grouped Month")
```

:::

## Exercise 4: Custom Header and Footer

**Dataset:** `iris`

1.  Load the `iris` dataset.
2.  Create a flextable with the first 15 rows of the dataset.
3.  Modify the header by renaming columns `Sepal.Length` and `Sepal.Width` to `Sepal Length` and `Sepal Width`, respectively.
4.  Add a custom footer explaining what the dataset represents.

Here is the finished table you are trying to make:

```{r}
#| echo: false
(
ft <- 
  flextable(iris[1:15,]) %>%
  # Modify header labels
  set_header_labels(Sepal.Length = "Sepal Length", Sepal.Width = "Sepal Width") %>%
  # Add a footer
  add_footer_lines("Iris dataset: Measurements of iris flowers from three species") 
)
```

::: {.panel-tabset}
### Work here
```{webr-r}
# Load iris dataset
data(iris)
```

### Answer
```{r, eval = FALSE}
#| code-fold: true
ft <- 
  flextable(iris[1:15,]) %>%
  # Modify header labels
  set_header_labels(Sepal.Length = "Sepal Length", Sepal.Width = "Sepal Width") %>%
  # Add a footer
  add_footer_lines("Iris dataset: Measurements of iris flowers from three species")
```

:::
