---
title: "Ex-8-2: Regression and Summary Tables with `modelsummary`"
abstract: "Tables"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

These run in RStudio rather than in your browser, because the point of the
chapter is producing tables that end up in a paper, and that means writing Word
files and images to disk.

The deliverable is not a table printed in your console. It is a table that lands
in a document, numbered, captioned, and cross-referenced. Several exercises end
by putting the table into a qmd, which is where the work from
[Ex-2-3](2-3-quarto-article.qmd) comes back.

# Setup {#sec-setup}

```{r}
#| eval: false
install.packages(
  "r.spatial.workshop.datasets",
  repos = c("https://tmieno2.r-universe.dev", "https://cran.r-project.org")
)

install.packages(c("modelsummary", "fixest", "flextable", "gt", "sandwich"))
```

Work in a project with the folder structure from
[Ex-6-1](6-1-project-organization.qmd) if you have one. Tables belong in
`Results/`.

```{r}
#| eval: false
library(tidyverse)
library(modelsummary)
library(fixest)
library(here)

data(county_yield, package = "r.spatial.workshop.datasets")

#--- the data is spatial; we do not need the polygons here ---#
cy <- dplyr::select(county_yield, -geometry)

cy
```

The data is county-level corn and soybean yields for Kansas, Nebraska, and
Colorado, 2000 to 2018, with drought indicators.

+ `corn_yield`, `soy_yield`: yield in bu/acre
+ `d0_5_9` through `d4_5_9`: share of weeks May-September at each drought
  severity, 0 mildest through 4 most severe

::: {.callout-important title="It worked if"}
`dim(cy)` is 1956 by 10, and `unique(cy$state_name)` gives three states.
:::

# Three models, one table

**Task.**

1. Estimate corn yield on `d1_5_9` and `d2_5_9` three ways: pooled, with county
   fixed effects, and with county and year fixed effects.
2. Put all three in one table.
3. Read across the row for `d1_5_9` and say what happens to it.

::: {.callout-important title="It worked if"}
Your table has three columns, all with 1,956 observations, and R² rising from
about 0.05 to about 0.54.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
m1 <- feols(corn_yield ~ d1_5_9 + d2_5_9, data = cy)
m2 <- feols(corn_yield ~ d1_5_9 + d2_5_9 | county_code, data = cy)
m3 <- feols(corn_yield ~ d1_5_9 + d2_5_9 | county_code + year, data = cy)

modelsummary(list(m1, m2, m3))
```

In `fixest`, everything after the `|` is a fixed effect, absorbed rather than
reported. That is why the intercept disappears in columns 2 and 3.

Reading across `d1_5_9`: about -0.22 pooled, -0.25 with county effects, then
**+0.04** once year effects are added. The coefficient changes sign.

That is what the table is for. Drought years are bad years everywhere, so
without year effects the coefficient is partly picking up "this was a bad year
nationally" rather than "this county was in drought". Adding year fixed effects
removes the common shock, and the mild-drought effect essentially vanishes.
`d2_5_9`, the more severe category, shrinks but stays negative.

A regression table is not a formality. Putting the specifications side by side
is what makes a result like this visible.
:::

# Make it fit for a paper

The default table is for you. This one is for a referee.

**Task.** Take the three models and:

1. rename the coefficients to something a reader understands
2. show only the number of observations and R², not the eight default statistics
3. add significance stars
4. give the columns meaningful headings instead of (1), (2), (3)
5. add a title and a note explaining the fixed effects

::: {.callout-important title="It worked if"}
Nothing in the table is named after a variable in your data frame.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
modelsummary(
  list(
    "Pooled"        = m1,
    "County FE"     = m2,
    "County+Year FE" = m3
  ),
  coef_map = c(
    "d1_5_9"      = "Moderate drought (share of weeks)",
    "d2_5_9"      = "Severe drought (share of weeks)",
    "(Intercept)" = "Constant"
  ),
  gof_map = c("nobs", "r.squared"),
  stars   = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  title   = "Drought exposure and county corn yields, 2000-2018.",
  notes   = "Standard errors in parentheses. Drought shares measured May-September."
)
```

`coef_map` does two jobs at once, and the second is easy to miss: it renames the
coefficients **and** sets their order, and it **drops anything not listed**. That
is the cleanest way to suppress a control you estimated but do not want in the
table.

`gof_map` does the same for the goodness-of-fit rows. The default prints
everything the model object offers, most of which nobody wants.

Naming the list elements is what names the columns.
:::

::: {.callout-tip collapse="true" title="Finding the names to use"}
```{r}
#| eval: false
get_gof(m1)          # every available gof statistic and its name
get_estimates(m1)    # every available coefficient and its name
```

Use these rather than guessing at what `gof_map` will accept.
:::

# Change the standard errors, not the model

**Task.** Take `m3` and present it four times in one table, with: default
standard errors, heteroskedasticity-robust, clustered by county, and clustered
by county and year. Do it **without re-estimating the model**.

Then say which you would report and why.

::: {.callout-important title="It worked if"}
The coefficients are identical across all four columns and only the parentheses
change.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
modelsummary(
  list(m3, m3, m3, m3),
  vcov = list(
    "IID"            = "iid",
    "Robust"         = "hetero",
    "Cluster: county" = ~county_code,
    "Two-way"        = ~county_code + year
  ),
  coef_map = c("d1_5_9" = "Moderate drought", "d2_5_9" = "Severe drought"),
  gof_map  = c("nobs", "r.squared")
)
```

The `vcov` argument recomputes the variance-covariance matrix after the fact.
The point estimates cannot change, because the estimator has not changed, only
the uncertainty around it.

This is the right way to present a robustness check on inference, and it is
much better than estimating the same model four times: it makes it obvious that
the specification is fixed and only the standard errors move.

Which to report: with repeated observations on the same counties over time,
errors are almost certainly correlated within county, so clustering by county is
the honest default. Standard errors typically get **larger**, and a result that
survives clustering is worth more than one that does not.
:::

# Get it out of R

**Task.** Save your paper-ready table as:

1. a Word document
2. a png image
3. an object you can edit further with `flextable`

Put all of them in `Results/`.

::: {.callout-important title="It worked if"}
You can open the Word file and edit the table as a table, not as a picture.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
tbl <- modelsummary(
  list("Pooled" = m1, "County FE" = m2, "County+Year FE" = m3),
  coef_map = c("d1_5_9" = "Moderate drought", "d2_5_9" = "Severe drought"),
  gof_map  = c("nobs", "r.squared"),
  output   = here("Results", "tbl-drought-yield.docx")
)

modelsummary(..., output = here("Results", "tbl-drought-yield.png"))

#--- keep it as an object to modify ---#
ft <- modelsummary(list(m1, m2, m3), output = "flextable")

ft %>%
  flextable::bold(i = 1, part = "header") %>%
  flextable::autofit()
```

`output =` decides everything. A file path writes a file; `"flextable"` or
`"gt"` returns an object you can keep styling; `"markdown"` gives you plain text
useful for pasting into an email.

Word output is worth the trouble because coauthors and journals ask for it, and
because a real Word table can be edited by someone who does not use R. A png of
a table cannot, and it also looks wrong next to the document's own fonts.

Use png only when the destination cannot accept anything better, which in
practice means slides.
:::

# Summary statistics without writing any code twice

**Task.**

1. Produce a quick summary of every numeric variable.
2. Then produce a controlled version: yields and the two drought measures only,
   with mean, standard deviation, minimum, and maximum, with readable row names.

::: {.callout-important title="It worked if"}
The second table has exactly four rows and four statistic columns.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
#--- 1: the quick look ---#
datasummary_skim(cy)

#--- 2: the controlled version ---#
datasummary(
  (`Corn yield (bu/ac)` = corn_yield) +
  (`Soy yield (bu/ac)`  = soy_yield) +
  (`Moderate drought`   = d1_5_9) +
  (`Severe drought`     = d2_5_9) ~
  Mean + SD + Min + Max,
  data = cy,
  title = "Summary statistics, 2000-2018."
)
```

`datasummary()` uses a two-sided formula: **rows on the left, columns on the
right**, joined by `~`. Once you see that, the syntax stops being mysterious.

Renaming happens inline with `(\`Nice name\` = variable)`, which is convenient
but does mean the backticks matter. A missing one produces a confusing error
about an object not being found.

`datasummary_skim()` is for you, during analysis. `datasummary()` is for the
paper. Do not put a skim table in a manuscript: it reports things nobody asked
for and omits the units.
:::

# Statistics by group

**Task.** Extend the summary table so the statistics are shown separately for
each state, with an "All" column alongside. Then add a count of observations per
state.

::: {.callout-important title="It worked if"}
There are four column groups: three states and All.
:::

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
datasummary(
  (`Corn yield` = corn_yield) +
  (`Soy yield`  = soy_yield) +
  (`Severe drought` = d2_5_9) ~
  state_name * (Mean + SD) + (`All` = 1) * (Mean + SD),
  data = cy,
  title = "Summary statistics by state."
)
```

The `*` is nesting: `state_name * (Mean + SD)` means "for each state, show mean
and SD". Reading it aloud as "crossed with" makes the syntax behave.

`(\`All\` = 1)` is the idiom for a total column. The `1` is a grouping variable
that takes the same value for every row, so it produces one group containing
everything.

This is exactly the table most empirical papers open with, and building it by
hand with `group_by()` and `pivot_wider()` is an afternoon's work that has to be
redone every time a variable changes.
:::

# Two convenience tables

**Task.**

1. Produce a balance table comparing all variables across states.
2. Produce a correlation matrix of the yield and drought variables.
3. Say what you would check in each before putting it in a paper.

::: {.callout-note collapse="true" title="Answer"}
```{r}
#| eval: false
cy %>%
  dplyr::select(corn_yield, soy_yield, d1_5_9, d2_5_9, state_name) %>%
  datasummary_balance(~state_name, data = .)

cy %>%
  dplyr::select(corn_yield, soy_yield, d0_5_9, d1_5_9, d2_5_9, d3_5_9) %>%
  datasummary_correlation()
```

What to check in the balance table: whether the groups differ on things they
should not, and whether the counts per group are wildly unequal. It is designed
for treatment-versus-control comparisons, so read the differences as
descriptive here, not causal.

What to check in the correlation matrix: the drought categories are shares of
the same weeks, so they are mechanically related — more weeks at severity 2
means fewer at severity 0. Strong negative correlations between adjacent
categories are structural, not a finding, and putting all five in one regression
is a collinearity problem rather than a specification.

A correlation matrix is most useful for catching this **before** you estimate,
which is why it is worth producing early even if it never reaches the paper.
:::

# Put it in a document

**Task.** Write a short qmd in `Writing/` that:

- reads the data and estimates the three models in a hidden chunk
- prints the paper-ready regression table with a caption and a `tbl-` label
- prints the summary statistics table, also labelled
- refers to both from the prose by cross-reference
- renders to both html and Word

::: {.callout-important title="It worked if"}
The rendered document says "Table 1" and "Table 2" in the prose, and no
`?tbl-` appears anywhere.
:::

::: {.callout-note collapse="true" title="Answer"}
````{verbatim}
```{r}
#| label: tbl-drought
#| tbl-cap: "Drought exposure and county corn yields, 2000-2018."
#| echo: false
modelsummary(
  list("Pooled" = m1, "County FE" = m2, "County+Year FE" = m3),
  coef_map = c("d1_5_9" = "Moderate drought", "d2_5_9" = "Severe drought"),
  gof_map  = c("nobs", "r.squared"),
  stars    = TRUE,
  output   = "flextable"
)
```

Severe drought reduces yields substantially, though the effect
falls once year fixed effects absorb common shocks (@tbl-drought).
Summary statistics are in @tbl-summary.
````

Two practical points.

Set `output = "flextable"` or `"gt"` when the table is going into a Quarto
document. Leave `output` at its default and `modelsummary` picks a format based
on the destination, which mostly works and occasionally produces something that
looks wrong in Word.

Do not pass a **file path** as `output` in a chunk you also want to display. That
writes the file and returns nothing, so your document gets a caption with an
empty space under it and no error to explain why.
:::
