---
title: "Ex-9-1: Creating Maps with `ggplot2`"
abstract: "Spatial Data"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
webr:
  packages: ['dplyr', 'ggplot2', 'sf', 'viridis', 'ggspatial', 'ggrepel']
  cell-options:
    editor-font-scale: 0.7
    out-width: 90%
filters:
  - webr
---

These run in your browser. The first cell installs the data package, which takes
a little while on first load; wait for it to finish before running anything
else.

A map in `ggplot2` is an ordinary figure whose geometry happens to be spatial.
Everything you know from Chapter 4 — layers, `aes()`, scales, facets, themes —
works unchanged. The only genuinely new ideas are `geom_sf()`, which knows how
to draw geometry, and coordinate reference systems, which are where the real
difficulty lives.

# Data

```{webr-r}
#| context: setup
webr::install(
  "r.spatial.workshop.datasets",
  repos = c(
    "https://tmieno2.github.io/rwasm-package/repo",
    "https://repo.r-wasm.org"
  )
)

library("r.spatial.workshop.datasets")

data(wells_ne)
data(ne_counties)
data(railroads_ne)
data(corn_acres_ne)
```

```{webr-r}
#| autorun: true
library(sf)
library(dplyr)
library(ggplot2)

ne_counties      # 93 Nebraska counties (polygons)
corn_acres_ne    # corn acreage by county and year (polygons)
head(wells_ne)   # 1000 irrigation wells
```

# Not everything with coordinates is spatial

**Task.**

1. Check the class of `wells_ne` and of `ne_counties`. They are not the same.
2. Try to plot `wells_ne` with `geom_sf()`. Read the error.
3. Convert `wells_ne` into an `sf` object. The coordinates are in `longdd` and
   `latdd`, and the CRS is EPSG 4269, the same as the counties.
4. Confirm the conversion worked.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
class(wells_ne)     # "data.frame"   <- not spatial
class(ne_counties)  # "sf" "data.frame"

ggplot(wells_ne) + geom_sf()
#> Error: stat_sf requires the following missing aesthetics: geometry

wells_sf <-
  st_as_sf(
    wells_ne,
    coords = c("longdd", "latdd"),   # longitude FIRST
    crs    = 4269
  )

class(wells_sf)        # "sf" "data.frame"
st_crs(wells_sf)       # NAD83
ggplot(wells_sf) + geom_sf()
```

A data frame with a longitude column and a latitude column is not spatial. It is
a table with two numbers in it, and R has no idea those numbers mean anything.
`st_as_sf()` is what promotes it: it folds the two columns into a single
`geometry` column and attaches a coordinate reference system.

Two things that catch people:

- **longitude comes first** in `coords`. Longitude is the x axis. Everyone says
  "lat/long" out loud and then writes it in the wrong order, and the result is a
  map of Nebraska somewhere off the coast of Somalia.
- **the CRS is not optional.** Without `crs =` the numbers have no meaning, and
  `sf` cannot combine your points with anything else later.
:::

# Your first map

**Task.** Draw the counties of Nebraska. Then draw the wells on top of them.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
#--- counties alone ---#
ggplot(ne_counties) +
  geom_sf()

#--- wells on top ---#
ggplot() +
  geom_sf(data = ne_counties, fill = "white", color = "grey60") +
  geom_sf(data = wells_sf, color = "blue", size = 0.5, alpha = 0.4)
```

`geom_sf()` works out for itself whether it is drawing points, lines, or
polygons, so the same function draws both layers.

Note where `data =` goes. With two different datasets you cannot put one in
`ggplot()` and expect the other layer to find it, so give each `geom_sf()` its
own `data` argument and leave `ggplot()` empty. This is the global-versus-local
data distinction from Chapter 4, and maps are where you need it constantly,
because a map almost always combines several sources.

`alpha` matters with a thousand overlapping points. At full opacity you see a
blue smear; at 0.4 you see where the wells actually concentrate.
:::

# A choropleth

**Task.** Map corn acreage by county for 2023 alone, with counties shaded by
acreage. Use a viridis scale, and give the legend a readable title.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
corn_2023 <- dplyr::filter(corn_acres_ne, year == 2023)

ggplot(corn_2023) +
  geom_sf(aes(fill = acre), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "magma", labels = scales::comma) +
  labs(fill = "Corn acres") +
  theme_void()
```

The only change from an ordinary bar chart is that the thing being filled is a
polygon. `aes(fill = acre)` inside `geom_sf()` is doing exactly what
`aes(fill = ...)` does anywhere else.

`scale_fill_viridis_c` with a `_c` because acreage is continuous. The `_d`
version is for discrete categories and silently does nothing useful here.

`theme_void()` removes the panel, grid, and axes. Latitude and longitude
gridlines are rarely informative on a small-area map and mostly add clutter; the
reader knows where Nebraska is.

Notice that not every county appears — there are 69 counties in 2023, not 93.
Counties with no reported acreage are simply absent, and they render as holes.
Whether that is acceptable or misleading depends on whether "not reported" means
"no corn", and it is worth stating in the caption either way.
:::

# Facet by year

**Task.** Show all four years at once as a two-by-two grid of maps, sharing one
colour scale.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
ggplot(corn_acres_ne) +
  geom_sf(aes(fill = acre), color = NA) +
  facet_wrap(~year, nrow = 2) +
  scale_fill_viridis_c(option = "magma", labels = scales::comma) +
  labs(fill = "Corn acres") +
  theme_void() +
  theme(legend.position = "bottom")
```

`facet_wrap()` works on maps exactly as it does on scatter plots. There is no
spatial version of faceting to learn.

The shared colour scale is the entire reason this is better than four separate
maps: with one scale, a county that is darker in 2023 than in 2020 really did
grow more corn. Four separately-scaled maps would each be normalised to their
own range, and comparing across them would be meaningless while looking
perfectly reasonable.

That is the most common way a multi-panel map misleads, and it is worth checking
whenever you see one.
:::

# Make it presentable

**Task.** Take your 2023 choropleth and add:

1. the railroads as a line layer
2. a scale bar and a north arrow
3. a title and a source note
4. county names for the five counties with the most corn acreage, without the
   labels overlapping

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
library(ggspatial)
library(ggrepel)

top5 <-
  corn_2023 %>%
  dplyr::arrange(desc(acre)) %>%
  head(5) %>%
  dplyr::left_join(sf::st_drop_geometry(ne_counties),
                   by = c("county_code" = "countyfp"))

ggplot() +
  geom_sf(data = corn_2023, aes(fill = acre), color = "white", linewidth = 0.1) +
  geom_sf(data = railroads_ne, color = "grey30", linewidth = 0.2, alpha = 0.5) +
  geom_sf_text(data = top5, aes(label = name), size = 3, color = "white") +
  scale_fill_viridis_c(option = "magma", labels = scales::comma) +
  annotation_scale(location = "bl") +
  annotation_north_arrow(location = "tr", height = unit(1, "cm"),
                         width = unit(1, "cm")) +
  labs(
    title   = "Corn acreage by county, Nebraska, 2023",
    caption = "Source: USDA NASS",
    fill    = "Corn acres"
  ) +
  theme_void()
```

Layer order is drawing order. Railroads added after the polygons sit on top of
them; put them first and the county fills paint over them entirely.

`annotation_scale()` and `annotation_north_arrow()` come from `ggspatial` and
know how to compute a scale bar correctly for the map's projection, which is why
you should not draw one by hand.

For labels, `geom_sf_text()` places text at each feature's centroid. When labels
collide, `ggrepel::geom_text_repel()` pushes them apart, though on an `sf` object
it needs `stat = "sf_coordinates"` and an `aes(geometry = geometry)`.

A north arrow is conventional rather than always necessary. On a map that is
obviously north-up of a place the reader recognises, it is decoration; on a
rotated or unfamiliar extent, it is essential.
:::

# Zoom without throwing data away

**Task.** Produce a map of the southeastern corner of Nebraska only, showing
counties and wells. Do it **without** filtering the data.

Then do it by filtering, and say why the first approach is usually better.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
ggplot() +
  geom_sf(data = ne_counties, fill = "white", color = "grey60") +
  geom_sf(data = wells_sf, color = "blue", size = 0.8, alpha = 0.5) +
  coord_sf(xlim = c(-98, -95.3), ylim = c(40, 41)) +
  theme_void()
```

`coord_sf()` changes the **window**, not the data. Counties that straddle the
edge are drawn and clipped, so the boundary looks the way a map should.

Filtering instead removes whole features, so a county whose centroid falls
outside the window vanishes completely, leaving a ragged edge and a hole where
part of a county should be. The map then implies there is nothing there, which
is a different claim from "outside the area shown".

Filter when you mean "these are the observations I am analysing". Use
`coord_sf()` when you mean "this is the part of the map I am showing". They look
similar and say different things.
:::

::: {.callout-tip collapse="true" title="Getting the limits"}
```{r}
#| eval: false
st_bbox(ne_counties)
```

Start from the full bounding box and narrow it, rather than guessing
coordinates.
:::

# Put it together

**Task.** Produce a single publication-quality figure answering: **where in
Nebraska is corn grown, and where are the irrigation wells?**

Requirements:

- corn acreage for the most recent year as the base layer
- wells on top, visible against the fill
- a scale bar
- a title, a legend title, and a source caption
- no axis clutter
- a colour scale that stays readable in greyscale

Then say in one sentence what the figure shows that a table could not.

::: {.panel-tabset}

### Work here

```{webr-r}

```

### Answer

```{r}
#| eval: false
#| code-fold: true
ggplot() +
  geom_sf(data = corn_2023, aes(fill = acre), color = "white", linewidth = 0.1) +
  geom_sf(data = wells_sf, color = "white", size = 0.35, alpha = 0.5) +
  scale_fill_viridis_c(option = "viridis", labels = scales::comma) +
  annotation_scale(location = "bl") +
  labs(
    title    = "Corn acreage and irrigation wells, Nebraska, 2023",
    subtitle = "Each point is one registered irrigation well",
    caption  = "Sources: USDA NASS; Nebraska DNR",
    fill     = "Corn acres"
  ) +
  theme_void() +
  theme(legend.position = "right")
```

On the greyscale requirement: viridis is designed so that its lightness
increases monotonically, which means it survives being printed in black and
white and is readable to colour-blind viewers. A red-to-green scale fails both
tests, and journals still print in greyscale more often than you would expect.

On the white points: the well colour has to be chosen against the fill, not in
isolation. Blue points on a dark viridis background disappear.

What the figure shows that a table cannot: the *spatial coincidence* of the two
variables. A table can tell you that wells and corn acres are both concentrated
in certain counties, but only a map shows you that they follow the river valleys
together, which is a claim about geography rather than about correlation.
:::
