---
title: "07-1: Date and String Management"
abstract: "Chapter 7: Date and String"
format: 
  revealjs:
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    include-in-header:
      - ../transcript-support.html
      - ../webr-layout.html
      - ../katex-amd-fix.html
    footer: "[back to the lecture slides on the course website](https://tmieno2.github.io/Data-Science-with-R-Quarto/lectures/)" 
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
webr:
  packages: ['dplyr', 'stringr', 'lubridate', 'gt']
  cell-options:
    editor-font-scale: 0.7
    dpi: 216
    fig-width: 10
    fig-height: 6
    out-width: 100%
filters:
  - webr
---

## Tips to make the most of the lecture notes

```{r}
#| include: false 
library(stringr) 
library(dplyr)
library(ggplot2) 
library(gt)

source(here::here("lectures/_lecture-theme.R"))
```

::: {.panel-tabset}

### Interactive navigation tools

<details class="transcript">
<summary>Transcript</summary>

The usual two navigation tricks. Three stacked lines in the bottom left corner give you a table of contents, and the letter o gives you a panel view of every slide. Worth knowing on this deck in particular, because it splits cleanly into two halves you will want to come back to separately. The first half is dates, the second is string manipulation, and in practice you tend to need one or the other rather than both at once. So when you are actually cleaning a dataset in a few weeks, being able to jump straight to the padding tab or the detect tab is the whole point.

</details>

+ Click on the three horizontally stacked lines at the bottom left corner of the slide, then you will see table of contents, and you can jump to the section you want

+ Hit letter "o" on your keyboard and you will have a panel view of all the slides

### Running and writing codes

<details class="transcript">
<summary>Transcript</summary>

The code boxes are live. Blue area is an editor, Run Code runs it, and you can highlight a piece and press Command Enter, or Control Enter on Windows, to run just that. Copy button to take code to your own machine, reload button to undo your edits. One thing specific to this deck: a couple of cells deliberately produce an error, because the error is the lesson. When you hit one, read it rather than assuming you mistyped. I will flag them as we reach them, but do not be alarmed when red text appears on a slide that is working exactly as intended.

</details>

+ The box area with a hint of blue as the background color is where you can write code (hereafter referred to as the "code area").
+ Hit the "Run Code" button to execute all the code inside the code area.
+ You can evaluate (run) code selectively by highlighting the parts you want to run and hitting Command + Enter for Mac (Ctrl + Enter for Windows).
+ If you want to run the codes on your computer, you can first click on the icon with two sheets of paper stacked on top of each other (top right corner of the code chunk), which copies the code in the code area. You can then paste it onto your computer.
+ You can click on the reload button (top right corner of the code chunk, left to the copy button) to revert back to the original code.

:::
<!--end of panel-->

## Data Preparation

<details class="transcript">
<summary>Transcript</summary>

One dataset for the whole lecture. Pizzaplace comes from the gt package, and it is a year of pizza orders from twenty fifteen, with a date, a name, a size and a type. It is a good teaching dataset here for two reasons. The date column arrives as text rather than as a proper Date, which is exactly the problem the first half of this lecture solves. And the name, size and type columns are all strings with regular structure, which gives the second half something realistic to work on. Run the cell and have a look before we start.

</details>

We use the `pizzaplace` dataset, which is available in the `gt` package.

```{webr-r}
#| autorun: true
data(pizzaplace, package = "gt")
pizzaplace
```

# Date

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1200px></html>

## `Date`

::: {.panel-tabset}

### Motivation

<details class="transcript">
<summary>Transcript</summary>

The point of this tab is that these two things look identical and are not. Run both cells. The first gives you character, the second gives you Date. On screen they may print much the same way, but R treats them completely differently. The three benefits underneath are why you should care. You can do calendar arithmetic, which means adding three months without worrying about how long February is. You can filter chronologically, so later than a given date actually means later rather than alphabetically after. And reformatting becomes one function call. Text that merely looks like a date gives you none of that.

</details>

R has an object class called `Date`.

This is a date as `character`.

```{webr-r}
a_date <- "2024-09-12"
class(a_date)
```

<br>

This is a date as `Date`.

```{webr-r}
#--- date as Date ---#
a_date_as_Date <- as.Date(a_date)
class(a_date_as_Date)
```

<br>

Recording dates as a `Date` object instead of a string has several benefits:

+ calendar math is possible with `Date` objects
+ you can `filter()` based on the chronological order of dates
+ converting a date into another format is easy

### `as.Date()`

<details class="transcript">
<summary>Transcript</summary>

Dates arrive in whatever format somebody felt like using, and the four examples at the top are all the same day. as.Date needs you to describe the format you have, using the percent codes listed. Lowercase d is day, lowercase m is numeric month, lowercase b is the abbreviated month name, uppercase B is the full name, lowercase y is a two-digit year and uppercase Y is four digits. Look at the example: Dec fifteen ten, with format percent b percent d percent y. You are not telling R what the date is. You are telling it how to read what you have.

</details>

Dates (as string) come in various formats. Several of them are:

+ `2010-12-15`
+ `12/15/2010`
+ `Dec 15 10`
+ `15 December 2010`

They all represent the same date. 

We can use `as.Date()` to transform dates stored as characters into `Date`s. 

```{r eval = F}
#--- NOT RUN ---#  
as.Date(date in character, format)
```

<br>

In `format` you specify how day, month, and year are represented in the date characters you intend to convert using special symbols including:

+ `%d`: day as a number (01-31)
+ `%m`: month (01, 02, $\dots$, 12)
+ `%b`: abbreviated month (Jan, $\dots$, Dec)
+ `%B`: unabbreviated month (January, $\dots$, December)
+ `%y`: 2-digit year (96 for 1996, 02 for 2002)
+ `%Y`: 4-digit year (1996, 2012)

**Example**

```{webr-r}
#--- convert from a character to a Date ---#
(
a_date <- as.Date("Dec 15 10", format = "%b %d %y")
)

#--- check the class ---#
class(a_date)
```

### `lubridate`

<details class="transcript">
<summary>Transcript</summary>

And here is the easier way. With lubridate you do not describe the format at all. You pick the function whose name matches the order the pieces appear in. Year month day for the first, month day year for the second and third, day month year for the fourth. Look carefully at the third one, though. Month day year handles Dec fifteen ten perfectly well, even though that is a month name and a two-digit year, with no format string anywhere. That is the whole appeal. lubridate works out the details, as long as you tell it the order.

</details>

Alternatively, you can use the `lubridate` package to easily convert dates recorded in characters into `Date`s.

Using `lubridate`, you do not need to provide the format information unlike `as.Date()`

Instead, you simply use `y` (year), `m` (month), `d` (day) in the order they appear in the dates in character.

<br>

**Example**

```{webr-r}
ymd("2010-12-15")
mdy("12/15/2010")
mdy("Dec 15 10")
dmy("15 December 2010") 
```

### format for presentation

<details class="transcript">
<summary>Transcript</summary>

Converting text into a Date is one direction. This is the other. You have a proper Date and you want it to appear a particular way in a figure or a table. The format function takes the same percent codes you have just learned and runs them backwards, turning a Date into text of your choosing. Look at the two examples and see how differently the same date can be presented. And note the sentence above the code, because it matters for the second half of this lecture. You can reformat dates with string manipulation, and later we watch somebody try, but format is the right tool.

</details>
  
It is often the case that date values are not formatted in the way you want (e.g., when you are creating figures). 

While you can use string manipulation functions to reformat dates (which we learn next in this lecture), it is easier to just use the `format()` function.

<br>

```{r eval = F}
#--- NOT RUN ---#  
format(Date, format)
```

<br>

You can use the same rule for the `format` argument as the one we saw earlier when using `as.Date()`.

<br>

**Example**

```{webr-r}
pizzaplace %>% 
  #--- convert to Date ---#
  mutate(date = ymd(date)) %>% 
  #--- format ---#
  mutate(date_txt_f1 = format(date, "%m/%d/%y")) %>% 
  mutate(date_txt_f2 = format(date, "%m %B %Y")) %>% 
  select(date_txt_f1, date_txt_f2)
```

### components

<details class="transcript">
<summary>Transcript</summary>

Often you do not want the whole date, you want a piece of it. The year for grouping, the month for a seasonal comparison, the day of week to see whether Fridays are busier. The lubridate helpers on the left pull each piece out. Note the two people forget exist. yday gives day of year, one through three hundred and sixty five, which is what you want for anything seasonal. And wday gives day of week. Run the example and look at the columns side by side for the same dates, because seeing mday and yday next to each other makes the distinction obvious.

</details>

::: {.columns}

::: {.column width="40%"}
You can extract components (year, month, day) from a `Date` object using various helper functions offered by `lubridate`. 

+ `year()`: year
+ `month()`: month
+ `mday()`: day of month 
+ `yday()`: day of year
+ `wday()`: day of week
:::
<!--end of the 1st column-->
::: {.column width="60%"}
**Examples**

```{webr-r}
#| autorun: true
pizzaplace %>% 
  mutate(date = ymd(date)) %>%
  distinct(date) %>% 
  filter(month(date) >= 8) %>% 
  mutate(
    year = year(date),
    month = month(date),
    mday = mday(date),
    yday = yday(date),
    wday = wday(date)
  ) %>% 
  select(year, month, mday, yday, wday)
```
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->



### math

<details class="transcript">
<summary>Transcript</summary>

This is the tab that justifies the whole first half of the lecture. Three nested tabs. You can add and subtract periods, so plus years three does the sensible thing across leap years without you thinking about it. You can build sequences with seq, stepping by days, weeks or years, which is how you make a complete calendar to join other data onto. And you can compare dates inside filter and get chronological order rather than alphabetical. Try any of these on the original character column and you get an error or nonsense. That is the difference a proper Date makes.

</details>

Unlike dates in character, you can do some math on `Date` objects.

::: {.panel-tabset}

#### addition and subtraction

You can use `years()`, `months()`, `days()` from the `lubridate` package to add specified years, months, and days, respectively.

<br>

```{webr-r}
ymd("2015-05-01") + years(3)
ymd("2015-05-01") - months(3)
ymd("2015-05-01") + days(1)
```

#### sequence of dates

You can use `seq()` to create a sequence of dates, where the incremental step is defined by the `by` option.

<br>

```{webr-r}
seq(ymd("2015-05-01"), ymd("2020-05-01"), by = "years")

seq(ymd("2015-05-01"), ymd("2015-06-01"), by = "weeks")

seq(ymd("2015-05-01"), ymd("2015-05-09"), by = "days")
```

#### filter (logical evaluation)

<br>

```{webr-r}
pizzaplace %>% 
  #--- convert to Date ---#
  mutate(date = ymd(date)) %>%  
  filter(date >= as.Date("2015-05-01"))
```

:::
<!--end of panel-->

:::
<!--end of panel-->

# Strings manipulation

<html><div style='float:left'></div><hr color='#EB811B' size=1px width=1200px></html>

## String manipulation 

::: {.panel-tabset}

### Introduction

<details class="transcript">
<summary>Transcript</summary>

Now the second half. For strings we use stringr, which comes with tidyverse, so if you have loaded tidyverse you already have it. The right column lists what we cover, grouped by what you are trying to do: join and split, mutate, detect, and manage lengths. That grouping is worth internalising, because the function names then almost write themselves, and almost everything is str underscore something. And do bookmark the cheatsheet linked on the left. String functions are the sort of thing you look up every time rather than memorise, and one page beats searching the internet.

</details>
::: {.columns}

::: {.column width="50%"}
**Package**

For string (character) manipulation, we use the `stringr` package, which is part of the `tidyverse` package. So, you have installed it already.

`stringr` is loaded automatically when you load `tidyverse`. So, just load `tidyverse`.

```{r eval = F}
library(tidyverse)  
```

<br>

**Resources**

+ [stringr cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/strings.pdf)
+ [stringr packagedown website](https://stringr.tidyverse.org/index.html)
:::
<!--end of the 1st column-->
::: {.column width="50%"}
**Functions**

Here are the select functions we learn in this lecture:

+ join and split
  * `stringr::str_c()`
  * `stringr::str_split()` (`tidyr::separate()`)
+ mutate strings
  * `stringr::str_replace()`
+ detect matches
  * `stringr::str_detect()`
+ manage lengths
  * `stringr::str_trim()`
  * `stringr::str_pad()`
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


### Concatenate

<details class="transcript">
<summary>Transcript</summary>

Joining strings together, with str underscore c. Read the sentence at the top carefully, because it has been corrected. str underscore c is close to paste, but not the same, and the difference shows up on the join three tab. Work through the four nested tabs in order. Join one is the basics, with a separator. Join two introduces recycling a single string against a vector. Join three is where it gets interesting, because two of those cells now fail deliberately. And use cases shows the two things you will actually do with this: building a combined variable, and generating file names.

</details>

`stringr::str_c()` lets you concatenate a vector of strings. It is close to `paste()`, with one important difference you will meet on the "join 3" tab: `str_c()` refuses to recycle vectors of unequal length, whereas `paste()` does it silently. <!-- CHANGED: said "It is basically the same as paste()". Since stringr 1.5.0 that is no longer true in the one respect this slide goes on to demonstrate. -->

::: {.panel-tabset}

#### join 1

**concatenate**

```{webr-r}
stringr::str_c("R", "rocks")
```

<br>

**order matters**

```{webr-r}
stringr::str_c("rocks", "R")
```

<br>

**separator**

```{webr-r}
stringr::str_c("R", "rocks", sep = "+")
```

<br>

**more than two strings**


```{webr-r}
stringr::str_c("R ", "rocks", ", right", "?", sep = "")
```

#### join 2

**a string and a vector of strings**

```{webr-r}
verbs <- c("sucks", "rocks", "is just okay")

stringr::str_c("R", verbs, sep = "+")
```

+ Each of the vector elements (`verbs`) is concatenated with the single string (`"R"`), which is recycled because it has length 1
+ The separator (`"+"`) is applied to all the vector elements

<br>

**collapsing a vector of strings to a single string**

```{webr-r}
stringr::str_c("R", verbs, sep = "+", collapse = "%")
```

+ The `collapse` option collapses all the vector elements into a single string with the collapse separator (here, `%`) placed between the individual vector elements
+ `sep = "+"` is applied when concatenating a vector of strings and a string, and `collapse = "%"` is applied when concatenating the resulting vector of strings. 

#### join 3

**two vectors of equal length**

```{webr-r}
software_types <- c("R", "Python", "SAS")

stringr::str_c(software_types, verbs, sep = " ")
```

+ The `n`th element of one vector (`software_types`) is paired with the `n`th element of the other (`verbs`).

<br>

<!-- CHANGED: both examples below used to rely on str_c() recycling two vectors
of unequal length. stringr 1.5.0 adopted vctrs recycling rules, under which only
a length-1 vector recycles. Both cells now ERROR:

    str_c(software_types, verbs, sep = " ")
    #> Error: Can't recycle `..1` (size 5) to match `..2` (size 3).

Verified on stringr 1.5.1, which is also the version the wasm repo serves to
students, so these cells failed for everybody. Rewritten to teach the rule that
actually applies, and the all-combinations example now builds the pairs
explicitly with expand.grid() rather than relying on 5 and 3 being coprime. -->

**two vectors of different lengths**

```{webr-r}
software_types <- c("R", "Python", "SAS", "STATA", "Matlab")

#--- 5 against 3: this is an ERROR ---#
stringr::str_c(software_types, verbs, sep = " ")
```

::: {.callout-important title="`str_c()` will not recycle unequal lengths"}
Only a **length-1** vector gets recycled. Anything else must match exactly, or `str_c()` stops:

```
Error: Can't recycle `..1` (size 5) to match `..2` (size 3).
```

This is deliberate. `paste()` *will* recycle 5 against 3 and silently give you five results built from a repeating pattern, which is very often not what anyone meant. `str_c()` makes you say what you want.
:::

<br>

**all combinations**

If you genuinely want every pairing, build the pairs first and then join them:

```{webr-r}
combos <- expand.grid(
  software = software_types,
  verb = verbs,
  stringsAsFactors = FALSE
)

stringr::str_c(combos$software, combos$verb, sep = " ")
```

+ `expand.grid()` produces all 5 $\times$ 3 = 15 pairings, so both columns are the same length and `str_c()` is happy
+ The intent is visible in the code, rather than depending on how two lengths happen to divide into each other

#### use cases

::: {.panel-tabset}

##### Concatenating string variables in a dataset

Sometimes, you want to concatenate two (or more) string variables into one variable. 

For example, suppose you would like to combine pizza `size` and `type` into a single variable to make it easier to create faceted figures by `size`-`type`.

<!-- CHANGED: added autorun. This cell creates `type_size`, and the Split tab
further down uses `pizzaplace$type_size`. Without autorun, a student who goes
straight to Split gets a NULL column and a confusing result, because the
variable was only ever created on a tab they had not clicked. -->

```{webr-r}
#| autorun: true
pizzaplace <- mutate(pizzaplace, type_size = stringr::str_c(type, size, sep = "-"))

pizzaplace
```

##### Reading files

You can use `stringr::str_c()` to create a vector of file names that have a common pattern. 

For example suppose you have files that are named following this convention: "corn_yield_X.csv", where X represents year. 

You have such csv files starting from 2000 to 2020. Then,

```{r }
file_names <- stringr::str_c("corn_yield_", 2000:2020, ".csv")  
```

```{r }
head(file_names) 
```

Now, you can easily read each of them iteratively using a loop.

:::
<!--end of panel-->

:::
<!--end of panel-->

### Split

<details class="transcript">
<summary>Transcript</summary>

The reverse operation. str underscore split takes a string and a pattern, and breaks the string apart wherever the pattern occurs. Run the first cell and look at what comes back, because it is a list, one element per input string, each holding the pieces. That is often awkward to work with. So read the sentence underneath. If what you actually want is to turn one column into two columns of a data frame, separate from tidyr is the better tool, because it hands you a data frame rather than a list you then have to unpack. Same operation, output shaped for where it is going.

</details>

`stringr::str_split()` splits a string based on a pattern you provide:

<br>

```{webr-r}
stringr::str_split(pizzaplace$type_size, "-") %>% head()
```

<br>

But, if you are splitting a variable into two variables, `tidyr::separate()` is a better option.

<br>

```{webr-r}
pizzaplace %>% 
  tidyr::separate(type_size, c("type_2", "size_2"), "-") %>% 
  dplyr::select(id, type, type_2, size, size_2)
```

### Replace 

<details class="transcript">
<summary>Transcript</summary>

Replacing part of a string, and this tab has a genuine trap in it. The introduction is straightforward. str underscore replace swaps the first match, str underscore replace all swaps every match. Then the use case tries to turn a four-digit year into two digits by deleting the characters two and zero. Read the red callout, because that approach quietly breaks. On the twentieth of any month the first two-zero it finds is the day, not the year, so you get a mangled date and no warning whatsoever. Twelve days a year. The tool worked. The pattern was simply not specific enough.

</details>

::: {.panel-tabset}

#### Introduction

**How**

You can use `stringr::str_replace()` to replace parts of the texts matched with the user-specified texts.

<br>

```{r eval = F}
#--- Syntax ---#
stringr::str_replace(string, pattern, replacement)
```

<br>

**Example**

```{webr-r}
strings_vec <- c("I think R rocks, but some think it does not rock", "Python also rocks, but Julia does not rock")  

stringr::str_replace(strings_vec, "rock", "rock big time")
```

<br>

Note that only the first occurrence of "rock" in each of the string vector element was replaced with "rock big time." 

You need to use `stringr::str_replace_all()` to replace all the occurrences. 

```{webr-r}
stringr::str_replace_all(strings_vec, "rock", "rock big time") 
```

#### Use case

Suppose you would like to have a particular format of date in a figure you are trying to create using `pizzaplace`: e.g., 07/08/20 (month, day, year without the first 2 digits).

Pretend that `date_text` is the variable that indicates date and it looks like this:

<br>

```{webr-r}
#| autorun: true
pizzaplace_for_plot <- 
  mutate(pizzaplace, date_text = ymd(date) %>% 
  format("%m/%d/%Y")) 

pizzaplace_for_plot$date_text %>% head()
```

<br>

So, you would like to replace "20" with "" (nothing).

<br>

```{webr-r}
pizzaplace_for_plot %>%
  mutate(date_text = stringr::str_replace(date_text, "20", "")) %>%
  select(id, date_text, size, type)
```

<br>

<!-- ADDED: this example is presented as if it works, and on most rows it does.
It does not on the 20th of any month: str_replace() removes the FIRST "20" it
finds, which for "01/20/2015" is the day, giving "01//2015". 12 of the 358
distinct dates in pizzaplace come out invalid. Left the demonstration in place,
since replacing the first match is exactly the behaviour being taught, but the
failure is now shown rather than hidden. -->

::: {.callout-important title="Look at the 20th of the month"}
`str_replace()` removes the **first** match, and on a date like `01/20/2015` the first "20" is the [day]{style="color: red;"}, not the year:

```
01/20/2015  ->  01//2015
```

Twelve dates a year come out broken this way, and nothing warns you. Try it yourself:

```{webr-r}
stringr::str_replace("01/20/2015", "20", "")
```

The lesson is not that `str_replace()` is faulty. It did exactly what you asked. The lesson is that matching on a bare "20" was never a precise enough way to say "the century digits of the year".
:::

<br>

Which is why, for a job like this, you should not be manipulating the text at all. From `pizzaplace`, you could have just done this:

<br>

```{webr-r}
pizzaplace %>% 
  mutate(
    date = ymd(date),
    date = format(date, "%m/%d/%y")
  ) %>% 
  .$date %>% head()
```

:::
<!--end of panel-->

### Detect 

<details class="transcript">
<summary>Transcript</summary>

Detecting whether a pattern appears. str underscore detect takes strings and a pattern and gives you back TRUE and FALSE, which makes it the natural thing to feed into filter, or to use for subsetting. The two use cases are both realistic. The first picks out just the soy files from a folder that also holds corn files, and notice the loop reads all files bracket is soy, not all files, because the entire point was to exclude the corn. The second builds a grouping variable with case when, which is the pattern you will use constantly for messy category labels.

</details>

::: {.panel-tabset}

#### Introduction

You can use `stringr::str_detect()` to check whether a user-specified text appears inside each of a set of strings.

It takes a vector of strings and a text pattern, and returns a vector of TRUE/FALSE.

<br>

**Example**

```{webr-r}
fruit <- c("apple", "banana", "pear", "pineapple")

stringr::str_detect(fruit, "apple")
``` 

#### use cases

::: {.panel-tabset}

##### get the list of file names 

First clone [this repository](https://github.com/tmieno2/quarto-examples). 

Inside **supplementary-material/data/data-for-loop-demo**, there are two sets of files in a single folder: `corn_experiment_x.rds` and `soy_experiment_y.rds`, where both `x` and `y` range from 1 to 30.

You want to read only the soy files. 

First, let's get the name of the whole list of files in the working directory:

```{r}
all_files <- 
  list.files(
    here::here("supplementary-material/data/data-for-loop-demo"),
    full.names = TRUE
  )

head(all_files, 2)
tail(all_files, 2)
```

<br>

Now use `stringr::str_detect()` to find which elements of `all_files` include "soy."

```{r }
is_soy <- stringr::str_detect(all_files, "soy")  
```

<br>

Okay so, here is the list of all the "soy" files:

```{r }
all_files[is_soy]
```

<br>

Now, you can loop to read just those files.

<!-- CHANGED: this looped over `all_files`, not `all_files[is_soy]`, so it read
all 60 files (60,000 rows) including every corn file the tab had just gone to
the trouble of filtering out. The intended soy-only result is 30 files and
30,000 rows. As written, the str_detect() work on this tab was discarded on the
very next line. -->

```{r}
(
soy_data <-
  lapply(all_files[is_soy], \(x) readRDS(x)) %>%
  bind_rows()
)
```

::: {.callout-note title="Check you got what you asked for"}
Thirty soy files, not sixty files. If you had looped over `all_files` here, everything would still have run without complaint — you would simply have silently read the corn data back in alongside it.
:::

##### Define a group from a variable 

Consider the following dataset of plant genes.

```{r }
gene_data <- expand.grid(
  id = c("Zm_1", "Zm_2"), 
  gene = c("20_WW_BL_TP1", "20_WW_BL_TP", "20_WW_ML_TP1", "20_WW_ML_TP", "20_WW_TL_TP1", "20_WW_TL_TP3")
) 
```

```{r echo = F}
gene_data 
```

<br>

There are three different types of genes: those that have `_BL_`,`_ML_`, and `_TL_`. The objective here is to make a variable that indicates gene group from the `gene` variable.

```{r }
gene_data %>% 
  mutate(gene_group = case_when(
    stringr::str_detect(gene, "_BL_") ~ "BL",
    stringr::str_detect(gene, "_ML_") ~ "ML",
    stringr::str_detect(gene, "_TL_") ~ "TL"
  ))
```

:::
<!--end of panel-->

:::
<!--end of panel-->

### letter case 

<details class="transcript">
<summary>Transcript</summary>

Three small functions for changing case. To upper, to lower, and to title, which capitalises the first letter of each word. They look trivial, and individually each one is. What makes them worth a slide is that inconsistent capitalisation is one of the most common reasons a join silently fails. If one dataset says Lancaster and another says LANCASTER, those are different strings as far as R is concerned, and your join drops both rows without complaining. Forcing everything to one case before you join is a cheap habit that prevents a category of bug that is genuinely hard to spot afterwards.

</details>

Here is the collection of functions that let you change the letter case of strings.

<br>

**To upper case**

```{webr-r}
pizzaplace %>% 
  mutate(name = stringr::str_to_upper(name))
```

<br>

**To lower case**

```{webr-r}
pizzaplace %>% 
  mutate(size = stringr::str_to_lower(size))
```

<br>

**Only the first letter is capitalized**

```{webr-r}
pizzaplace %>%  
  mutate(type = stringr::str_to_title(type)) 
```

### padding

<details class="transcript">
<summary>Transcript</summary>

Padding makes strings a fixed width by adding a character until they reach the length you specify. The examples pad with question marks and dashes, which is deliberately silly, but the real use is almost always the same one: padding numbers with leading zeros. If you have identifiers one through one hundred and you pad them to three characters with zeros, they sort correctly as text, and they match identifiers in other datasets that were stored that way. Note the side argument, which controls whether padding goes left, right or both. Left is what you want for leading zeros.

</details>

You can pad strings with symbols of your choice so the resulting strings are of the length you specify.

```{r eval = F}
#--- NOT RUN ---#
stringr::str_pad(strings, string length, side, padding symbol)
```

<br>

**Examples**

```{webr-r}
stringr::str_pad("I am sick of R", 20, side = "left", pad = "?") 
```

<br>

```{webr-r}
stringr::str_pad("I am sick of R", 30, side = "both", pad = "?") 
```

<br>

```{webr-r}
stringr::str_pad("I am sick of R", 20, side = "right", pad = "-") 
```

:::
<!--end of panel-->

## Exercises 

::: {.panel-tabset}

### Data preparation

<details class="transcript">
<summary>Transcript</summary>

A small dataset built for the exercises, and it is worth seeing how it was made. We take every date from the first of April to the thirtieth of September in twenty twelve, pull out the year, month, day and day of year from each, and then throw the actual date column away. So you have the pieces of a date but not the date itself. That is the situation the two exercises put you in, and it is a realistic one. Plenty of datasets record year and day of year in separate columns and leave you to reassemble what they mean.

</details>

We will work with the following data:

<br>

```{webr-r}
#| autorun: true
(
  date_data <-
    tibble(
      date = seq(ymd("2012-04-01"), ymd("2012-09-30"), by = "days")
    ) %>%
    mutate(
      year = year(date),
      month = month(date),
      day = mday(date),
      day_of_year = yday(date)
    ) %>%
    select(-date)
)
```

### Exercise 1

<details class="transcript">
<summary>Transcript</summary>

The first exercise goes the string route. You have year, month and day as separate numbers, and you want a Date. So join them into one piece of text with str underscore c and a separator, then hand that text to lubridate. Think about which lubridate function you want, given the order you have chosen to put the pieces in. One thing you may notice when you run it: the intermediate text does not always look tidy, because month four prints as four rather than zero four. Try it anyway, and see whether lubridate actually minds.

</details>

Use `stringr::str_c()` to combine, `year`, `month`, and `day` using "-" as the separator and convert the combined text to Date using `lubridate`.

<br>

::: {.panel-tabset}
### Work here

```{webr-r}

```

### Answer

```{r, eval = FALSE}
#| code-fold: true
date_data %>%
  mutate(date_as_str = str_c(year, month, day, sep = "-")) %>%
  mutate(date_as_Date = ymd(date_as_str)) %>%
  select(date_as_Date)
```


:::
<!--end of panel-->
### Exercise 2

<details class="transcript">
<summary>Transcript</summary>

The second exercise takes the other route, and it is the more interesting one, because it uses date arithmetic rather than string work. You have the year and the day of year, so you can build the first of January for that year and then add the day of year to it. Watch the off-by-one, though. The first of January is day one, not day zero, so adding the day of year directly overshoots by exactly one day. That minus one at the end of the answer is not a fudge. It is the difference between counting days and counting offsets.

</details>

Using `Date` math to recover the dates from `year` and `day_of_year`.

<br>

::: {.panel-tabset}

### Work here

```{webr-r}

```


### Answer
```{r, eval = FALSE}
#| code-fold: true
date_data %>%
  mutate(first_day_of_year = ymd(str_c(year, "01-01"))) %>%
  mutate(date = first_day_of_year + day_of_year - 1)
```
:::
<!--end of panel-->

:::
<!--end of panel-->



