---
title: "09-2: R as GIS: Raster Data Basics"
abstract: "Chapter 9: Spatial Data"
format: 
  revealjs: 
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    include-in-header:
      - ../transcript-support.html
      - ../webr-layout.html
      - ../katex-amd-fix.html
    footer: "[back to the lecture slides on the course website](https://tmieno2.github.io/Data-Science-with-R-Quarto/lectures/)"
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    chalkboard: true
webr:
  packages: ['ggplot2', 'dplyr', 'terra', 'stars', "raster", "tidyterra", "sf"] # Install R packages on document open
# autoload-packages: false       # Disable automatic loading of packages
# show-startup-message: false    # Disable displaying status of webR initialization
  cell-options:
    editor-font-scale: 0.7
    dpi: 216
    fig-width: 7
    fig-height: 4.5
    out-width: 100%
filters:
  - webr
---

```{r, include = FALSE}
library(terra)
library(dplyr)
library(stars)
library(data.table)
```

## Before you start

<details class="transcript">
<summary>Transcript</summary>

Raster data is the other half of GIS, and it is a completely different animal from what we did last time. A raster is a grid: fixed rows and columns of equally sized cells, each holding a number. A raster has no vector-style geometry column or row-per-feature attribute table, though categorical layers can have associated category tables. Otherwise, it is a rectangle of values plus enough metadata to say where on Earth that rectangle sits. That difference drives everything in this deck. We will cover what the object looks like, how to read and write the various file formats weather and satellite data arrive in, and then three operations you will actually reach for: arithmetic across layers, aggregating to a coarser resolution, and resampling one grid onto another.

</details>

<br>

### Learning objectives

Learn how to handle raster datasets using R.

<br>

::: {.columns}

::: {.column width="50%"}
### Table of contents

+ [Raster data basics](#sec-basics)
   - object type
   - accessing metadata
   - quick plotting
+ [Input and output](#sec-io)
   - read raster data files
   - write to raster data files
+ [Raster operations](#sec-operations)
   - arithmetic operations
   - aggregate
   - resample
   - merge
:::
<!--end of the 1st column-->
::: {.column width="50%"}
### Related Contents (Links)

+ [Create maps from raster data using `ggplot2`](https://tmieno2.github.io/R-as-GIS-workshop/L03_create_map.html)
+ [Vector and raster data interactions](https://tmieno2.github.io/R-as-GIS-workshop/L05_vector_raster_interaction.html)
  + cropping
  + masking
  + value extraction
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

```{webr-r}
#| context: setup
 
#--- install and library the data package ---#
#--- Both repos are required. r.spatial.workshop.datasets declares         ---#
#--- Imports: sf, and webr::install() resolves dependencies ONLY in the     ---#
#--- repos named here. With the custom repo alone, sf cannot be found and   ---#
#--- the install never completes, so nothing downstream ever runs.          ---#
webr::install(
  "r.spatial.workshop.datasets",
  repos = c(
    "https://tmieno2.github.io/rwasm-package/repo",
    "https://repo.r-wasm.org"
  )
)

library("r.spatial.workshop.datasets")

data(reflec_blue)
reflec_blue <- terra::rast(reflec_blue)
data(reflec_red)
reflec_red <- terra::rast(reflec_red)
data(reflec_green)
reflec_green <- terra::rast(reflec_green)
data(NIR)
NIR <- terra::rast(NIR)
data(RED)
RED <- terra::rast(RED)
reflec_all <- c(reflec_blue, reflec_red, reflec_green)

#--- ADDED: explicit data() calls for the two prism objects, matching the      ---#
#--- lines above. NOTE: not a bug fix. r.spatial.workshop.datasets is          ---#
#--- LazyData: true, so terra::rast(prism_saunders) already worked without     ---#
#--- one — confirmed in the WebR runtime itself.                               ---#
data(prism_saunders)
prism_saunders <- terra::rast(prism_saunders)
data(prism_douglas)
prism_douglas <- terra::rast(prism_douglas)

# Mirror of lecture_theme() in lectures/_lecture-theme.R. WebR runs in the
# browser against a virtual filesystem that does not hold the repo, so it
# cannot source that file and has to carry a copy. Edit the shared file first,
# then run `Rscript lectures/check-lecture-theme.R`, which fails if this copy
# has drifted and prints the block to paste.
lecture_theme <- function(base_size = 16) {
  ggplot2::theme_bw(base_size = base_size) +
    ggplot2::theme(
      axis.text = ggplot2::element_text(size = base_size * 0.9),
      axis.title = ggplot2::element_text(size = base_size),
      legend.text = ggplot2::element_text(size = base_size * 0.9),
      legend.title = ggplot2::element_text(size = base_size * 0.9),
      plot.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.key = ggplot2::element_rect(fill = "transparent", colour = NA),
      # theme_bw's grey92 grid was chosen against a white panel and all but
      # disappears on the warm paper once the panel stops being opaque.
      panel.grid = ggplot2::element_line(colour = "grey87")
    )
}

# WebR draws its canvas at 72 dpi whatever `dpi` says, so on the 216-dpi canvas
# every size ggplot measures in points or millimetres comes out a third of what
# was asked for. Scaling ggplot2's own mm-to-point constants and the theme by
# the same factor puts WebR figures at the same type size as the knitr ones.
# Keep this in step with `dpi` in the header: change one, change the other.
webr_scale <- 216 / 72
assignInNamespace(".pt", ggplot2::.pt * webr_scale, "ggplot2")
assignInNamespace(".stroke", ggplot2::.stroke * webr_scale, "ggplot2")
theme_lecture <- lecture_theme(18 * webr_scale) +
  # base_size also sets the theme's line widths, which are in millimetres and
  # already pass through the scaled .pt; pin them to what 18pt gives, or they
  # come out nine times too thick.
  ggplot2::theme(
    line = ggplot2::element_line(linewidth = 18 / 22),
    rect = ggplot2::element_rect(linewidth = 18 / 22)
  )
ggplot2::theme_set(theme_lecture)
```

## Tips to make the most of the lecture notes

<details class="transcript">
<summary>Transcript</summary>

Same arrangement as the other decks. The blue-tinted boxes are live R sessions running in your browser, so hit Run Code and things happen. One warning specific to this deck: it loads terra, stars, raster, sf and tidyterra, which is a lot of packages to download, and satellite imagery on top of that. The first cell you run may take a couple of minutes before anything appears. That is the download, not a hang. And because every cell shares one session, order matters more here than usual, so work top to bottom the first time through.

</details>

<br>

### Interactive navigation tools

+ Click on the three horizontally stacked lines at the bottom left corner of the slide, then you will see table of contents, and you can jump to the section you want

+ Hit letter "o" on your keyboard and you will have a panel view of all the slides

<br>

### Running and writing codes

+ The box area with a hint of blue as the background color is where you can write code (hereafter referred to as the "code area").
+ Hit the "Run Code" button to execute all the code inside the code area.
+ You can evaluate (run) code selectively by highlighting the parts you want to run and hitting Command + Enter for Mac (Ctrl + Enter for Windows).
+ If you want to run the codes on your computer, you can first click on the icon with two sheets of paper stacked on top of each other (top right corner of the code chunk), which copies the code in the code area. You can then paste it onto your computer.
+ You can click on the reload button (top right corner of the code chunk, left to the copy button) to revert back to the original code.

## The `terra` and `raster` packages: Basics {#sec-basics}

::: {.panel-tabset}

### `terra` and `raster`

<details class="transcript">
<summary>Transcript</summary>

There are two packages for raster data and you need to know both names, though you will only write one of them. The raster package came first, terra replaced it, and the same person wrote both. terra is faster at essentially everything and is what we will use. So why mention raster at all? Because a decent number of packages, especially older ones, only accept raster's object classes and will refuse a terra object outright. When you hit that, you convert, which is a one-liner we will see in a moment. Think of raster as a language you read but do not speak.

</details>

+ The most popular R packages to handle raster data is the `raster` and `terra` packages. They are authored by the same person and `terra` is the successor of `raster`. 

+ The `terra` package is now mature and does everything faster than `raster`

+ There are lots of packages that still depend on `raster` and do not work well with R object class defined by `terra`. 

+ We primarily learn how to use the `terra` package

### Object class

<details class="transcript">
<summary>Transcript</summary>

Here is where the two packages diverge in a way you will notice. terra has exactly one class, SpatRaster, and it handles one layer or fifty layers with no distinction. The raster package splits that across three classes: RasterLayer for a single layer, then RasterStack and RasterBrick for multiple. That is three names for what terra calls one thing, which is a fair summary of why terra replaced it. You do not need to know when to use a Brick versus a Stack. You only need to recognise these names when an error message or an old package's documentation mentions them.

</details>

::: {.columns}

::: {.column width="50%"}
The `terra` and `raster` packages use different R object classes to represent raster data:

+ `terra`
  + `SpatRaster`
+ `raster` (collectively referred to as `Raster`*)
  + `RasterLayer`
  + `RasterStack`
  + `RasterBrick`
:::

:::{.callout-note}
+ We will learn how to convert `SpatRaster` to `Raster`* and vice versa.
+ This is important as there are (and will be) packages that accept only `Raster`*, especially the ones that are old.
:::

::: {.column width="50%"}

:::

:::

### `SpatRaster`

<details class="transcript">
<summary>Transcript</summary>

Print a SpatRaster and you get a compact block of metadata, and it repays reading carefully. Dimensions tell you rows, columns and layers. Resolution is the size of a single cell, not the whole grid, and it is in whatever units the CRS uses, so degrees for lon-lat data and metres for a projected one. Extent is the bounding box. The field I want you to notice is source. If it says memory, the values are in RAM. If it names a file, terra left them on disk and reads them on demand, which is how it copes with rasters bigger than your computer. Move that file and the object breaks.

</details>

::: {.panel-tabset}

#### Look at one

Let's take a look at a `SpatRaster` object, `reflec_blue`.

Check the class first:

```{webr-r}
class(reflec_blue)
```

<br>

What is inside?

```{webr-r}
reflec_blue
```

#### What the metadata means

Here are the explanations of the information provided:

+ class
+ dimensions
  + nrow: number of rows
  + ncol: number of columns
  + nlyr: number of layers
+ resolution: the size of a **single cell**, in the units of the CRS
  + x: cell width
  + y: cell height
+ extent: bounding box of the whole grid (just like what you get with `st_bbox(sf)`)
+ coord. ref.: CRS
+ source: `memory` if the values are held in RAM, otherwise the file they are read from
+ name: name of the attribute (`names`, plural, once there is more than one layer)
+ min value / max value: the range of the cell values

#### Watch `source`

`source` is the one worth noticing. `terra` does not necessarily read the cell
values into memory when you open a file — it can leave them on disk and fetch
them only as needed, which is how it works with rasters far larger than your RAM.

<br>

So a `SpatRaster` printing `source(s) : memory` holds its values; one printing a
file path is still tied to that file on your disk. Move or delete the file and
the object stops working:

```{r, eval = FALSE}
r <- terra::rast("data/reflec_blue.tif")   # source(s) : reflec_blue.tif
file.remove("data/reflec_blue.tif")
terra::values(r)
#> Error: [readStart] file does not exist: data/reflec_blue.tif
```

:::
<!--end of panel-->



### Quick plotting

<details class="transcript">
<summary>Transcript</summary>

Plot on a SpatRaster gives you a usable map in one word, with a colour scale picked for you. This is your first move whenever you load raster data, before any analysis. It answers questions you did not know you had: is the extent where you expected, are the values in a plausible range, is half the grid missing. In this deck we use it constantly as a check. When you want a map for a paper rather than a glance, that is ggplot and tidyterra, which is the next deck. For now, plot is enough.

</details>

You can use `plot()` to make a map quickly:

```{webr-r}
plot(reflec_blue)
```

It automatically color the grids by the value of the attribute (`blue`). 

### Multi-layer `SpatRaster`

<details class="transcript">
<summary>Transcript</summary>

Stacking layers is just c, the same function you use to build a vector, which is a genuinely nice piece of design. The requirement is that the layers line up exactly: identical dimensions, extent and resolution. If they do not, c refuses, and that refusal is doing you a favour, because a stack of misaligned grids would silently pair up cells that are not in the same place. When layers do not align, you resample first, which is later in this deck. After stacking, notice nlyr has gone to three and names lists all three attributes. Plot now draws a panel per layer.

</details>

You can simply use `c()` function to create a multi-layer `SpatRaster` just like you create a vector as long as all the layers have exactly the same dimensions, extent, and resolution.  

```{webr-r}
(
reflec_all <- c(reflec_blue, reflec_red, reflec_green)
)
```

Notice that `nlyr` is 3 now and you see three attribute names in `names`.

<br>

`plot()` create maps for all the attributes.

```{webr-r}
plot(reflec_all)
```

### `Raster`*

<details class="transcript">
<summary>Transcript</summary>

This is the conversion I promised. Going from terra to raster is a single as call with the string Raster, and it works out which of the three classes you need: a single-layer SpatRaster becomes a RasterLayer, a multi-layer one becomes a RasterBrick. Going the other way is terra's rast function, which accepts either. That is the whole story. You only need it when some package demands the old classes, and the symptom is an error complaining about the object type rather than anything you did wrong. Worth knowing this exists, so that a package refusing your SpatRaster costs you a minute instead of an afternoon.

</details>

::: {.panel-tabset}

#### Introduction

The `raster` package differentiates single-layer and multi-layer raster data. 

+ `RasterLayer`: single-layer
+ `RasterStack`: multi-layer
+ `RasterBrick`: multi-layer

#### Converting to `Raster`* 

You can convert a `SpatRaster` to a `Raster`* object using `as(SpatRaster, "Raster")`. 

```{webr-r}
(
reflec_blue_rl <- as(reflec_blue, "Raster")
)
```

Since, `reflec_blue` is a single-layer `SpatRaster`, it was converted into a `RasterLayer`.

<br>

---

```{webr-r}
(
reflec_all_rl <- as(reflec_all, "Raster")
)
```

Since, `reflec_all` is a multi-layer `SpatRaster`, it was converted into a `RasterBrick`.

<br>

:::{.callout-note}
You can convert a `RasterBrick` to a `RasterStack` by applying `stack()` to the `RasterBrick` if you want. But, you do not need to.
:::

#### Converting to `SpatRaster`

You can convert an `Raster`* object to `SpatRaster` using `terra::rast()` function.

```{webr-r}
terra::rast(reflec_all_rl)
```

:::
<!--end of panel-->

:::


## Basic operations

::: {.panel-tabset}

### Accessing the metadata

<details class="transcript">
<summary>Transcript</summary>

Everything printed in that metadata block has a function that returns it on its own, and these tabs walk through them: crs, ncol and nrow and nlyr and ncell, res, and ext. You use these constantly, not for display but inside code. The most important is crs. Many operations that combine a raster with an sf object require both to share a CRS, and they are not always polite about telling you why they failed. Pulling the raster's CRS with crs and transforming the vector data to match is a line you will write many times in the next few decks.

</details>

::: {.panel-tabset}

#### CRS

```{webr-r}
terra::crs(reflec_blue)
```

This is very useful. As we will see later, when interacting two spatial objects (e.g., extracting values from a raster data to `sf`) some functions require that the two spatial objects has the same CRS. You can use `terra::crs()` to get the CRS of the raster data and apply it to another spatial object.

#### dimensions

```{webr-r}
#--- number of columns ---#
terra::ncol(reflec_blue)

#--- number of rows ---#
terra::nrow(reflec_blue)

#--- number of layers ---#
terra::nlyr(reflec_blue)

#--- number of cells ---#
terra::ncell(reflec_blue)
```

#### resolution

```{webr-r}
terra::res(reflec_blue)
```

#### extent

This is like `st_bbox()` for `sf`.

```{webr-r}
terra::ext(reflec_blue)
```

:::
<!-- end of panel -->


### Extract cell information

<details class="transcript">
<summary>Transcript</summary>

Two things here. Square brackets get you cell values by cell number, and cell numbers run left to right along the top row, then the next row, the way you read a page. And xyFromCell converts a cell number into an actual coordinate. Read the note in the first tab, because the indexing looks like a vector but does not return one. Without an index you get a matrix, with an index a data frame, one column per layer, which is the price of a SpatRaster being able to hold many layers. Honestly, in practice you rarely need either function except to check your own arithmetic.

</details>

::: {.panel-tabset}

#### cell values

You can access the cell values using `[]` and a cell number, just like a vector. Note that `head()` is there to avoid printing all half a million cells to the console.

```{webr-r}
#--- first six values ---#
reflec_blue[] %>% head()
```

<br>

Yes, there are so many NAs in this raster data. Let's look at the value of 10100th through 10120th cells:

```{webr-r}
reflec_blue[10100:10120]
```

<br>

:::{.callout-note title="You index like a vector, but you do not get one back"}
`[]` **looks** like vector indexing and the cell numbering behaves that way, but
what comes back is rectangular, because a `SpatRaster` can hold many layers:

+ `reflec_blue[]` (no index) returns a **matrix**, one column per layer.
+ `reflec_blue[10100:10120]` returns a **data.frame**, again one column per layer.

So `reflec_blue[10100]` is a one-row data.frame, not a number. If you need a plain
numeric vector, ask for the column — `reflec_blue[10100:10120]$blue` — or use
`terra::values(reflec_blue, mat = FALSE)`.
:::

<br>

:::{.callout-note title="Note"}
+ We use this function to check our arithmetic operations are successful.
+ I have never had to use this function before in practice.
:::

#### coordinates of the cell centers

You can use `terra::xyFromCell()` to get the geographic coordinates of the cell centers.

<br>

**Syntax**

```{r, eval = F}
terra::xyFromCell(SpatRaster, cell)
```

<br>

+ `cell`: cell numbers

<br>

**Example**

```{webr-r}
terra::xyFromCell(reflec_blue, cell = 1:ncell(reflec_blue)) %>% head()
```

:::
<!--end of panel-->

### Subset

<details class="transcript">
<summary>Transcript</summary>

Subset pulls layers out of a multi-layer SpatRaster. You can name them, which is what I would encourage, or use their position numbers. Names are worth the extra typing, because a script that says quote blue is obvious when you reread it in six months while one that says three is a small mystery, and position numbers also break the moment you add or reorder a layer. Note that this selects whole layers, not areas. Cutting a raster down geographically is cropping, which is a different function in a later deck. Subset is about which variables you are carrying, not about which part of the map you keep.

</details>

You can access specific layers using `subset()`.

<br>

**Syntax**
```{r, eval = F}
subset(SpatRaster, subset)
```

+ `subset`: layer names or corresponding integers 

<br>

**Examples**

```{webr-r}
#--- get the second and third layer ---#
subset(reflec_all, c(2, 3))
```

<br>

```{webr-r}
#--- get the blue layer ---#
subset(reflec_all, "blue")
```


:::
<!--end of panel-->

## Raster data input and output {#sec-io}

::: {.panel-tabset}

### Introduction

<details class="transcript">
<summary>Transcript</summary>

Raster data arrives in an unreasonable number of file formats. GeoTIFF, with the tif extension, is the common one and what you will produce yourself. But PRISM weather data comes as BIL, Daymet often comes as netCDF with an nc extension, and there are plenty more. The good news is that this is almost entirely not your problem. terra sits on top of GDAL, which reads essentially everything, so the same function opens all of them and you rarely have to care what you were handed. The next tabs prove that by opening three different formats with identical code.

</details>

Raster data files can come in numerous different formats. 

+ The most common format is GeoTiff (.tif as the file extension)
+ PRISM weather data comes in the Band Interleaved by Line (BIL) format
+ Some of the Daymet data comes in netCDF format. 
+ Other popular formats include SAGA, ENVI, and many others. 

You can read data of almost all the existing file formats with the `terra` package.

### Read with `terra`

<details class="transcript">
<summary>Transcript</summary>

One function, rast, with a file path. Give it a vector of paths and you get a multi-layer object in one call, provided the files align. Then work through the three examples, and the point of them is how boring they are: the tif, the BIL and the netCDF all open with exactly the same code and only the path changes. That is the payoff of GDAL underneath. Do download the files and get the paths right, because path problems are the actual difficulty here, not formats. If a path fails, check your working directory before anything else.

</details>

::: {.panel-tabset}

#### How

You can use `terra::rast()` to read raster data files.

<br>

**single raster data file**

```{r, eval = FALSE}
drone_blue_sr <- terra::rast(path to the file)
```

<br>

**multiple raster data files**

```{r, eval = FALSE}
drone_blue_sr <- terra::rast(c(path to the file 1, path to the file 2, ...))
```

#### Let's try (.tif)

:::{.callout-note title="Instruction"}
+ download **reflec_blue.tif**, **reflec_red.tif**, and **reflec_green.tif** from this [link](https://www.dropbox.com/scl/fo/ljr76u5azr7u4zx7mk55p/AB1RtEs0KyWqL6rKsiJHRKk?rlkey=ogas3pbhuhhhy94kprbd2m553&st=oq35p7rx&dl=0).
+ find the path to the three files
:::

<br>

**single file**

It looks like this for me:

```{r, eval = F}
reflec_blue <- terra::rast("data-files-for-participants/reflec_blue.tif")
```

<br>

**multiple files**

It looks like this for me:

```{r, eval = F}
reflec_all <- 
  terra::rast(
    c(
      "data-files-for-participants/reflec_blue.tif",
      "data-files-for-participants/reflec_red.tif",
      "data-files-for-participants/reflec_green.tif"
    )
  )
```

#### Let's try (.bil)

:::{.callout-note title="Instruction"}
+ download the folder named **PRISM_ppt_stable_4kmD2_20120801_bil** from this [link](https://www.dropbox.com/scl/fo/ljr76u5azr7u4zx7mk55p/AB1RtEs0KyWqL6rKsiJHRKk?rlkey=ogas3pbhuhhhy94kprbd2m553&st=oq35p7rx&dl=0).
+ find the path to the file named **PRISM_ppt_stable_4kmD2_20120801_bil.bil**, which is PRISM precipitation data observed on 08-01-2012.
:::

<br>

It has a different file extension of **.bil**. Well, it does not matter. Just use `terra::rast()` with path to the file inside it just like you did with the GeoTiff files.

<br>

This is what the code looks like for me:

```{r, eval = FALSE}
prism_precip <- terra::rast("data-files-for-participants/PRISM_ppt_stable_4kmD2_20120801_bil/PRISM_ppt_stable_4kmD2_20120801_bil.bil")
```

#### Let's try (.nc)

Many weather data comes in the netCDF format with the file extension of **.nc**. (e.g., [gridMET](https://www.climatologylab.org/))

:::{.callout-note title="Instruction"}
+ download the file named **gm_precip_2018.nc** from this [link](https://www.dropbox.com/scl/fo/ljr76u5azr7u4zx7mk55p/AB1RtEs0KyWqL6rKsiJHRKk?rlkey=ogas3pbhuhhhy94kprbd2m553&st=oq35p7rx&dl=0).
+ find the path to the file
:::

<br>

Let's read the data now. This is what the code looks like for me:

```{r, eval = FALSE}
gm_precip <- terra::rast("data-files-for-participants/gm_precip_2018.nc")
```


:::
<!-- end of panel -->

### Write with `terra`

<details class="transcript">
<summary>Transcript</summary>

writeRaster going out, rast coming in, and the file extension decides the format, so ending a name in tif gets you a GeoTIFF with no further instruction. Two things to remember. You need overwrite equals TRUE to replace an existing file, and without it you get an error rather than a silent replacement, which is the right default. And netCDF is the exception to the extension rule: writeRaster will work but nudges you toward writeCDF, which handles netCDF's extra structure properly. Do the round trip in these tabs and confirm what you read back matches what you wrote.

</details>

::: {.panel-tabset}

#### How

You can use `terra::writeRaster()` to write raster data to a data file.

<br>

**Syntax**

```{r, eval = FALSE}
terra::writeRaster(SpatRaster, path to the file)
```

<br>

**Example (does not run)**

```{r, eval = FALSE}
writeRaster(reflec_blue, "./data/reflec_blue.tif", overwrite = TRUE) 
```

+ This code saves `reflec_blue` (a `SpatRaster` object) as a GeoTiff file. 

+ `writeRaster()` infers the correct format from the extension of the file name, which is **.tif** here. 

+ The `overwrite = TRUE` option is necessary if a file with the same name already exists and you are overwriting it. 

<br>

:::{.callout-note title="Note"}
No distinction is necessary for single-layer and multi-layer `SpatRaster` objects.
:::

#### Let's try (.tif)

**single-layer**

Write `reflec_blue` on your computer. Mine looks like this:

```{r, eval = FALSE}
terra::writeRaster(reflec_blue, "data/reflec_blue.tif")
```

<br>

Then read it back:

```{r, eval = FALSE}
reflec_blue_re_read <- terra::rast("data/reflec_blue.tif")
```

<br>

Confirm that  `reflec_blue_re_read` is the same as `reflec_blue`.

<br>

---

**multi-layer**

Write `reflec_all` on your computer. Mine looks like this:

```{r, eval = FALSE}
terra::writeRaster(reflec_all, "data/reflec_all.tif")
```

<br>

Then read it back:

```{r, eval = FALSE}
reflec_all_re_read <- terra::rast("data/reflec_all.tif")
```

<br>

Confirm that  `reflec_all_re_read` is the same as `reflec_all`.

#### Let's try (.nc)

While you can use `terra::writeRaster()` to write to a netCDF file, you will get a note saying that you should consider using `terra::writeCDF()`. 

Write `reflec_all` to a netCDF file. Mine looks like this.

```{r, eval = FALSE}
terra::writeCDF(reflec_all, "data/reflec_all.nc")
```

:::

:::

## Raster Data Operations {#sec-operations}

::: {.panel-tabset}

### Arithmetic operations

<details class="transcript">
<summary>Transcript</summary>

Raster arithmetic works cell by cell, exactly like vector arithmetic. Add two layers and each cell is added to the cell in the same position. Apply log and every cell is logged. The requirement, again, is matching extent and resolution, since otherwise cell-to-cell has no meaning. This is how vegetation indices get computed, and the second exercise is the real example: NDVI is near-infrared minus red over near-infrared plus red, which is three raster operations and one of the most used calculations in remote sensing. Note the result inherits its name from the first layer in the expression, which is worth knowing when you print it.

</details>

::: {.panel-tabset}

#### Introduction

+ You can do basic arithmetic operations (addition, subtraction, division, etc) using raster layers as long as they share the same spatial extent and resolution
  
+ You can also apply a function like `log()` to transform the value of the cells

+ Raster arithmetic operations are done element-by-element (cell-by-cell) just like vector arithmetic operations.

For example, when two `RasterLayer`s are added, then the two values associated with the same cell are added and the resulting value becomes the new value for the cell in the newly created `RasterLayer`.


#### Example

```{webr-r}
(
  reflec_b_plus_g <- reflec_blue + log(reflec_green)
)
```

<br>

Did it work? The first number below should be the sum of the other two.

```{webr-r}
c(reflec_b_plus_g[10100], reflec_blue[10100], log(reflec_green[10100])) %>% unlist()
```

<br>

Yes, looks like it did. Look at different cells yourself. Note that the name of the attribute in `reflec_b_plus_g` inherited the name of the attribute in `reflec_blue` (first `SpatRaster` in the addition above) — which is why two of the three numbers above are labelled `blue`.

#### Exercise 1

Multiply `reflec_blue` with `reflec_red` and add square root of `reflec_green`:

```{webr-r}

```

<br>

Look at several cells to confirm that multiplication was successful.

```{webr-r}

``` 


<br>

**Answer:**

```{r, eval = FALSE}
#| code-fold: true
reflec_temp <- reflec_blue * reflec_red + sqrt(reflec_green)

#--- confirm if the calculations were done right ---#
reflec_temp[10100] == reflec_blue[10100] * reflec_red[10100] + sqrt(reflec_green[10100])
```

#### Exercise 2

Using `NIR` and `RED`, calculate NDVI and name it `NDVI`.

$NDVI = (NIR-RED)/(NIR+RED)$

```{webr-r}
NIR
RED
```

<br>

Then, plot it to see what it looks like.

```{webr-r}

```

<br>

**Answer:**

```{r, eval = FALSE}
#| code-fold: true
 
NDVI <- (NIR - RED) / (NIR + RED)

#--- see what it looks like ---#
plot(NDVI)
```

:::

### Aggregate

<details class="transcript">
<summary>Transcript</summary>

Aggregating makes cells bigger and the grid coarser. The reason is usually practical: drone imagery at thirty centimetre resolution is enormous, and plotting or analysing it at full resolution wastes a lot of time for detail you cannot see. The factor argument says how many cells get combined in each direction, so five turns each five-by-five block into one cell and cuts the count twenty-five-fold. Read the note about fun, because the default is the mean. That is right for reflectance or temperature and wrong for anything categorical, where averaging land cover codes gives you a number that means nothing.

</details>

::: {.panel-tabset}

#### Motivation

Sometimes, you want to make your raster data have a lower resolution. For example, satellite image is often very fine with spatial resolution of say 30cm. When trying to create a map using the data, it takes a long time for R to render a plot and also the size of the figure can be very large.

```{webr-r}
reflec_blue
```

<br>


#### How and Example

**Syntax**

```{r, eval = FALSE}
aggregate(SpatRaster, fact, fun)
```

+ `fact`: aggregation factor — how many cells in each direction get combined into
one. `fact = 5` turns each 5-by-5 block of cells into a single cell, so the
result has 25 times fewer cells. Give one number to use it both horizontally and
vertically, or `c(row_factor, column_factor)`, equivalently
`c(y_factor, x_factor)`, to differ.
+ `fun`: how the values in each block are combined. **Defaults to `"mean"`.**
Others include `"sum"`, `"min"`, `"max"`, `"median"` and `"modal"`.

:::{.callout-note title="The default averages"}
Leaving `fun` out averages, which is what you want for a continuous variable like
reflectance or temperature. It is wrong for a categorical one: averaging land
cover codes where 1 = corn and 2 = soybean gives you 1.4. Use `fun = "modal"`
there, which takes the most common value in the block.
:::

**Example**

```{webr-r}
#| autorun: true
(
reflec_blue_agg_5 <- aggregate(reflec_blue, fact = 5)
)
```

<br>

Let's compare before and after. After aggregating by a factor of 5, the map is visibly coarser. Maybe this was too much. 


```{webr-r}
#| autorun: true
#| code-track: 0.3
plot(reflec_blue)
```

<br>

```{webr-r}
#| autorun: true
#| code-track: 0.3
plot(reflec_blue_agg_5)
```

#### Exercise 

Aggregate `NIR` by factor of 2. 

```{webr-r}
NIR
```

<br>

Then, check if its plot is acceptable or not.

```{webr-r}
```

<br>

**Answer:**

```{r, eval = FALSE}
#| code-fold: true
NIR_ag_2 <- aggregate(NIR, fact = 2)

plot(NIR_ag_2)
```
:::

### Resample

<details class="transcript">
<summary>Transcript</summary>

Resampling is the fix for two rasters that do not line up, and you will need it more often than you would like, since weather data, soil data and satellite data all come on their own grids. It transfers the values of one raster onto the cell geometry of another so the two become comparable. Look at the picture in the second tab, with red and blue grids overlapping, to see the problem before the syntax. Then read the important note: the default is bilinear for noncategorical data and nearest neighbour when the first layer is categorical. Specifying `method` explicitly is still clearer.

</details>

::: {.panel-tabset}

#### Motivation

+ You have two raster layers that differ in their dimensions and resolution. 

+ You want to assign a value from one layer to each of the cells in the other layer so that you have consistent observation units for the variables from the two layers.

::: {.columns}

::: {.column width="50%"}

Fake precipitation data:

```{webr-r}
#| out-width: 100%
#| autorun: true 
precip <- rast(xmin=1, xmax=11, ymin=1, ymax=11, nrows=10, ncols = 10, vals = rnorm(100), crs = "EPSG:4326")

plot(precip)
```
:::

::: {.column width="50%"}

Fake soil data:

```{webr-r}
#| autorun: true
#| out-width: 100%
soil <- rast(xmin=0, xmax=12, ymin=0, ymax=12, nrows=26, ncols = 26, vals = 100 * runif(26 * 26), crs = "EPSG:4326")

plot(soil)
```
:::

:::

#### What does it do really?

+ Map below shows grids from `precip` (red border) and from `soil` (blue border). 

+ As you can see, the grids from the two layers are not nicely aligned. 

+ See the top left grid with green fill color from the `precip` layer. Resampling will assign a single value to that grid based on the values of the nearby grids from the `soil` layer.

+ Even though the name "resample" sounds like it is a random process, there is no randomness.


```{webr-r}
#| context: output 

precip_sf <- 
  st_as_stars(precip) %>%
  st_as_sf()

soil_sf <- 
  st_as_stars(soil) %>%
  st_as_sf()

ggplot() +
  geom_sf(data = precip_sf, fill = NA, col = "red", linewidth = 0.8) +
  geom_sf(data = precip_sf[1, ], fill = "green", col = "red", alpha = 0.4) +
  geom_sf(data = soil_sf, fill = NA, col = "blue") +
  theme_void()
```

#### How

**Syntax**
```{r, eval = FALSE}
resample(sr_1, sr_2, method)
```

<br>

+ `sr_1`:	`SpatRaster` to be resampled
+ `sr_2`:	`SpatRaster` with the geometry that `sr_1` should be resampled to
+ `method`: method of assigning values
  + "bilinear": bi-linear interpolation (**the default** for a continuous layer)
  + "near": nearest neighbor (the default only when the first layer is categorical)
  + "cubicspline": cubic-spline interpolation
  + others (run `?terra::resample` to see all the `method` options)

:::{.callout-important title="`near` is not the default"}
It is tempting to assume nearest neighbor is the default, since it is the
simplest rule. It is not. `terra` picks `bilinear` unless the first layer of
`sr_1` is categorical, so **omitting `method` interpolates your values**.

That matters when the values are not interpolatable. Resampling a land-cover
raster where 1 = corn and 2 = soybean with `bilinear` will happily hand you 1.4,
which is neither. Pass `method = "near"` explicitly for any categorical layer.
:::

#### Example

Since we are resampling values from the `soil` layer onto the grids of `precip`, `soil` comes first and `precip` second:

```{webr-r}
#| layout: stacked
soil_to_precip_grids <- resample(soil, precip, method = "cubicspline")

plot(soil_to_precip_grids)
```

<br>

:::{.callout-note title="Try yourself"}
+ Change the `method` and see how the resampling results change.
+ Many of the spatial datasets tend to exhibit a positive spatial correlation and the resampling outcomes are not as sensitive to `method` as you see here where cell values are completely independent.
:::

:::

### Merge

<details class="transcript">
<summary>Transcript</summary>

Merging is for rasters covering different areas that you want as one, which is the situation whenever your study region spans two data tiles. It is not the same as stacking, so keep that straight: stacking puts several variables over one area into layers, merging puts one variable over several areas into a single wider grid. Because a raster must be a rectangle, the result is the rectangle enclosing both inputs, and anything neither one covered comes back as NA. The caveat tab is the part to remember: where they overlap, the first object wins. The experiment there makes that visible by flattening one input to a constant.

</details>

::: {.panel-tabset}

#### Motivation

Sometimes, you have two or more raster layers that have different spatial coverages. In such a case, you might want to merge them into a single raster layer.

For demonstration purpose, we will use two `SpatRaster` objects: `prism_saunders` and `prism_douglas`. They are PRISM maximum temperature observed on 08/01/2012 in the Saunders and Douglas counties in Nebraska, which are adjacent to each other.

```{webr-r}
#| autorun: true
#| code-track: 0.4 
ggplot() +
  geom_spatraster(data = prism_saunders) +
  geom_spatraster(data = prism_douglas) +
  scale_fill_viridis_c()
```

:::{.callout-note}
Note that this is different from combining multiple single-layer raster data of the same spatial extent and resolution into multi-layer raster data.
:::

#### How

You can use the `terra::merge()` function to merge two raster datasets into one.

```{webr-r}
prism_merged <- terra::merge(prism_saunders, prism_douglas)
```

<br>

You can check the result of the merging below:

```{webr-r}
#| code-track: 0.4 
ggplot() +
  geom_spatraster(data = prism_merged) +
  scale_fill_viridis_c()
```

<br>

:::{.callout-note}
Remember that raster object has to be perfectly rectangular. The result of merging will construct a rectangle that encompasses both `prism_saunders` and `prism_douglas`. All the cells that are not covered by `prism_saunders` and `prism_douglas` will be assigned NA.
:::


#### Caveat

When merging two `SpatRaster` objects and when they have spatial overlaps, the value of the first `SpatRaster` object will be respected.

Let's run a little experiment. We will assign a high value to all the cells in `prism_saunders`. This will make this phenomenon easy to detect.

```{webr-r}
values(prism_saunders) <- 120
```

:::{.callout-warning title="This overwrites `prism_saunders` for the rest of the session"}
The line above does not make a copy — it replaces every temperature in
`prism_saunders` with 120. Every cell on this deck shares one R session, so if you
go back to the **How** tab afterwards and re-run it, you will get the flattened
version rather than the real temperatures.

Re-run the setup by reloading the page if you want the original data back.
:::

<br> 

**`prism_saunders` first**

```{webr-r}
merge(prism_saunders, prism_douglas) %>% plot()
```

<br>

**`prism_douglas` first**

```{webr-r}
merge(prism_douglas, prism_saunders) %>% plot()
```

:::
<!--end of panel-->

:::
