---
title: "09-4-1: R as GIS: Interaction of Vector Datasets I"
abstract: "Chapter 9: Spatial Data"
format: 
  revealjs: 
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    include-in-header:
      - ../transcript-support.html
      - ../webr-layout.html
      - ../katex-amd-fix.html
    footer: "[back to the lecture slides on the course website](https://tmieno2.github.io/Data-Science-with-R-Quarto/lectures/)"
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    fig-dpi: 400
    chalkboard: true
webr:
  packages: ['ggplot2', 'dplyr', 'sf'] # Install R packages on document open
# autoload-packages: false       # Disable automatic loading of packages
# show-startup-message: false    # Disable displaying status of webR initialization
  cell-options:
    editor-font-scale: 0.7
    dpi: 216
    fig-width: 7
    fig-height: 4.5
    out-width: 100%
filters:
  - webr
---

```{r}
#| include: false

source(here::here("lectures/_lecture-theme.R"))
```


```{r, include = FALSE}
library(sf)
library(dplyr)
library(data.table)
```

## Before you start

<details class="transcript">
<summary>Transcript</summary>

Up to now every deck has handled one spatial object at a time. This one is about what happens when two of them meet, which is where spatial analysis actually starts. The questions are ordinary: which wells fall inside this aquifer, which railroads cross this county, which weather station is nearest to this field. Three tools answer nearly all of them. Topological relations tell you what touches what. Spatial subsetting keeps the features that satisfy a relation. Cropping cuts geometries down to a rectangle. Everything in the next deck, on spatial joins, is built on the first of those, so it is worth getting comfortable here.

</details>

<br>

### Learning objectives

The objective of this chapter is to learn spatial operations that involve two `sf` objects. Specifically,

+ understand topological relations
+ subsetting an `sf` object based on another `sf` object

<br>

::: {.columns}

::: {.column width="50%"}
### Table of contents

1. [Topological Relations](#sec-topological)
2. [Spatial Subsetting](#sec-subsetting)
3. [Spatial Cropping](#sec-cropping)
:::
<!--end of the 1st column-->
::: {.column width="50%"}
### Pre-requisite (Links)

+ [`ggplot2` primer](https://tmieno2.github.io/R-as-GIS-workshop/LA_1_ggplot2_primer.html)
+ [`dplyr` primer](https://tmieno2.github.io/R-as-GIS-workshop/LA_2_dplyr_primer.html)
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

```{webr-r}
#| context: setup
nc <-
  st_read(system.file("shape/nc.shp", package = "sf")) %>%
  dplyr::select(AREA, NAME, FIPS)

#--- install and library the data package ---#
#--- Both repos are required. r.spatial.workshop.datasets declares         ---#
#--- Imports: sf, and webr::install() resolves dependencies ONLY in the     ---#
#--- repos named here. With the custom repo alone, sf cannot be found and   ---#
#--- the install never completes, so nothing downstream ever runs.          ---#
webr::install(
  "r.spatial.workshop.datasets",
  repos = c(
    "https://tmieno2.github.io/rwasm-package/repo",
    "https://repo.r-wasm.org"
  )
)

library(r.spatial.workshop.datasets)

#--- load datasets ---#
data(wells_ne)
data(points)
data(polygons)
data(lines)

# Mirror of lecture_theme() in lectures/_lecture-theme.R. WebR runs in the
# browser against a virtual filesystem that does not hold the repo, so it
# cannot source that file and has to carry a copy. Edit the shared file first,
# then run `Rscript lectures/check-lecture-theme.R`, which fails if this copy
# has drifted and prints the block to paste.
lecture_theme <- function(base_size = 16) {
  ggplot2::theme_bw(base_size = base_size) +
    ggplot2::theme(
      axis.text = ggplot2::element_text(size = base_size * 0.9),
      axis.title = ggplot2::element_text(size = base_size),
      legend.text = ggplot2::element_text(size = base_size * 0.9),
      legend.title = ggplot2::element_text(size = base_size * 0.9),
      plot.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.key = ggplot2::element_rect(fill = "transparent", colour = NA),
      # theme_bw's grey92 grid was chosen against a white panel and all but
      # disappears on the warm paper once the panel stops being opaque.
      panel.grid = ggplot2::element_line(colour = "grey87")
    )
}

# WebR draws its canvas at 72 dpi whatever `dpi` says, so on the 216-dpi canvas
# every size ggplot measures in points or millimetres comes out a third of what
# was asked for. Scaling ggplot2's own mm-to-point constants and the theme by
# the same factor puts WebR figures at the same type size as the knitr ones.
# Keep this in step with `dpi` in the header: change one, change the other.
webr_scale <- 216 / 72
assignInNamespace(".pt", ggplot2::.pt * webr_scale, "ggplot2")
assignInNamespace(".stroke", ggplot2::.stroke * webr_scale, "ggplot2")
theme_lecture <- lecture_theme(18 * webr_scale) +
  # base_size also sets the theme's line widths, which are in millimetres and
  # already pass through the scaled .pt; pin them to what 18pt gives, or they
  # come out nine times too thick.
  ggplot2::theme(
    line = ggplot2::element_line(linewidth = 18 / 22),
    rect = ggplot2::element_rect(linewidth = 18 / 22)
  )
ggplot2::theme_set(theme_lecture)
```


## Tips to make the most of the lecture notes

<details class="transcript">
<summary>Transcript</summary>

The usual arrangement. One thing specific to this deck: it leans on small made-up datasets, three points, two lines and three polygons, drawn on a single picture you will see over and over. That is deliberate. With real data you cannot check the answer by eye, and every function here returns a list of numbers that means nothing unless you can verify it against a picture. So when a result appears, go back to the figure and count. Later, once the pattern is clear, we switch to real Nebraska data where checking by eye is no longer possible.

</details>

<br>

### Interactive navigation tools

+ Click on the three horizontally stacked lines at the bottom left corner of the slide, then you will see table of contents, and you can jump to the section you want

+ Hit letter "o" on your keyboard and you will have a panel view of all the slides

<br>

### Running and writing codes

+ The box area with a hint of blue as the background color is where you can write code (hereafter referred to as the "code area").
+ Hit the "Run Code" button to execute all the code inside the code area.
+ You can evaluate (run) code selectively by highlighting the parts you want to run and hitting Command + Enter for Mac (Ctrl + Enter for Windows).
+ If you want to run the codes on your computer, you can first click on the icon with two sheets of paper stacked on top of each other (top right corner of the code chunk), which copies the code in the code area. You can then paste it onto your computer.
+ You can click on the reload button (top right corner of the code chunk, left to the copy button) to revert back to the original code.



## Topological Relations {#sec-topological}

::: {.panel-tabset}

### Introduction

<details class="transcript">
<summary>Transcript</summary>

Topological relations describe how two geometries sit relative to one another: touching, containing, crossing, overlapping, being disjoint. The sf package implements the whole standard vocabulary, and you can see the full list by running the help query on this slide. In practice you will use one of them almost all the time, st_intersects, which asks the loosest possible question: do these two share any point at all. We spend most of this section on it and then look briefly at two distance-based cousins. Do not memorise the list; know that it exists and look it up when intersects is too blunt.

</details>


**Definition**

Topological relations refer to the way multiple spatial objects are spatially related to one another. 

<br>

**Goals**

You can identify various types of spatial relations using the `sf` package 

+ Our main focus is on the intersections of spatial objects, which can be found using `st_intersects()`.

+ We also briefly cover `st_is_within_distance()` and `st_nearest_feature()`

You can run `?geos_binary_pred` to find all the topological relations you can use:

```{webr-r}
?geos_binary_pred
```

### `st_intersects()`

<details class="transcript">
<summary>Transcript</summary>

This is the workhorse, and the tabs walk through all three combinations: points against polygons, lines against polygons, polygons against polygons. Start with the Visualized Data tab and keep that picture in mind, because every result that follows can be checked against it by eye. The output takes a moment to get used to. It is a list, one element per feature in the first object, holding the indices of everything in the second object that it intersects. So the first element tells you what point one touches, and so on. Add sparse equals FALSE if you would rather have a matrix of TRUE and FALSE.

</details>

::: {.panel-tabset}

#### Data Preparation

We use three `sf` objects: `points`, `lines`, and `polygons`. Inspect each of them.

<br>

```{webr-r}
points
lines
polygons
```

<br>

If you are interested in codes to create them see below.

```{r}
#| code-fold: true
#| eval: false
#--- create points ---#
point_1 <- sf::st_point(c(2, 2))
point_2 <- sf::st_point(c(1, 1))
point_3 <- sf::st_point(c(1, 3))

#--- combine the points to make a single  sf of points ---#
points <- list(point_1, point_2, point_3) %>% 
  sf::st_sfc() %>% 
  sf::st_as_sf() %>% 
  mutate(point_name = c("point 1", "point 2", "point 3"))

#--- create lines ---#
line_1 <- sf::st_linestring(rbind(c(0, 0), c(2.5, 0.5)))
line_2 <- sf::st_linestring(rbind(c(1.5, 0.5), c(2.5, 2)))

#--- combine the points to make a single  sf of points ---#
lines <- list(line_1, line_2) %>% 
  sf::st_sfc() %>% 
  sf::st_as_sf() %>% 
  mutate(line_name = c("line 1", "line 2"))

#--- create polygons ---#
polygon_1 <- sf::st_polygon(list(
  rbind(c(0, 0), c(2, 0), c(2, 2), c(0, 2), c(0, 0)) 
))

polygon_2 <- sf::st_polygon(list(
  rbind(c(0.5, 1.5), c(0.5, 3.5), c(2.5, 3.5), c(2.5, 1.5), c(0.5, 1.5)) 
))

polygon_3 <- sf::st_polygon(list(
  rbind(c(0.5, 2.5), c(0.5, 3.2), c(2.3, 3.2), c(2, 2), c(0.5, 2.5)) 
))

#--- combine the polygons to make an sf of polygons ---#
polygons <- list(polygon_1, polygon_2, polygon_3) %>% 
  sf::st_sfc() %>% 
  sf::st_as_sf() %>% 
  mutate(polygon_name = c("polygon 1", "polygon 2", "polygon 3"))
```

#### Visualized Data

```{webr-r}
#| autorun: true
(
g_all <- 
  ggplot() +
  geom_sf(data = polygons, aes(fill = polygon_name), alpha = 0.3) +
  scale_fill_discrete(name = "Polygons") +
  geom_sf(data = lines, aes(color = line_name)) +
  scale_color_discrete(name = "Lines") + 
  geom_sf(data = points, aes(shape = point_name), size = 3) +
  scale_shape_discrete(name = "Points") +
  theme_void()
)

```

#### points and polygons

`st_intersects()` checks which of the `sfg`s in one `sf` geographically intersect with which of the `sfg`s in another `sf`.

:::: {.columns}

::: {.column width="40%"}
```{webr-r}
#| context: output 
#| out-width: 120% 
g_all 
```
:::

::: {.column width="10%"}
:::

::: {.column width="50%"}

<br>

```{webr-r}
st_intersects(points, polygons)
```

<br>

+ The output is a list of which polygon(s) each of the points intersect with. 

+ The numbers 1, 2, and 3 in the first row mean that 1st (polygon 1), 2nd (polygon 2), and 3rd (polygon 3) objects of the `polygons` intersect with the first point (point 1) of the `points` object. 

+ The fact that point 1 is considered to be intersecting with polygon 2 means that the area inside the border is considered a part of the polygon (of course).

<br>

If you would like the results of `st_intersects()` in a matrix form with boolean values filling the matrix, you can add `sparse = FALSE` option. 

```{webr-r}
st_intersects(points, polygons, sparse = FALSE)
```
:::

::::

#### lines and polygons

:::: {.columns}

::: {.column width="40%"}
```{webr-r}
#| context: output 
#| out-width: 120% 
g_all 
```
:::

::: {.column width="10%"}
:::

::: {.column width="50%"}

<br>

```{webr-r}
st_intersects(lines, polygons)
```

The output is a list of which polygon(s) each of the lines intersect with.
:::

::::

#### polygons and polygons

:::: {.columns}

::: {.column width="40%"}
```{webr-r}
#| context: output 
#| out-width: 120% 
g_all 
```
:::

::: {.column width="10%"}
:::

::: {.column width="50%"}

<br>

```{webr-r}
st_intersects(polygons, polygons)
```

For polygons vs polygons interaction, `st_intersects()` identifies any polygons that either touch (even at a single point, as polygons 1 and 3 do) or share some area.
:::
::::
:::

### `st_is_within_distance()`

<details class="transcript">
<summary>Transcript</summary>

The first of the two distance-based relations. Rather than asking whether two things touch, it asks whether anything in the second object lies within a given distance of each feature in the first. Note that the distance argument is required and carries no default. A numeric distance uses metres for geographic coordinates and coordinate units for projected or CRS-free data. That catches people constantly. Work through the tabs in order: the expected outcome tab draws buffers of the right radius around each red point, so you can read the answer off the picture before you run the function.

</details>

::: {.panel-tabset}

#### Syntax

`sf::st_is_within_distance()` function identifies whether any of `sfg`s in `sf_2` is within the specified distance from each of the `sfg`s in `sf_1`.

```{r eval = F}
st_is_within_distance(sf_1, sf_2, dist)
```

+ `dist`: the distance threshold. It is **required**; there is no default. A numeric
distance uses metres for geographic coordinates and coordinate units for
projected or CRS-free data.


#### Data Preparation

Create two sets of points and then inspect each of them.

```{webr-r}
#| autorun: true 
set.seed(884738)

points_set_1 <- lapply(1:5, function(x) st_point(runif(2))) %>% 
  st_sfc() %>% st_as_sf() %>% 
  mutate(id = 1:nrow(.))

points_set_2 <- lapply(1:5, function(x) st_point(runif(2))) %>% 
  st_sfc() %>% st_as_sf() %>% 
  mutate(id = 1:nrow(.))
```

<br>

Here is the visualization of the two sets of points we just created.

```{webr-r}
#| code-track: 0.4
ggplot() +
  geom_sf_text(
    data = points_set_1,
    aes(label = id),
    color = "red"
  ) +
  geom_sf_text(
    data = points_set_2,
    aes(label = id),
    color = "blue"
  ) 
```

#### Expected Outcomes

We want to know which of the blue points (`points_set_2`) are located within 0.2 from each of the red points (`points_set_1`). 

The following figure gives us the answer visually.

```{webr-r}
#| autorun: true
#| code-track: 0.4
#--- 0.2 buffers around points in points_set_1 ---#
buffer_1 <- st_buffer(points_set_1, dist = 0.2)

ggplot() +
  geom_sf(
    data = buffer_1, 
    color = "red", 
    fill = NA
  ) +
  geom_sf_text(
    data = points_set_1, 
    aes(label = id), 
    color = "red"
  ) +
  geom_sf_text(
    data = points_set_2, 
    aes(label = id), 
    color = "blue"
  ) 
```

#### Apply the function

```{webr-r}
#| out-width: 100%
#| code-track: 0.4 
ggplot() +
  geom_sf(
    data = st_buffer(points_set_1, dist = 0.2), color = "red", 
    fill = NA
  ) +
  geom_sf_text(
    data = points_set_1, 
    aes(label = id), 
    color = "red"
  ) +
  geom_sf_text(
    data = points_set_2, 
    aes(label = id), 
    color = "blue"
  )
```  

<br>

```{webr-r}
#| code-track: 0.4
st_is_within_distance(
  points_set_1,
  points_set_2,
  dist = 0.2
)
```

<br>

Confirm that your visual inspection results are consistent with the outcome of the `st_is_within_distance()` code above.

:::

### `st_nearest_feature()`  

<details class="transcript">
<summary>Transcript</summary>

The second distance relation, and a simpler one. For each feature in the first object it returns the single closest feature in the second, always exactly one, no threshold involved. That difference matters: within-distance can return several matches or none at all, while nearest-feature always returns precisely one answer per feature. Which you want depends on the question. If you are attaching the nearest weather station to each field, this is the function. If you want every station within twenty kilometres, it is the previous one. Check the result against the same picture of red and blue points.

</details>

::: {.panel-tabset}

#### Syntax

`sf::st_nearest_feature()` identifies which `sfg` in `sf_2` is closest to each of the `sfg`s in `sf_1`.

```{r eval = F}
st_nearest_feature(sf_1, sf_2)
```

#### Example

```{webr-r}
#| code-track: 0.4
ggplot() +
  geom_sf(
    data = buffer_1, 
    color = "red", 
    fill = NA
  ) +
  geom_sf_text(
    data = points_set_1, 
    aes(label = id), 
    color = "red"
  ) +
  geom_sf_text(
    data = points_set_2, 
    aes(label = id), 
    color = "blue"
  )
```

<br>

```{webr-r}
st_nearest_feature(points_set_1, points_set_2)
```

Confirm that your visual inspection results are consistent with the outcome of the `st_nearest_feature()` code above.

:::

### Exercises

<details class="transcript">
<summary>Transcript</summary>

Now the same function on real data, where you cannot check the answer by eye. The mower sensor dataset is a dense cloud of GPS readings from a fairway, and the fairway grid is a set of polygons over the same area. Notice the first cell builds an sf from plain longitude and latitude columns with st_as_sf, taking its CRS from the grid so the two objects match — a step you will repeat often with real data. Then run st_intersects and accept that the output is a long unhelpful list. Filtering spatially, which is what you actually want, is the next section.

</details>

::: {.panel-tabset}

#### Data Preparation

Run the following codes to get `mower_sensor` and `fairway_grid` data.

```{webr-r}
#--- mower sensor ---#
data(mower_sensor, package = "r.spatial.workshop.datasets")

mower_sensor_sf <- st_as_sf(mower_sensor, coords = c("LNG", "LAT"), crs = st_crs(fairway_grid))

#--- fairway grid ---#
data(fairway_grid, package = "r.spatial.workshop.datasets")
```

<br>

Visualize the datasets.

```{webr-r}
#| code-track: 0.4
ggplot() +
  geom_sf(data = fairway_grid, col = "red") +
  geom_sf(data = mower_sensor_sf)
```

#### `st_intersects()`

Use `sf::st_intersects()` to find out which of the points in `mower_sensor_sf` are inside of any of the polygons in `fairway_grid`.

```{webr-r}
#--- type your answer here ---#

```

<br>

Since there are so many points in `mower_sensor_sf`, you won't really see which ones are inside of any of the polygons in `fairway_grid`.

That is okay for now. We will later learn how to filter `sf` objects spatially. 

<br>

**Answer codes**
```{r}
#| code-fold: true 
#| eval: false 
st_intersects(mower_sensor_sf, fairway_grid)
```


:::
:::
<!--end of panel-->



## Spatial Subsetting {#sec-subsetting}

::: {.panel-tabset}

### Introduction

<details class="transcript">
<summary>Transcript</summary>

Subsetting is the natural next step. Topological relations told you which features are related; subsetting uses that answer to keep only the features you want. The phrasing to hold on to is that you narrow one spatial object down using another. Everything in this section uses one compact piece of syntax that looks like ordinary data frame indexing, and the three tabs after this one apply it to each geometry type in turn: polygons by polygons, points by polygons, and lines by polygons. The mechanics are identical in all three, which is the point.

</details>

Spatial subsetting refers to operations that narrow down the geographic scope of a spatial object based on another spatial object. 


### Data preparation

<details class="transcript">
<summary>Transcript</summary>

Four real Nebraska datasets for this section. County boundaries as polygons, irrigation wells as points, railroads as lines, and the High Plains aquifer boundary, which is the thing we will keep subsetting against. Print each one and look at the columns before going further, because you will be referring to them by name shortly. Note that the wells arrive as a plain table of longitude and latitude and get converted into an sf object with st_as_sf, using the North American Datum as the CRS. That conversion is one of the most common first steps in any real spatial workflow.

</details>

Here are the datasets we will use here. Inspect them to familiarize yourself with the datasets.

```{webr-r}
#--- Nebraska county boundary ---# 
ne_counties

#--- Irrigation wells in Nebraska ---#
wells_ne

#--- Railroads in Nebraska ---#  
railroads_ne

#--- High-Plains aquifer boundary ---#
hp_boundary
```


```{webr-r}
#| context: setup
#--- Nebraska county boundary ---# 
data(ne_counties, package = "r.spatial.workshop.datasets")

#--- Irrigation wells in Nebraska ---#
data(wells_ne)
wells_ne <-
  sf::st_as_sf(
    wells_ne,
    coords = c(
      "longdd",
      "latdd"
    ),
    crs = 4269
  )

#--- Railroads in Nebraska ---#  
data(railroads_ne, package = "r.spatial.workshop.datasets")

#--- High-Plains aquifer boundary ---#
data(hp_boundary, package = "r.spatial.workshop.datasets")
```

### polygons vs polygons

<details class="transcript">
<summary>Transcript</summary>

Here is the syntax, and it is deliberately familiar: square brackets, with another sf object where you would normally put row numbers. That is the whole thing. What you get back are the counties that intersect the aquifer boundary, because intersects is the default relation. Read the note about how loose that is: a county touching the aquifer by a sliver survives. The last two tabs are the more interesting ones. Flagging keeps every row and adds a zero or one instead of dropping anything, which is usually what you want for regression. And the op argument lets you demand containment rather than mere contact.

</details>
 
::: {.panel-tabset}

#### Goal

Select only the counties that intersect with the HPA boundary.

```{webr-r}
ggplot() +
  geom_sf(data = ne_counties) +
  geom_sf(
    data = hp_boundary,
    fill = "blue",
    alpha = 0.4
  ) +
  theme_void()
```

#### How

When subsetting a data.frame by specifying the row numbers you would like to select, you can do 

```{r}
#| eval: false
#--- NOT RUN ---#
data.frame[vector of row numbers, ]
```
<br>

Spatial subsetting of sf objects works in a similar syntax:   

```{r}
#| eval: false
#--- NOT RUN ---#
sf_1[sf_2, ]
```

<br>

where you are subsetting `sf_1` based on `sf_2`. Instead of row numbers, you provide another `sf` object in place.

#### Demonstration

The following code spatially subsets Nebraska counties based on the HPA boundary. 

```{webr-r}
#| autorun: true
ne_counties_in_hpa <- ne_counties[hp_boundary, ]
```

<br>

::: {.columns}

::: {.column}
```{webr-r}
ggplot() +
  geom_sf(data = ne_counties_in_hpa) +
  geom_sf(data = hp_boundary, fill = "blue", alpha = 0.4) +
  theme_void()
```
:::

::: {.column}
+ You can see that only the counties that intersect with the HPA boundary remained. 

+ This is because when you use the above syntax of `sf_1[sf_2, ]`, the default underlying topological relation is `st_intersects()`. 

+ So, if an object in `sf_1` intersects with any of the objects in `sf_2` even slightly, then it will remain after subsetting.
:::

:::

#### Create a flag variable 

Sometimes, you just want to flag whether two spatial objects intersect or not, instead of dropping non-overlapping observations like we saw with `sf_1[sf_2, ]` syntax. In that case, you can get a list of the IDs and then assign 1 (or TRUE) if in the list, 0 (or FALSE) otherwise.

---
<br>

Get the list of countyfp (ID) of the intersected counties:

```{webr-r}
#| autorun: true
county_fp_intersected_list <- ne_counties_in_hpa$countyfp
```

<br>

Assign 1 or 0 to a new variable called `in_hpa` based on the list.

```{webr-r}
#| autorun: true
(
ne_counties <- dplyr::mutate(ne_counties, in_hpa = ifelse(countyfp %in% county_fp_intersected_list, 1, 0))
)
```

#### Other topological relations

You can specify the topological relation as in 

```{r}
#| eval: false
#--- NOT RUN ---#
sf_1[sf_2, , op = topological_relation_type] 
```

<br>

For example, if you only want counties that are completely within the HPA boundary, you can do the following:

```{webr-r}
#| autorun: true
ne_counties_completely_within_hpa <- ne_counties[hp_boundary, , op = st_within]
```

<br>

Check visually:

```{webr-r}	
#| layout: stacked
ggplot() +
  geom_sf(data = ne_counties_completely_within_hpa) +
  geom_sf(data = hp_boundary, fill = "blue" ,alpha = 0.4) +
  theme_void()
```
:::
<!--end of panel-->

### points vs polygons

<details class="transcript">
<summary>Transcript</summary>

Exactly the same syntax, now with points. Wells inside the aquifer boundary, in one line of brackets. There is a parenthetical worth noticing: for points, intersects and within give the same answer only when none of the points lies on the polygon boundary. A point on the boundary intersects the polygon but is not within it. The flagging tab repeats the pattern from before, pulling out the well identifiers that survived and marking every well as in or out. That flagged version, rather than the subset, is what you would carry into an analysis where you want to compare the two groups.

</details>
 
::: {.panel-tabset}

#### Goal

Select only the wells that intersect with the HPA boundary. This is equivalent to
selecting wells within it only when no wells lie on the polygon boundary.

```{webr-r}
ggplot() +
  geom_sf(
    data = hp_boundary,
    fill = "blue",
    alpha = 0.3,
    size = 0.05
  ) +
  geom_sf(
    data = wells_ne,
    size = 0.1
  ) +
  theme_void()  
```

#### Demonstration

We can select only the wells that reside within the HPA boundary using the same syntax as the polygon-polygon example.

```{webr-r}
#| autorun: true
wells_ne_in_hpa <- wells_ne[hp_boundary, ]
```
<br>

As you can see in the figure below, only the wells that intersect the HPA remain, because the default topological relation is `st_intersects()`. (Here you get the same result even if you use `op = st_within`.)

```{webr-r}
ggplot() +
  geom_sf(
    data = hp_boundary,
    fill = "blue",
    alpha = 0.3,
    size = 0.05
  ) +
  geom_sf(
    data = wells_ne_in_hpa,
    size = 0.1
  ) +
  theme_void()   
```

#### Flagging

Get the list of `wellid` (ID) of the intersected wells:

```{webr-r}
#| autorun: true
wellid_list <- wells_ne_in_hpa$wellid
```

<br>

Assign 1 or 0 to a new variable called `in_hpa` based on the list.

```{webr-r}
#| autorun: true
(
wells_ne <- dplyr::mutate(wells_ne, in_hpa = ifelse(wellid %in% wellid_list, 1, 0))
)
```

:::

### lines vs polygons

<details class="transcript">
<summary>Transcript</summary>

The third combination, and by now it should feel routine: railroads intersecting Lancaster county, using the same brackets as before. What is worth pausing on is what intersects actually means for a line. A railroad that merely clips the corner of the county is kept in full, not trimmed at the boundary. The whole line survives because subsetting selects features and never modifies their geometry. If what you want is the piece inside the county rather than the whole line, use st_intersection to clip it to the county boundary. The st_crop function, covered in the next section, clips only to a rectangular bounding box. The flagging tab closes the section by marking every railroad as in or out of Lancaster.

</details>
 
::: {.panel-tabset}

#### Goal

Select only railroads that intersect with the Lancaster county in Nebraska.

```{webr-r}
#| autorun: true
lancaster_county <- dplyr::filter(ne_counties, name == "Lancaster")
```

<br>

```{webr-r}
#| code-track: 0.3
ggplot() +
  geom_sf(
    data = ne_counties,
    color = "red",
    fill = NA,
    size = 0.1
  ) +
  geom_sf(
    data = lancaster_county,
    fill = "blue",
    alpha = 0.4
  ) +
  geom_sf(
    data = railroads_ne,
    size = 0.1
  ) +
  theme_void() 
```

#### Demonstration

Just like we did in the two previous examples:

```{webr-r}
#| autorun: true
railroads_lancaster <- railroads_ne[lancaster_county, ] 
```
<br>

Successful?

```{webr-r}
#| code-track: 0.3
ggplot() +
  geom_sf(
    data = ne_counties,
    color = "red",
    fill = NA,
    size = 0.1
  ) +
  geom_sf(
    data = lancaster_county,
    fill = "blue",
    alpha = 0.4
  ) +
  geom_sf(
    data = railroads_lancaster,
    size = 0.1
  ) +
  theme_void()
```

#### Flagging

Get the list of `LINEARID` (ID) of the intersected railroads:

```{webr-r}
#| autorun: true
railroads_id_list <- railroads_lancaster$LINEARID
```

<br>

Assign 1 or 0 to a new variable called `in_lancaster` based on the list.

```{webr-r}
#| autorun: true
(
railroads_ne <- dplyr::mutate(railroads_ne, in_lancaster = ifelse(LINEARID %in% railroads_id_list, 1, 0))
) 
```

:::
<!--end of panel-->

:::
<!--end of panel-->

## Spatial Cropping {#sec-cropping}

::: {.panel-tabset}

### Goal

<details class="transcript">
<summary>Transcript</summary>

Cropping is the operation that does change geometry. You give it a spatial object and a bounding box, and it cuts everything down to that rectangle, chopping features at the edge rather than keeping or dropping them whole. Notice the word rectangle: st_crop works to a bounding box, the minimum and maximum x and y that enclose an object, not to the shape of the object itself. That surprises people who expect to be cropping to the outline of the aquifer. The next tab builds the bounding box explicitly so you can see it.

</details>

+ We can use `st_crop()` to crop spatial objects to a spatial bounding box (extent) of a spatial object. 

+ The bounding box of an `sf` is a rectangle represented by the minimum and maximum of `x` and `y` that encompass/contain all the spatial objects in the `sf`. 

+ You can use `st_bbox()` to find the bounding box of an `sf` object.


### Bounding box

<details class="transcript">
<summary>Transcript</summary>

st_bbox gives you four numbers, the minimum and maximum of x and y. Check its class and you will find it is a bbox object, not an sf, and that matters because you cannot plot it or use it in most spatial operations directly. st_as_sfc converts it into a geometry you can actually draw, which is what the red rectangle on this slide is. Look at how much larger that rectangle is than the aquifer itself, because that gap is exactly what makes cropping different from subsetting, and it is the point of the last tab.

</details>

Let's get the bounding box of the High-Plains aquifer using `st_bbox()`.

```{webr-r}
#| autorun: true
#| code-track: 0.4
hp_bbox <- st_bbox(hp_boundary)
```
<br>

Check its class:
```{webr-r}
#| code-track: 0.4
class(hp_bbox)
```

<br>

You can convert a `bbox` to `sfc` by applying `st_as_sfc()` to the `bbox` object (you cannot use a `bbox` for mapping and other interactive spatial operations).

```{webr-r}
#| autorun: true
hp_bbox_sfc <- st_as_sfc(hp_bbox)
```
<br>

The bounding box looks like this (red rectangle):

```{webr-r}
#| code-track: 0.4
ggplot() +
  geom_sf(data = hp_boundary) +
  geom_sf(
    data = hp_bbox_sfc, 
    fill = NA, 
    color = "red"
  )
```


### Crop

<details class="transcript">
<summary>Transcript</summary>

The crop itself is one line. Note the remark that passing the aquifer object and passing its bounding box give the same result: st_crop takes the bounding box of whatever you hand it, so there is no need to convert first. The figure at the end is the one to study. The blue outline is the aquifer, the red rectangle its bounding box, and the orange shapes the cropped counties. Counties are sliced along the rectangle edge, mid-county, which no amount of subsetting would ever do. Some of those orange pieces are nowhere near the aquifer.

</details>

Now, let's crop Nebraska counties to the bounding box of the High-Plains aquifer boundary.

```{webr-r}
(
ne_counties_cropped_to_hp <- st_crop(ne_counties, hp_boundary)
)
``` 
<br>

Note that you do not need to do the following — it produces the same outcome, because `st_crop()` uses the bounding box of whatever you give it:

```{webr-r}
st_crop(ne_counties, hp_bbox_sfc)
```
<br>

```{webr-r}
#| code-track: 0.4
ggplot() +
  geom_sf(
    data = hp_boundary, 
    color = "blue", 
    linewidth = 1
  ) +
  geom_sf(
    data = ne_counties, 
    fill = NA
  ) +
  geom_sf(
    data = hp_bbox_sfc, 
    fill = NA, 
    color = "red"
  ) +
  geom_sf(
    data = ne_counties_cropped_to_hp, 
    fill = "orange", 
    alpha = 0.4
  )
```


### Cropping vs Subset

<details class="transcript">
<summary>Transcript</summary>

The summary slide, and worth a moment because the two operations are easy to confuse. Subsetting keeps whole features that satisfy a relation and never touches their shape. Cropping cuts every feature to a rectangle and will happily leave you with half a county. The figure overlays the green subset and the blue cropped version so the difference is visible directly. Which you want depends on the question: use subsetting when the unit of analysis is the county and you need it intact, and cropping when you want a study area of a particular extent regardless of what it cuts through.

</details>

Note that `st_crop()` will chop off the parts that are not intersecting.

```{webr-r}
#| code-track: 0.4
ggplot() +
  #--- subsetted ---#
  geom_sf(
    data = ne_counties_in_hpa, 
    fill = "green", 
    alpha = 0.3
  ) +
  #--- cropped ---#
  geom_sf(
    data = ne_counties_cropped_to_hp, 
    fill = "blue", 
    alpha = 0.3
  )
  # +
  # #--- HPA ---#
  # geom_sf(data = hp_boundary, fill = NA, color = "red")
```

:::


