---
title: "09-5: R as GIS: Interaction of Vector and Raster Datasets"
abstract: "Chapter 9: Spatial Data"
format: 
  revealjs: 
    html-math-method:
      method: katex
      url: "https://cdn.jsdelivr.net/npm/katex@0.18.4/dist/"
    include-in-header:
      - ../transcript-support.html
      - ../webr-layout.html
      - ../katex-amd-fix.html
    footer: "[back to the lecture slides on the course website](https://tmieno2.github.io/Data-Science-with-R-Quarto/lectures/)"
    theme: [default, ../custom.scss, ../notebook.scss]
    fontsize: 1.2em
    callout-icon: false
    scrollable: true
    echo: true
    fig-dpi: 400
    chalkboard: true
webr:
  packages: ['ggplot2', 'dplyr', 'sf', 'terra', 'stars', 'tidyterra'] # Install R packages on document open
# autoload-packages: false       # Disable automatic loading of packages
# show-startup-message: false    # Disable displaying status of webR initialization
  cell-options:
    editor-font-scale: 0.7
    dpi: 216
    fig-width: 7
    fig-height: 4.5
    out-width: 100%
filters:
  - webr
---

```{r}
#| include: false

source(here::here("lectures/_lecture-theme.R"))
```


```{r, include = FALSE}
library(sf)
library(dplyr)
library(data.table)
```

## Before you start

<details class="transcript">
<summary>Transcript</summary>

This is the deck that makes raster data useful to an economist. Almost every environmental variable you might want as a regressor arrives as a raster: rainfall, temperature, elevation, soil properties, satellite imagery. Almost every unit of analysis you care about is a vector: a field, a farm, a county, a well. So the recurring task is pulling raster values onto vector objects, and that is what extraction does. Before that we cover cropping and masking, which cut a raster down to the area you actually need. Both matter for speed, and on a large raster they matter a lot.

</details>

<br>

### Learning objectives

Learn the spatial interactions of a vector and raster dataset. Specifically,

+ Crop (spatially subset) a raster dataset based on the geographic extent of a vector dataset. 

+ Extract values from raster data for points and polygons. 

<br>

::: {.columns}

::: {.column width="50%"}
### Table of contents

1. [Cropping and masking raster data](#sec-crop-mask)
2. [Extract values from raster data](#sec-extract)
3. [Going the other way: `rasterize()` and `zonal()`](#sec-rasterize)
:::
<!--end of the 1st column-->
::: {.column width="50%"}
### Pre-requisite (Links)

+ [`ggplot2` primer](https://tmieno2.github.io/R-as-GIS-workshop/LA_1_ggplot2_primer.html)
+ [`dplyr` primer](https://tmieno2.github.io/R-as-GIS-workshop/LA_2_dplyr_primer.html)
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->


```{webr-r}
#| context: setup
#--- install and library the data package ---#
#--- Both repos are required. r.spatial.workshop.datasets declares         ---#
#--- Imports: sf, and webr::install() resolves dependencies ONLY in the     ---#
#--- repos named here. With the custom repo alone, sf cannot be found and   ---#
#--- the install never completes, so nothing downstream ever runs.          ---#
webr::install(
  "r.spatial.workshop.datasets",
  repos = c(
    "https://tmieno2.github.io/rwasm-package/repo",
    "https://repo.r-wasm.org"
  )
)

library(r.spatial.workshop.datasets)
data(wells_ne_sf)
data(ne_counties)
data(treatment_blocks)
data(corn_yield)
data(NDRE)
NDRE <- terra::rast(NDRE)
data(prism_us)
prism_us <- terra::rast(prism_us)

# Mirror of lecture_theme() in lectures/_lecture-theme.R. WebR runs in the
# browser against a virtual filesystem that does not hold the repo, so it
# cannot source that file and has to carry a copy. Edit the shared file first,
# then run `Rscript lectures/check-lecture-theme.R`, which fails if this copy
# has drifted and prints the block to paste.
lecture_theme <- function(base_size = 16) {
  ggplot2::theme_bw(base_size = base_size) +
    ggplot2::theme(
      axis.text = ggplot2::element_text(size = base_size * 0.9),
      axis.title = ggplot2::element_text(size = base_size),
      legend.text = ggplot2::element_text(size = base_size * 0.9),
      legend.title = ggplot2::element_text(size = base_size * 0.9),
      plot.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      legend.key = ggplot2::element_rect(fill = "transparent", colour = NA),
      # theme_bw's grey92 grid was chosen against a white panel and all but
      # disappears on the warm paper once the panel stops being opaque.
      panel.grid = ggplot2::element_line(colour = "grey87")
    )
}

# WebR draws its canvas at 72 dpi whatever `dpi` says, so on the 216-dpi canvas
# every size ggplot measures in points or millimetres comes out a third of what
# was asked for. Scaling ggplot2's own mm-to-point constants and the theme by
# the same factor puts WebR figures at the same type size as the knitr ones.
# Keep this in step with `dpi` in the header: change one, change the other.
webr_scale <- 216 / 72
assignInNamespace(".pt", ggplot2::.pt * webr_scale, "ggplot2")
assignInNamespace(".stroke", ggplot2::.stroke * webr_scale, "ggplot2")
theme_lecture <- lecture_theme(18 * webr_scale) +
  # base_size also sets the theme's line widths, which are in millimetres and
  # already pass through the scaled .pt; pin them to what 18pt gives, or they
  # come out nine times too thick.
  ggplot2::theme(
    line = ggplot2::element_line(linewidth = 18 / 22),
    rect = ggplot2::element_rect(linewidth = 18 / 22)
  )
ggplot2::theme_set(theme_lecture)
```

## Tips to make the most of the lecture notes

<details class="transcript">
<summary>Transcript</summary>

The usual arrangement, and this deck is one of the more practical ones, built around a single real on-farm experiment. Two habits will serve you here. First, check the coordinate reference systems before you combine any two objects, because most of the failures on this deck come from mismatched projections rather than from anything conceptual. Second, after every extraction, look at how many rows came back compared to what you put in. Extraction to points and extraction to polygons return quite differently shaped results, and knowing which you are holding matters.

</details>

<br>

### Interactive navigation tools

+ Click on the three horizontally stacked lines at the bottom left corner of the slide, then you will see table of contents, and you can jump to the section you want

+ Hit letter "o" on your keyboard and you will have a panel view of all the slides

<br>

### Running and writing codes

+ The box area with a hint of blue as the background color is where you can write code (hereafter referred to as the "code area").
+ Hit the "Run Code" button to execute all the code inside the code area.
+ You can evaluate (run) code selectively by highlighting the parts you want to run and hitting Command + Enter for Mac (Ctrl + Enter for Windows).
+ If you want to run the codes on your computer, you can first click on the icon with two sheets of paper stacked on top of each other (top right corner of the code chunk), which copies the code in the code area. You can then paste it onto your computer.
+ You can click on the reload button (top right corner of the code chunk, left to the copy button) to revert back to the original code.



## Cropping and masking raster data {#sec-crop-mask}

::: {.panel-tabset}

### Story and Data

<details class="transcript">
<summary>Transcript</summary>

A real setup, and a good one to have in mind. A nitrogen trial on a single field, with three objects. Corn yield as points, thousands of readings logged by the combine as it moved. Treatment blocks as polygons, the areas that received each nitrogen rate. And NDRE, a drone image of how green the crop is, as a raster. Print all three and look at them, then look at the map. The analytical problem is plain to see: three objects describing the same field, sharing no identifier, related only by position.

</details>

+ You have run an on-farm randomized nitrogen experiment on a field to quantify the impact of nitrogen on corn yield. 

+ You have three datasets
  + `corn_yield`: `sf` of corn yield (bu/acre) observations data represented by points
  + `treatment_blocks`: `sf` of treatment blocks represented by polygons
  + `NDRE`: `SpatRaster` of NDRE (roughly put, an indicator of how green the field is) taken by a drone

```{webr-r}
corn_yield
treatment_blocks
NDRE
```

<br>

Here is what they look like on a map:

```{webr-r}
#| fig-width: 3
#| code-track: 0.3
#| fig-max-height: 800 
ggplot() +
  geom_spatraster(
    data = NDRE,
    alpha = 0.4
  ) +
  geom_sf(
    data = corn_yield,
    aes(color = yield),
    size = 0.05
  ) +
  geom_sf(
    data = treatment_blocks,
    color = "red",
    fill = NA
  ) +
  scale_fill_viridis_c(
    name = "NDRE"
  ) +
  scale_color_continuous(
    name = "Corn Yield (bu/acre)"
  ) +
  theme_void()
```

### Objectives

<details class="transcript">
<summary>Transcript</summary>

Two operations, and they sound similar but do different things. Cropping cuts the raster down to a rectangle, the bounding box of a vector object, and it is fundamentally about size and speed. Masking sets cells outside your shapes to missing, following the actual outline rather than a rectangle, and it is mostly about appearance. Read the bullets under objective one, because the payoff is real: a smaller raster is faster to extract from and takes less memory. On drone imagery or a national weather grid, that difference is the one between a script that finishes and one that does not.

</details>


**Objective 1**

<span style = "color: red;"> Crop </span> the NDRE raster data (`NDRE`) to the bounding box of `corn_yield`

+ We can create a map that is more focused on the area of interest rather than the entire field 

+ We can stop carrying around the unnecessary parts of the data, which reduces its size (This matters when your raster data is spatially very fine and large.)

+ Extracting value to an `sf` from a smaller raster data is faster (we will talk about this later)

<br>

**Objective 2**

<span style = "color: blue;"> Mask </span> the `NDRE` data to `treatment_blocks` (assign NA to all the cells that are not intersecting with `treatment_blocks`)

+ We can create a map without any unnecessary cells presented 


### Crop

<details class="transcript">
<summary>Transcript</summary>

The syntax is one function and two arguments, but do read the demonstration carefully, because it is designed to fail first. Running crop with the untransformed yield points throws an error, and the message says extents do not overlap, which never mentions projections at all. The cause is that one object is measured in degrees and the other in metres, so their extents describe entirely different places. Note the callout's advice on which object to transform: always move the vector data to the raster's CRS, never the reverse, because re-projecting a raster resamples its values and cannot be undone.

</details>

::: {.panel-tabset}

#### How

You can crop a raster layer by using `terra::crop()`.

<br>

**Syntax**

```{r eval = F}
terra::crop(SpatRaster, sf)
```

<br>

The resulting `SpatRaster` object is the original cropped to the <span style = "color: red;"> bounding box </span> of the `sf` object.

#### Demonstration

```{webr-r}
NDRE_sub <- terra::crop(NDRE, corn_yield)
```

<br>

If you run the code above, then you should see an error. This happened because they do not share the same CRS.

```{webr-r}
crs(NDRE)
st_crs(corn_yield)
```

<br>

:::{.callout-important title="Important!"}
+ Projecting (or re-projecting to a different CRS) a raster dataset is typically a bad idea as it is irreversible. 
+ Re-project the vector data to the CRS of the raster data.
:::

<br>

Let's change the CRS of `corn_yield` to that of `NDRE` and then try to crop again.

```{webr-r}
NDRE_sub <- terra::crop(NDRE, st_transform(corn_yield, crs(NDRE)))
```

<br>

Let's check visually.

```{webr-r}
#| code-track: 0.3
#| fig-width: 3
#| fig-max-height: 800

ggplot() +
  geom_spatraster(
    data = NDRE_sub,
    alpha = 0.4
  ) +
  geom_sf(
    data = corn_yield,
    aes(color = yield),
    size = 0.05
  ) +
  geom_sf(
    data = treatment_blocks,
    color = "red",
    fill = NA
  ) +
  scale_fill_viridis_c(
    name = "NDRE"
  ) +
  scale_color_continuous(
    name = "Corn Yield (bu/acre)"
  ) +
  theme_void()
```
:::
<!--end of panel-->

### Mask 

<details class="transcript">
<summary>Transcript</summary>

Masking takes the same two arguments and sets every cell outside your polygons to missing. Compare its result with the crop from the previous tab: cropping gave you a rectangle that still included ground between and around the treatment blocks, whereas masking follows the block outlines exactly. Note also that we mask the already-cropped raster rather than the original, which is the usual order — crop first to make the object small, then mask for appearance. And read the note about na.value, because without it the missing cells render as grey rather than disappearing.

</details>

::: {.panel-tabset}

#### How

**Syntax**

```{r eval = F}
terra::mask(SpatRaster, sf)
```

<br>

The resulting `SpatRaster` object will have NA assigned to all the cells that are <span style = "color: red;"> not intersecting </span> with any of the geometries in the `sf` object.

#### Demonstration

```{webr-r}
NDRE_sub_masked <- terra::mask(NDRE_sub, st_transform(treatment_blocks, terra::crs(NDRE_sub)))
```

<br>

Let's check visually.

```{webr-r}
#| code-track: 0.3
#| fig-width: 3
#| fig-max-height: 700
ggplot() +
  geom_spatraster(
    data = NDRE_sub_masked, 
    alpha = 0.4
  ) +
  geom_sf(
    data = treatment_blocks, 
    color = "red", 
    fill = NA
  ) +
  scale_fill_viridis_c(
    name = "NDRE", 
    na.value = "transparent"
  ) +
  theme_void()
```

<br>

:::{.callout-note title="Note"}
+ adding `na.value = "transparent"` in `scale_fill_*()` will make the cells with NA value transparent (cannot be seen).
+ remove `na.value = "transparent"`, run the code again, and you will see that cells with NA are grey. 
:::

:::
<!--end of panel-->

### Exercise

<details class="transcript">
<summary>Transcript</summary>

Your turn, on a different scale entirely: PRISM precipitation covering the whole of the contiguous United States, cropped and masked down to Nebraska. This is the case where cropping earns its keep, since you are discarding the overwhelming majority of the raster. Do them in that order, crop and then mask, and notice that the two functions chain together with a pipe perfectly happily. Then compare what you get against the map in the Data tab, which shows the whole country, and the difference should make the point about size on its own.

</details>

::: {.panel-tabset}

#### Data

We use 

+ `prism_us`: a coarser version of PRISM precipitation data on 08/01/2012 covering the entire contiguous U.S. (`SpatRaster`)

+ `ne_counties`: counties in Nebraska (`sf`)

Here is what they look like:

```{webr-r}
#| context: output
ggplot() +
  geom_spatraster(data = prism_us) +
  geom_sf(data = ne_counties, color = "orange") +
  theme_void()
```

#### Exercise 1

Create the map you saw in the previous tab.

```{webr-r}
#| layout: stacked
ggplot() +
  geom_spatraster() +
  geom_sf() +
  theme_void()
```

<br>

**Answer**
```{r, eval = FALSE}
#| code-fold: true
ggplot() +
  geom_spatraster(data = prism_us) +
  geom_sf(data = ne_counties, color = "orange") +
  theme_void()
```

#### Exercise 2

Crop and then mask `prism_us` using `ne_counties`. 

```{webr-r}
#| layout: stacked

```

<br>

Create a map using the cropped- and masked-`prism_us` and `ne_counties`.

```{webr-r}
#| layout: stacked
 
```

<br>

**Answer**
```{r, eval = FALSE}
#| code-fold: true

prism_us_cropped_masked <-
  prism_us %>%
  terra::crop(ne_counties) %>%
  terra::mask(ne_counties)

ggplot() +
  geom_spatraster(data = prism_us_cropped_masked) +
  geom_sf(data = ne_counties, color = "orange") +
  theme_void()
```

:::
<!--end of panel-->

:::
<!--end of panel-->

## Extract values from raster layers to a vector data {#sec-extract}

::: {.panel-tabset}

### What is it?

<details class="transcript">
<summary>Transcript</summary>

Two tabs, one for each kind of target, and they behave differently in a way worth understanding before you write any code. Extracting to a point is simple: each point sits in exactly one cell, so it gets exactly one value. Extracting to a polygon is not: a polygon covers many cells, so it gets many values, and you must decide how to reduce them to one. The two diagrams show this with an eight-by-eight grid small enough to check by hand. Count the cells the polygon touches in the second picture and you will see the problem immediately.

</details>

::: {.panel-tabset}

#### to points

**Definition**

For each of the points, find which raster cell it is located within, and assign the value of the cell to the point.  

<br>

**Example**

```{r echo = F}
set.seed(378533)
library(sf)
library(stars)
library(dplyr)
library(ggplot2)

#--- create polygons ---#
polygon <- st_polygon(list(
  rbind(c(0, 0), c(8, 0), c(8, 8), c(0, 8), c(0, 0))
))

raster_like_cells <-
  st_make_grid(polygon, n = c(8, 8)) %>%
  st_as_sf() %>%
  mutate(value = sample(1:64, 64))

stars_cells <- st_rasterize(raster_like_cells, nx = 8, ny = 8)

cell_centroids <-
  st_centroid(raster_like_cells) %>%
  st_as_sf()
```

::: {.columns}


::: {.column width="50%"}
```{r, echo = F, out.width = "120%"}
#--------------------------
# Create points for which values are extracted
#--------------------------
#--- points ---#
point_1 <- st_point(c(2.4, 2.2))
point_2 <- st_point(c(6.7, 1.8))
point_3 <- st_point(c(4.2, 7.1))

#--- combine the points to make a single  sf of points ---#
points <-
  list(point_1, point_2, point_3) %>%
  st_sfc() %>%
  st_as_sf() %>%
  mutate(point_name = c("Point 1", "Point 2", "Point 3"))

#--------------------------
# Create maps
#--------------------------
ggplot() +
  geom_stars(data = stars_cells, alpha = 0.5) +
  scale_fill_distiller(name = "Value", palette = "Spectral") +
  geom_sf_text(data = raster_like_cells, aes(label = value)) +
  geom_sf(data = points, aes(shape = point_name), size = 3) +
  scale_shape(name = "Points") +
  theme_void()
```
:::

<!--end of the 1st column-->

::: {.column width="50%"}
+ The numbers inside the cells are the values that the cells hold. 

+ After the extraction, 
	* Point 1 will be assigned $50$
	* Point 2 will be assigned $4$
	* Point 3 will be assigned $54$.
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

#### to polygons

**Definition**

For each of the polygons, identify all the raster cells that intersect with the polygon, and assign a vector of the cell values to the polygon. 

<br>

**Example**

::: {.columns}

::: {.column width="50%"}
```{r, echo = FALSE}
#--------------------------
# Create a polygon for which values are extracted
#--------------------------
polygon_extract <- st_polygon(list(
  rbind(c(1.5, 2), c(6, 2.3), c(7, 6.5), c(2, 5), c(1.5, 2))
))

polygons_extract_viz <- ggplot() +
  geom_stars(data = stars_cells, alpha = 0.5) +
  scale_fill_distiller(name = "Value", palette = "Spectral") +
  geom_sf(data = polygon_extract, fill = NA, linewidth = 1) +
  geom_sf(data = cell_centroids, color = "black", size = 0.8) +
  geom_sf_text(data = raster_like_cells, aes(label = value), nudge_x = -0.25, nudge_y = 0.25) +
  theme_void()

polygons_extract_viz
```
:::
<!--end of the 1st column-->
::: {.column width="50%"}
+ Find all the raster cells each of the polygons "intersect with"

+ Assign the value of all the intersecting cells to the polygon (n-to-1). 
:::
<!--end of the 2nd column-->
:::
<!--end of the columns-->

:::
<!--end of panel-->

### Story

<details class="transcript">
<summary>Transcript</summary>

Back to the field experiment, and here is the reason for everything in this section. You have yield points, treatment blocks and a drone image sitting as three unrelated objects. You want two analyses, one where each yield point is an observation and one where each treatment block is. Neither is possible until the NDRE values are attached to those objects, and nothing but geography can attach them. This is the same joining problem as the previous deck, with a raster on one side instead of a vector, and the next two tabs take the two targets in turn.

</details>

+ Right now, corn yields (`corn_yield`), NDRE (`NDRE`), and treatment blocks (`treatment_blocks`) are separate R objects. 

+ We would like to conduct two kinds of analysis
  + analysis based on data where yield points are the unit of observations
  + analysis based on data where treatment blocks are the unit of observations
 
+ To achieve this, we would like to `join` them based on their locations
  + extract values from `NDRE` to `corn_yield`
  + extract values from `NDRE` to `treatment_blocks`


### How (to points)

<details class="transcript">
<summary>Transcript</summary>

Extraction to points, and the same function serves both cases. Two things to take from these tabs. First, note that extract behaves differently from crop when the projections disagree: rather than refusing, it quietly re-projects for you and warns that it did. Convenient, but transform explicitly anyway so your code says what it means. Second, understand the ID column, because it is how the result gets back to your data. Row n of the extraction corresponds to row n of the sf, which is what makes the assignment on the post-processing tab safe.

</details>

::: {.panel-tabset}

#### How

You can use `terra::extract()` with the following syntax. 

**Syntax**

```{r eval = F}
terra::extract(SpatRaster, sf of points)
```

#### Demonstration

Extract NDRE values to each of the yield points:

```{webr-r}
NDRE_extracted <- terra::extract(NDRE, corn_yield)
```

<br>

This one does **not** fail. Run it and you get a result, along with a warning
reading `[extract] transforming vector data to the CRS of the raster`.

:::{.callout-important title="`crop()` refuses, `extract()` transforms for you"}
The two functions treat a CRS mismatch completely differently, which is worth
knowing before it confuses you:

+ `terra::crop()` **errors**. The message is `[crop] extents do not overlap`, which
  never mentions the CRS — degrees and metres simply describe different places.
+ `terra::extract()` **carries on**, re-projecting the vector data for you and
  warning that it did.

So `extract()` is forgiving, but do not rely on it. Transform the vector data to
the raster's CRS yourself, as below, and the intent of your code is explicit.
:::

```{webr-r}
(
  NDRE_extracted <- terra::extract(NDRE, st_transform(corn_yield, crs(NDRE)))
)
```

:::{.callout-note title="Important"}
+ `ID` variable represents the row number in the `sf` (here, `corn_yield`). For example, `ID == 3` in `NDRE_extracted` is for `corn_yield[3, ]`. 
+ This becomes more important when we do extraction for polygons
:::


#### post-extraction processing

Just extracting the raster values to the points is not where we stop. We need to merge the extracted values back to the points data so that we can use them for further analysis.

Let's first check the class of `NDRE_extracted`.

```{webr-r}
class(NDRE_extracted)
```

<br>

The **n**th row in `NDRE_extracted` is for the **n**th point in `corn_yield`.

So, you can simply do this:

```{webr-r}
corn_yield <- mutate(corn_yield, NDRE = NDRE_extracted$NDRE)

#--- or ---#
# corn_yield$NDRE <- NDRE_extracted$NDRE
```

#### Multiple layers

You can extract values from multiple layers at the same time using `terra::extract()` just like you did with a single-layer raster data. 

For demonstration, let's create a multi-layer raster data:

```{webr-r}
#| autorun: true
#--- Create another layer ---#
NDRE_100 <- NDRE * 100
names(NDRE_100) <- "NDRE_100" # change the attribute name

#--- combine to form a two-layer SpatRaster ---#
(
  NDRE_two_layers <- c(NDRE, NDRE_100)
)
```

<br>

```{webr-r}
(
  NDRE_two_layers_extracted <- terra::extract(NDRE_two_layers, st_transform(corn_yield, crs(NDRE)))
)
```

<br>

The resulting object is a `data.frame`, and the values from the first (second) layer are in the second (third) column.

```{webr-r}
(
  corn_yield <- cbind(corn_yield, NDRE_two_layers_extracted[, -1])
)
```

<br>

:::{.callout-note title="Look at the column names"}
`corn_yield` already picked up an `NDRE` column a couple of tabs ago, so this
`cbind()` brings in a second copy. R does not overwrite it or complain — it renames
the newcomer `NDRE.1`. Harmless here, since the two hold the same values, but a
column silently named `something.1` is always worth a second look.
:::

:::
<!--end of panel-->

### How (to polygons)

<details class="transcript">
<summary>Transcript</summary>

Same function, quite different result, and this is the most useful section of the deck. Each polygon returns many rows, one per intersecting cell, so you have to summarise before you can use them. Three ways are shown, in increasing order of care: group and average by hand, let extract do it with the fun argument, or weight by how much of each cell actually falls inside the polygon using exact. That last one matters whenever the cells are large relative to the polygons. Also read the callout about assigning a summary table into a column, which is a mistake that fails silently.

</details>

::: {.panel-tabset}

#### How

You can use `terra::extract()` with the following syntax. Yes, same as value extraction to points.

**Syntax**

```{r eval = F}
terra::extract(SpatRaster, sf of polygons)
```

#### Demo

Extract NDRE values to each of the treatment blocks:

```{webr-r}
#| warning: false
NDRE_extracted_tb <- terra::extract(NDRE, st_transform(treatment_blocks, crs(NDRE)))
```

<br>

It's a `data.frame`.

```{webr-r}
class(NDRE_extracted_tb)
```

<br>

As you can see below, there is more than one NDRE value for each of the treatment blocks, which is expected, as many grid cells fall inside each one.

```{webr-r}
NDRE_extracted_tb[1:10, ]
NDRE_extracted_tb[200:210, ]
```


#### Post-extraction processing

Let's check the class of `NDRE_extracted_tb`.

```{webr-r}
class(NDRE_extracted_tb)
```

<br>

We just want one NDRE value for each of the treatment blocks. So, let's summarize them. In doing so, we summarize by `ID` as it indicates the row number of `treatment_blocks`. Here, we are getting the average.

```{webr-r}
avg_NDRE <-
  NDRE_extracted_tb %>%
  dplyr::group_by(ID) %>%
  dplyr::summarize(avg_NDRE = mean(NDRE))
```

<br>

Now, we can assign the average NDRE values to `treatment_blocks` like below because `ID == n` is for **n**th row of `treatment_blocks`.

```{webr-r}
treatment_blocks <- mutate(treatment_blocks, NDRE = avg_NDRE$avg_NDRE)

#--- or ---#
# treatment_blocks$NDRE <- avg_NDRE$avg_NDRE
```

<br>

:::{.callout-important title="A data.frame will go into a column without complaining"}
`dplyr` lets you put a data frame in a column on purpose, so `NDRE = avg_NDRE`
is not an error — it quietly nests the whole summary table inside `treatment_blocks`
and the row count still matches. You find out much later, when printing or plotting
the object fails with something that looks unrelated. Whenever the right-hand side
of a `mutate()` came from `summarize()`, name the column you actually want.
:::

#### Extract and summarize

You can actually extract and summarize both in `terra::extract()` using the `fun` option.

```{webr-r}
#| warning: false
NDRE_extracted_tb <- terra::extract(NDRE, st_transform(treatment_blocks, crs(NDRE)), fun = mean)
```

#### Area-weighted summarization

::: {.panel-tabset}

##### Why

+ In the previous cases of extraction and summarization tasks, all the intersecting cells are given the same weight irrespective of the degree of spatial overlap. 

+ This is very much acceptable in the current application, because the resolution of the raster data is high (cells are so small) relative to the size of the polygons.

+ However, if the cells are relatively large, you might want to consider calculating area-weighted summary.

<br>


##### How

We can add `exact = TRUE` option, which returns `fraction` variable indicating the fraction of the cells intersecting with the polygon.

```{webr-r}
#| warning: false
NDRE_extracted_tb <- terra::extract(NDRE, st_transform(treatment_blocks, crs(NDRE)), exact = TRUE)
```

<br>

Now, we can do fraction (area)-weighted summary of the extracted values:

```{webr-r}
NDRE_extracted_tb %>%
  dplyr::group_by(ID) %>%
  dplyr::summarize(NDRE = sum(NDRE * fraction) / sum(fraction))
```

:::
<!--end of panel-->

#### Multiple layers

Just like the case with value extraction to points, we can extract values from multiple layers to polygons in a single call with `terra::extract()`.

```{webr-r}
NDRE_extracted_tb_twolayers <- terra::extract(NDRE_two_layers, st_transform(treatment_blocks, crs(NDRE)), exact = TRUE)

head(NDRE_extracted_tb_twolayers)
```

<br>

Let's get the weighted average for both variables:

```{webr-r}
#| code-track: 0.5
NDRE_extracted_tb_twolayers %>%
  dplyr::group_by(ID) %>%
  dplyr::summarize(
    NDRE = sum(NDRE * fraction) / sum(fraction),
    NDRE_100 = sum(NDRE_100 * fraction) / sum(fraction)
  )
```

:::
<!--end of panel-->

:::
<!--end of panel-->


## Going the other way: `rasterize()` and `zonal()` {#sec-rasterize}

::: {.panel-tabset}

### Why

<details class="transcript">
<summary>Transcript</summary>

Everything so far in this deck ran one way, pulling raster values onto vector objects, because that is what you need most of the time. This section covers the reverse, and the motivation is specific. Think about the field experiment we have been using. Yield arrives as a point cloud from the combine, treatments as polygons from the trial design, greenness as a drone image on its own grid. Three geometries describing one field. If you want a regression where every row is the same patch of ground, they all have to be on one common grid, and getting them there means turning the vector layers into rasters.

</details>

Everything so far has run in one direction: raster values onto vector objects.
Occasionally you want the reverse — vector data turned into a grid.

The usual reason is that you have several layers describing the same field, all on
different geometries — a yield point cloud from the combine, treatment polygons from
the trial design, a drone image on its own grid — and you want them **on one common
grid** so that every row of your regression is the same patch of ground.

`terra::rasterize()` does the conversion, and `terra::zonal()` summarises a raster
using another raster as the grouping.

### `rasterize()`

<details class="transcript">
<summary>Transcript</summary>

The function takes the vector data, wrapped in vect because terra prefers its own class, and a raster whose grid you want to borrow. Read the callout, because the second argument confuses people: the NDRE raster is there purely to donate its geometry, its resolution and extent and projection. None of its values are used at all. That is what makes the technique work, since passing the same template to every layer guarantees they all come out cell for cell identical. The field argument says which column supplies the values, and here it is the nitrogen rate the trial assigned to each block.

</details>

::: {.panel-tabset}

#### Syntax

```{r eval = F}
terra::rasterize(SpatVector, SpatRaster, field, fun)
```

+ `SpatVector`: the vector data, wrapped in `terra::vect()`
+ `SpatRaster`: **the grid to use** — its cells and extent, not its values
+ `field`: which column of the vector data supplies the cell values
+ `fun`: how to combine, when several features land in one cell

:::{.callout-note title="The second argument is a template"}
`NDRE` is there only to donate its geometry — its resolution, extent and CRS. None
of its values are used. That is the whole trick: pass the same template to every
layer and they all come out on the identical grid, cell for cell.
:::

#### Try it

Put the treatment blocks onto the NDRE grid, carrying the nitrogen rate:

```{webr-r}
#| autorun: true
tb_prj <- st_transform(treatment_blocks, crs(NDRE))

rate_raster <- terra::rasterize(terra::vect(tb_prj), NDRE, field = "target_rate")

rate_raster
```

<br>

```{webr-r}
#| code-track: 0.4
#| fig-width: 1.8
#| fig-max-height: 600
ggplot() +
  geom_spatraster(data = terra::trim(rate_raster)) +
  scale_fill_viridis_c(
    name = "N rate",
    na.value = "transparent"
  ) +
  theme_void()
```

<br>

:::{.callout-note title="A warning you can ignore here"}
In the browser, `rasterize()` prints
`GDAL Error 1: PROJ: proj_crs_get_coordinate_system: Object is not a SingleCRS`.
Despite saying Error it is a warning, and it comes from the WebAssembly build of
PROJ rather than from anything you did. The raster it returns is correct — the
nitrogen rates and the `zonal()` table below match what the same code produces in
RStudio. You will not see this message on your own machine.
:::

:::
<!--end of nested panel-->

### `zonal()`

<details class="transcript">
<summary>Transcript</summary>

Once two rasters share a grid, one of them can group the other, and that is what zonal does. It takes the raster you want summarised, the raster defining the zones, and a function. Here it gives mean greenness by nitrogen rate, which is the question the trial was designed to answer. Look at the numbers rather than just the fact that it ran. Greenness does rise with nitrogen, and it rises very slightly, from about point six zero four to point six one two across the whole range. Noticing how small that is matters more than the code.

</details>

Once two rasters share a grid, one can group the other.

**Syntax**

```{r eval = F}
terra::zonal(SpatRaster, zones_SpatRaster, fun)
```

<br>

So, mean greenness by nitrogen rate — the actual question the trial was run to
answer:

```{webr-r}
terra::zonal(NDRE, rate_raster, fun = "mean", na.rm = TRUE)
```

<br>

:::{.callout-note title="Read the result"}
Five nitrogen rates, five mean NDRE values, and greenness does rise with the rate —
though by very little, which is itself worth noticing before anyone gets excited.
:::

### Which route should you use?

<details class="transcript">
<summary>Transcript</summary>

You can get the same table with extract and group_by from earlier in the deck, and it is worth knowing that both are legitimate. Run both and the answers agree to about three decimal places, which is close but not identical, and the callout explains why. Rasterize assigns a whole cell by where its centre falls, so a half-covered cell is either in or out. Extract returns every intersecting cell however slightly it overlaps. Neither convention is more correct, but you should know which one produced your number, because on a coarse grid the two can diverge sharply.

</details>

You can reach that table the other way, with `extract()` and `group_by()`, as
earlier in this deck. Both are legitimate.

```{r eval = F}
#--- via extract ---#
terra::extract(NDRE, tb_prj) %>%
  dplyr::group_by(ID) %>%
  dplyr::summarize(NDRE = mean(NDRE, na.rm = TRUE))
```

<br>

:::{.callout-important title="They agree, but not exactly"}
Running both gives 0.6044 and 0.6045 for the first treatment rate, and similarly
close numbers for the rest. The gap is not an error in either one — the two count
cells differently:

+ `rasterize()` assigns a whole cell to a polygon based on **where the cell centre
falls**. A cell half-covered is either in or out.
+ `extract()` returns **every intersecting cell**, however slightly it overlaps,
unless you ask for `exact = TRUE`.

Neither is more correct in general. What matters is that you know which convention
produced your number, because on coarse grids the two can differ a great deal.
:::

<br>

:::{.callout-note title="So when is `rasterize()` worth it?"}
Use `extract()` when you want values on your vector objects, which is most of the
time. Reach for `rasterize()` when you need several **different** layers aligned to
one grid — which is exactly the setup for a cell-level regression, and the reason
this section exists.
:::

:::
<!--end of panel-->
